### 2011-12-05 paper/draft1.tex
draft1 init

\documentclass{article}
\begin{document}
\title{Machine Learning for Auto-Auto-Tuning}
\begin{abstract}
Each generation of GPU hardware presents unique programming challenges.
It is difficult to write software that obtains good performance on a range of hardware.

Auto-tuning is an effective technique for adapting a parametrized
GPU code template to current hardware.

Autotuning, however, is a relatively expensive kernel-selection mechanism - it
works by actually running the desired GPU computation in tens or hundreds of
ways before selecting the best implementation strategy for a particular
machine.

Futhermore library abstraction boundaries provide operations such as image
filtering and matrix multiplication, which actually correspond to a large set
of possible computations: for example the convolution of 1 large image with 1
small filters is  quite a different computational problem than the convolution
of 100 small images with 10 large filters.
This paper considers (XXX) strategies for using previous auto-tuning
measurements from similar problems to short-circuit computationally
expensive auto-tuning on a new problem of interest.

This paper presents a machine learning approach to generalized auto-tuning, in
which features of
(a) the current hardware platform,
(b) the kernel configuration
and (c) the problem instance are used to fit a regression model that predicts
how fast an implementation is (XXX).
We argue that autotuning using this surrogate model can be effective, while
being much faster than normal auto-tuning.

Whereas auto-tuning on problem XXX takes XXX seconds, our approach obtains
(XXX) of the speed in (XXX) of the time.
\end{abstract}
\end{document}

### 2011-12-06 paper/draft1.tex
m

@@
-\documentclass{article}
+\documentclass{sig-alternate}
+\usepackage{graphicx}
+\usepackage{amsmath}
+\usepackage{amssymb}
+
+\title{Model-Based Auto-Tuning}
+
+\numberofauthors{3}
+
+\author{
+%\alignauthor First Last\titlenote{}\\
+\alignauthor First Last\\
+\affaddr{Affiliation line 1}\\
+\affaddr{Affiliation line 2}\\
+\email{anon@mail.com}
+% 2nd author
+%\alignauthor First Last\titlenote{None}\\
+\alignauthor First Last\\
+\affaddr{Affiliation line 1}\\
+\affaddr{Affiliation line 2}\\
+\email{anon@mail.com}
+% 3rd author
+%\alignauthor First Last\titlenote{None}\\
+\alignauthor First Last\\
+\affaddr{Affiliation line 1}\\
+\affaddr{Affiliation line 2}\\
+\email{anon@mail.com}
+}
+
 \begin{document}
-\title{Machine Learning for Auto-Auto-Tuning}
+\maketitle
+
 \begin{abstract}
@@
 filtering and matrix multiplication, which actually correspond to a large set
-of possible computations: for example the convolution of 1 large image with 1
-small filters is  quite a different computational problem than the convolution
+of possible computations: for example the correlation of 1 large image with 1
+small filters is  quite a different computational problem than the correlation
 of 100 small images with 10 large filters.
@@
 (XXX) of the speed in (XXX) of the time.
+
+XXX: Hill-climbing (HC) search is effective alternative to grid.
+XXX: How is HC search affected by mutation rate?
+XXX: Purely random search is not an effective alternative to grid.
+
 \end{abstract}
+
+\section{Introduction}
+
+There is a natural function from
+problem space x implementation space x platform space to runtime:
+how much wall time elapses on the given platform when solving the given
+problem with the given implementation.
+
+Autotuning is a family of empirical techniques for finding the implementation
+that minimizes that runtime function, when the platform and the problem
+configuration are given.
+
+XXX: some math, or possibly a picture with the three boxes that was on Pinto's
+whiteboard.
+
+This paper is about different ways of doing the minimization.
+For a particular problem (filterbank correlation) we
+look at a parameterized kernel implementation (from GCG)
+and a few hardware platforms, and compare
+random search, grid search, and a hill-climbing strategy with a good, but
+generic, reference implementation.
+Then we show that we can model the LSM function with a regression tree
+well enough to usefully perform the argmin of Equation~\ref{eq:lsm_argmin}
+on the model rather than the actual system. Whereas it may take several
+minutes to perform a hill-climbing or grid search in the original system,
+a hill-climbing search in the model requires a small fraction of a second.
+Model-based autotuning makes it possible to simulate auto-tuning quickly and
+accurately enough to be used within a single library call of the computational
+routine.
+
+\subsection{Problem Space: Filter-bank Correlation}
+
+Our experiments on automatic auto-tuning will be focused on filterbank
+correlation.  Mathematically, we define filterbank correlation in terms of an
+image $x$ and a filterbank $f$.
+The image $x$ has $R$ rows, $C$ columns, and $D$ channels (e.g. color
+channels) that we call its {\em depth}. We index $x$ like $\mathbf{x}[i,j,d]$
+where $1 \leq i \leq R$ and $1 \leq j \leq C$.
+The filterbank $f$ has $K$ square filters of side-length $W$, each one of
+which has $D$ channels.
+The result of filterbank correlation of $x$ with $f$ is an image-like array
+$z$ with $R-W+1$ rows, $C-W+1$ columns, and depth $K$, whose elements are
+defined according to Equation~\ref{eq:z}:
+
+\begin{equation}
+    \mathbf{z}[i,j,k] = \sum_{u=1}^{W} \sum_{v=1}^{W} \sum_{d=1}^D
+        \mathbf{x}[i+u, j+v, d]~ \mathbf{f}[k, u, v, d].
+        \label{eq:z}
+\end{equation}
+
+Filterbank correlation is used extensively in image and video processing,
+where it is often the computational bottleneck.
+GPUs are well-suited to computing filterbank correlations, although the
+optimal strategy for doing so often depends on the shape of the problem, i.e.
+the choices of $R$, $C$, $D$, $F$, and $W$.
+
+We randomly sampled problem specifications from the following space:
+\begin{figure}
+\begin{center}
+\begin{minipage}{.4\textwidth}
+\begin{verbatim}
+R : one_of(256, 512, 1024, 2048, 4096)
+D : one_of(1, 4, 8, 16, 32, 64, 128, 256)
+F : one_of(1, 4, 8, 16, 32, 64, 128, 256)
+W : one_of(3, 5, 7, 9, 11)
+\end{verbatim}
+\end{minipage}
+\end{center}
+\caption{The full configuration space for filterbank correlation included a
+total of 1600 elements, but we restricted our experiments to problems that
+represented between 1 and 50 gigaflops. Our problem space included 602
+configurations. In a full application in which all image and filter sizes are
+valid, the number of configurations would be effectively infinite.
+XXX: define R, D, F, W here
+}
+\label{fig:probspace}
+\end{figure}
+% Problems were considered invalid if the filters were larger than the images.
+% but in the new setting with iheight >= 256 this never happened.
+Problems were tested if they represented between 1 and 50 gigaflops.  Problems
+that were smaller do not fully utilize the hardware, and are handled equally
+well by many kernel settings, including the reference (XXX cite GCG).
+Problems that were larger take so long to evaluate that there is negligible
+inefficiency in implementing them via multiple calls with smaller images and fewer filters.
+The configuration space described in Figure~\ref{fig:probspace} included 602
+configurations with between 1 and 50 gigaflops.
+
+In a full, reusable, library implementation of the filter correlation
+operation, we would also provide options for arbitrary image shapes,
+arbitrary integer depth and filter shapes, and also support a variety of
+physical input layouts via strides.  These additional options would make our
+approach of automatic auto-tuning even more important, because there would be
+a greater variety in the kinds of computations and memory transfers to
+perform.  In our experiments, we only consider the problem variations
+described in Figure ~\ref{tbl:}, which already suffice to show that automatic
+auto-tuning is a promising approach.
+
+\subsection{Kernel Space: Metaprogramming}
+
+\begin{figure}
+\begin{center}
+\begin{minipage}{.43\textwidth}
+\begin{verbatim}
+block_w   : one_of(4, 8, 16, 32, 64, 128)
+block_h   : one_of(4, 8, 16, 32, 64, 128)
+n_filter_r: one_of(1, 2)
+n_output4s: one_of("all", 1, 2)
+spill     : one_of(False, True)
+imul_fast : one_of(False, True)
+pad_shared: one_of(False, True)
+use_tex1d : one_of(False, True)
+maxrreg   : one_of(8, 16, 20, 24, 28, 32, inf)
+fast_math : one_of(False, True)
+\end{verbatim}
+\end{minipage}
+\end{center}
+\caption{The kernel used to perform
+filterbank correlation was parametrized by 10 parameters, some of which
+were binary and others of which were integer-valued. The full configuration
+space included 12000 kernels.
+XXX: how to make sense of these paramters without code listing?
+}
+\end{figure}
+
+XXX: What to all the parameters mean?
+XXX: Point to github / GCG for full code listing.
+
+XXX: how many configurations are in the configuration space
+
+XXX: what was the reference kernel, and how was it chosen?
+
+\iffalse
+\subsection{Platform Space: XXX}
+
+XXX: GCG chapter pg 13 notes that among the grid search, the best parameter
+setting is different for the various platforms.
+
+XXX: so far we only tried 3 cards... so we're using a 1-of-N feature.
+\fi
+
+\section{Results}
+
+\subsection{Random, Grid, and Hill-Climbing Search}
+
+XXX: Histogram (over problems) of gigaflops/sec of random, grid, random,
+hill-climbing for each/selected platform.
+This shows that our implementations and problems are stressing the cards,
+that our implementation is pretty good, and that different problem sizes pose
+different computational challenges.
+
+XXX: Scatterplot of gigaflops vs. gigaflops / second (of best grid point)
+should show that it's not just problem size that makes it possible to go fast.
+
+XXX: histogram of speedups of grid, random, hill-climbing over reference, for
+several hardware platforms. (GET ANOTHER -- macbook??)
+Sometimes the reference implementation is optimal, in many cases it is far
+from optimal: Which cases? What's wrong with the reference?
+
+TL;DR: Hill-climbing is a good way to search the implementation space, can
+search the full space and do about as well as grid search in the more
+restricted space.
+
+\begin{figure}
+\includegraphics[scale=.5]{speedup_295.pdf}
+\caption{Speedup of various auto-tuning strategies over a good reference
+implementation.
+Each strategy was evaluated the same set of 83 problem configurations, which
+was disjoint from the problem configurations used to build the model for
+model-based hill-climbing.
+The vertical dashed lines are positioned at the geometric mean speedup (XXX?)
+of each strategy. The grid and hill-climbing approaches
+tested 73 and 75 configurations respectively, and took 45?? XXX
+seconds on average.
+The model-based approach tested 0 configuration evaluations, and took 0.05
+seconds on average, even with a naive Python implementation.
+}
+\label{fig:speedup_295}
+\end{figure}
+
+\subsection{Model-based Autotuning}
+
+On the GTX 295, a single decision-tree fit to the log-speed-multiplier
+(LSM) function (XXX: how many problem configurations, how much vs. what
+kind of training data) achieves an average Spearman correlation of 0.9 on
+held-out test data (XXX: what kind of test data).
+
+Now: the big test is that if we use the genetic search on the model instead of
+the original function, do we get good performance?
+
+\section{Discussion}
+
+We're characterizing hardware by a 1-of-N feature vector, simply describing
+which hardware is the current hardware.
+To make better use of auto-tuning data from other platforms, it would be more
+useful to have precise and descriptive features such as: what compute
+capability is present, how many cores are active, what is the bandwidth
+between the various kinds of memory, and how much of various kinds of memory
+is present.  With these features a model-assisted auto-tuning approach
+might be able to make very good guesses on hardware for which no auto-tuning
+has ever been done.
+
 \end{document}

### 2011-12-06 experiment.py

@@ -35,6 +35,28 @@ def init_cuda(dev_id):
 #       time? (Should it be more important to tune the more expensive calls? I
 #       think yes)
 
+def main_count_problem_space_size():
+    n_valid = 0
+    for iheight in (256, 512, 1024, 2048, 4096):
+        for depth in (1, 4, 8, 16, 32, 64, 128, 256):
+            for n_filters in (1, 4, 8, 16, 32, 64, 128, 256):
+                for fsize in (3, 5, 7, 9, 11):
+                    prob_spec = wisdom.ProblemSpec(
+                            n_imgs=1,
+                            height=iheight,
+                            width=iheight,
+                            depth=depth,
+                            n_filters=n_filters,
+                            filter_height=fsize,
+                            filter_width=fsize,
+                            img_strides=None,
+                            filter_strides=None,
+                            border_mode='valid')
+                    if 1 < prob_spec.gflops() < 50:
+                        if prob_spec.is_valid():
+                            n_valid += 1
+    print n_valid
+
 def problem_generator(rng):
     # TODO: sample fbcorr parameters from within LFW models
     space = rSON2(
@@ -188,6 +210,190 @@ def main_step():
         cPickle.dump((wdb, results, rng), ofile)
         ofile.close()
 
+def main_dump_logspeedup_dataset():
+    timings = {}
+    refspeed = {}
+    try:
+        ofile = open(sys.argv[2])
+        print 'Please remove output file manually:', sys.argv[2]
+        return
+    except IOError:
+        pass
+    for pklfile in sys.argv[3:]:
+        assert len(sys.argv) == 4  # XXX: platformfeatures are not implemented
+        print "Loading data from ", pklfile
+        wdb, results, rng = cPickle.load(open(pklfile))
+        for ii, finding in enumerate(results):
+            ref = finding['ref']
+            ref_orig = finding['ref_orig']
+            key_ii = (ref.prob_spec, pklfile)
+
+            refspeed.setdefault(key_ii,
+                    np.mean([ref.speed(), ref_orig.speed()]))
+            timings.setdefault((ref.prob_spec, pklfile), [])
+
+            for k, t in finding.iteritems():
+                if not t.valid:
[…]

### 2011-12-06 wisdom.py

@@ -108,6 +108,10 @@ def random_op_cross(op1, op2, rng, r=.5):
             op1.use_fast_math if rng.rand() < r else op2.use_fast_math,
             )
 
+def resample_some_coords(op, rng, rate=.5):
+    rop = random_op_spec(rng)
+    return random_op_cross(rop, op, rng, rate)
+
 
 def reference_op_spec():
     return OpSpec(block_w=8,
@@ -149,20 +153,34 @@ class ProblemSpec(object):
         else:
             raise ValueError(self.border_mode)
 
+        if img_strides is not None:
+            # XXX add to features
+            raise NotImplementedError()
+
+        if filter_strides is not None:
+            # XXX add to features
+            raise NotImplementedError()
+
         if n_imgs != 1:
+            # XXX add to features
             raise NotImplementedError()
         if border_mode != 'valid':
+            # XXX add to features
             raise NotImplementedError()
 
-    def is_valid(self):
-        return self.out_height > 0 and self.out_width > 0
+    def __repr__(self):
+        assigns = ['%s=%s' % (n, v) for v, n in self.feature_pairs()]
+        return "ProblemSpec(%s)" % ", ".join(assigns)
 
-    def gflops(self):
-        return (self.n_imgs
-                * self.out_height * self.out_width * self.n_filters
-                * self.filter_height * self.filter_width * self.depth
-                * 2  # mul and add
-                / (1000.**3.)) #return as giga float ops
+    def __eq__(self, other):
+        return (type(self) == type(other)
+                and (self.feature_pairs() == other.feature_pairs()))
+
+    def __ne__(self, other):
+        return not self == other
+
+    def __hash__(self):
+        return hash((type(self),) + tuple(self.feature_pairs()))
 
     # relevant hand-designed features of problem specification
     def feature_pairs(self):
@@ -185,13 +203,16 @@ class ProblemSpec(object):
         # filters flipped vertically (conv vs. corr)
         # filters flipped horizontally (conv vs. corr)
         names = [
-            'n_imgs',
+            #'n_imgs',
             'height',
             'width',
             'depth',
             'n_filters',
             'filter_height',
             'filter_width',
+            #'img_strides',
+            #'filter_strides',
+            #'border_mode',
             ]
         return [(getattr(self, name), name) for name in names]
 
@@ -202,9 +223,15 @@ class ProblemSpec(object):
         return map(float, zip(*self.feature_pairs())[0])
 
 
[…]

### 2011-12-06 paper/abstract.tex
m

\documentclass{sig-alternate}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}

\title{Boosted Regression Trees for Model-Free Auto-Tuning}

\numberofauthors{3}

\author{
%\alignauthor First Last\titlenote{}\\
\alignauthor First Last\\
\affaddr{Affiliation line 1}\\
\affaddr{Affiliation line 2}\\
\email{anon@mail.com}
% 2nd author
%\alignauthor First Last\titlenote{None}\\
\alignauthor First Last\\
\affaddr{Affiliation line 1}\\
\affaddr{Affiliation line 2}\\
\email{anon@mail.com}
% 3rd author
%\alignauthor First Last\titlenote{None}\\
\alignauthor First Last\\
\affaddr{Affiliation line 1}\\
\affaddr{Affiliation line 2}\\
\email{anon@mail.com}
}

\begin{document}
\maketitle

\begin{abstract}

Autotuning is a widely used and effective technique for optimizing a
parametrized GPU code template for a particular type of computation on
particular hardware.
Its drawback is that exhaustive or even thorough autotuning can be very slow.
Furthermore, library abstraction boundaries provide operations such as image
filtering and matrix multiplication, which actually correspond to a large set
of potential problem configurations, with a potentially wide variety of memory
access patterns and computational bottlenecks.
How can we draw on data from previous auto-tuning of related problems on related hardware to make a
good decision on a novel problem?
This paper presents a machine learning approach to auto-tuning, in
which features of
(a) the current hardware platform,
(b) the kernel configuration
and (c) the problem instance are used to fit a regression model (boosted regression trees) that predicts
how much faster an implementation will be than a reference baseline.
Combinatorial optimization strategies that would normally be applied for autotuning can be applied to the regression model to find an implementation much more quickly.
We validate our approach using the filterbank correlation kernel described in [XXX].
We find that 0.05 seconds of a hill climbing on the regression model can achieve 80\% of the improvement brought by 120 seconds of real autotuning.
Our approach is not specific to filterbank correlation, or even to GPU autotuning: the approach of using a non-linear regression model on top of simple features applies to a variety of problem types, kernel types, and platforms.
A detailed model of the platform can provide useful features for cross-platform autotuning, but such a model is not necessary for same-platform autotuning.

\end{abstract}

\end{document}

### 2011-12-06 paper/abstract.tex
m

@@
 
-\title{Boosted Regression Trees for Model-Free Auto-Tuning}
+\title{Boosted Regression Trees for Data-Driven Auto-Tuning}
+% Automatic
+% of GPU Code
+% Just-In-Time Code Specialization
 
@@
 \begin{abstract}
-
-Autotuning is a widely used and effective technique for optimizing a
+Auto-tuning is a widely used and effective technique for optimizing a
 parametrized GPU code template for a particular type of computation on
 particular hardware.
-Its drawback is that exhaustive or even thorough autotuning can be very slow.
+Its drawback is that thorough or exhaustive auto-tuning requires compiling
+many kernels, and calling each one many times; this process is slow.
 Furthermore, library abstraction boundaries provide operations such as image
 filtering and matrix multiplication, which actually correspond to a large set
-of potential problem configurations, with a potentially wide variety of memory
+of potential problem configurations with a wide variety of memory
 access patterns and computational bottlenecks.
@@
 how much faster an implementation will be than a reference baseline.
-Combinatorial optimization strategies that would normally be applied for autotuning can be applied to the regression model to find an implementation much more quickly.
-We validate our approach using the filterbank correlation kernel described in [XXX].
-We find that 0.05 seconds of a hill climbing on the regression model can achieve 80\% of the improvement brought by 120 seconds of real autotuning.
-Our approach is not specific to filterbank correlation, or even to GPU autotuning: the approach of using a non-linear regression model on top of simple features applies to a variety of problem types, kernel types, and platforms.
-A detailed model of the platform can provide useful features for cross-platform autotuning, but such a model is not necessary for same-platform autotuning.
-
+Combinatorial optimization strategies that would normally be applied for auto-tuning can be applied to the regression model to find an implementation much more quickly.
+We validate our approach using the filterbank correlation kernel described in [anon].
+We find that 0.05 seconds of hill climbing on the regression model can achieve
+80\% of the improvement brought by 120 seconds of real auto-tuning.
+Our approach is not specific to filterbank correlation, or even to GPU auto-tuning: the approach of using a non-linear regression model on top of simple features applies to a variety of problem types, kernel types, and platforms.
+A detailed platform model can provide useful features for
+cross-platform auto-tuning, but such a model is not necessary for same-platform autotuning.
 \end{abstract}

### 2011-12-06 paper/draft1.tex
m

@@
 
-\title{Model-Based Auto-Tuning}
+\title{Boosted Regression Trees for Model-Free Auto-Tuning}
 
@@
 \begin{abstract}
-Each generation of GPU hardware presents unique programming challenges.
-It is difficult to write software that obtains good performance on a range of hardware.
-
-Auto-tuning is an effective technique for adapting a parametrized
-GPU code template to current hardware.
-
-Autotuning, however, is a relatively expensive kernel-selection mechanism - it
-works by actually running the desired GPU computation in tens or hundreds of
-ways before selecting the best implementation strategy for a particular
-machine.
-
-Futhermore library abstraction boundaries provide operations such as image
+Autotuning is a widely used and effective technique for optimizing a
+parametrized GPU code template for a particular type of computation on
+particular hardware.
+Its drawback is that thorough or exhaustive auto-tuning can be very slow.
+Furthermore, library abstraction boundaries provide operations such as image
 filtering and matrix multiplication, which actually correspond to a large set
-of possible computations: for example the correlation of 1 large image with 1
-small filters is  quite a different computational problem than the correlation
-of 100 small images with 10 large filters.
-This paper considers (XXX) strategies for using previous auto-tuning
-measurements from similar problems to short-circuit computationally
-expensive auto-tuning on a new problem of interest.
-
-This paper presents a machine learning approach to generalized auto-tuning, in
+of potential problem configurations with a wide variety of memory
+access patterns and computational bottlenecks.
+How can we draw on data from previous auto-tuning of related problems on related hardware to make a
+good decision on a novel problem?
+This paper presents a machine learning approach to auto-tuning, in
 which features of
@@
 (b) the kernel configuration
-and (c) the problem instance are used to fit a regression model that predicts
-how fast an implementation is (XXX).
-We argue that autotuning using this surrogate model can be effective, while
-being much faster than normal auto-tuning.
-
-Whereas auto-tuning on problem XXX takes XXX seconds, our approach obtains
-(XXX) of the speed in (XXX) of the time.
-
-XXX: Hill-climbing (HC) search is effective alternative to grid.
-XXX: How is HC search affected by mutation rate?
-XXX: Purely random search is not an effective alternative to grid.
-
+and (c) the problem instance are used to fit a regression model (boosted regression trees) that predicts
+how much faster an implementation will be than a reference baseline.
+Combinatorial optimization strategies that would normally be applied for auto-tuning can be applied to the regression model to find an implementation much more quickly.
+We validate our approach using the filterbank correlation kernel described in [anon].
+We find that 0.05 seconds of hill climbing on the regression model can achieve
+80\% of the improvement brought by 120 seconds of real auto-tuning.
+Our approach is not specific to filterbank correlation, or even to GPU auto-tuning: the approach of using a non-linear regression model on top of simple features applies to a variety of problem types, kernel types, and platforms.
+A detailed platform model can provide useful features for
+cross-platform auto-tuning, but such a model is not necessary for same-platform autotuning.
 \end{abstract}
@@
 
-\begin{figure}
-\includegraphics[scale=.5]{speedup_295.pdf}
+\begin{figure*}
+\centering
+\includegraphics[scale=.42]{speedup_295.pdf}
+\includegraphics[scale=.42]{speedup_480.pdf}
 \caption{Speedup of various auto-tuning strategies over a good reference
@@
 \label{fig:speedup_295}
-\end{figure}
+\end{figure*}
 

### 2011-12-06 experiment.py

@@ -57,6 +57,7 @@ def main_count_problem_space_size():
                             n_valid += 1
     print n_valid
 
+
 def problem_generator(rng):
     # TODO: sample fbcorr parameters from within LFW models
     space = rSON2(
@@ -83,6 +84,7 @@ def problem_generator(rng):
         if 1 < prob_spec.gflops() < 50:
             if prob_spec.is_valid():
                 yield prob_spec
+
 
 def main_step():
     _python, _cmd, dev_id_str, wisdomfile = sys.argv
@@ -210,6 +212,7 @@ def main_step():
         cPickle.dump((wdb, results, rng), ofile)
         ofile.close()
 
+
 def main_dump_logspeedup_dataset():
     timings = {}
     refspeed = {}
@@ -297,20 +300,30 @@ def main_train():
     import scipy.stats
 
     print "VAR: ", np.var(test_y)
-    for d in range(13, 14):
+    cur_train_y = train_y
+    cur_pred_y = np.zeros_like(test_y)
+    mdls = []
+    weak_depth = 4
+    for boosting_iter in range(100):
         mdl = sklearn.tree.DecisionTreeRegressor(
-                max_depth=d,
+                max_depth=weak_depth,
                 min_split=10,
                 )
-        mdl.fit(train_x, train_y)
-        pred_y = mdl.predict(test_x)
-        rcorr = [scipy.stats.spearmanr(pred_y[idxs], test_y[idxs])[0]
+        mdl.fit(train_x, cur_train_y)
+        cur_train_y -= mdl.predict(train_x)
+        cur_pred_y += mdl.predict(test_x)
+        rcorr = [scipy.stats.spearmanr(cur_pred_y[idxs], test_y[idxs])[0]
             for pspec, idxs in pspec_idxs.items()]
-        print rcorr
-        print "%i: MSE: %3.3f  MRC: %3.3f" % (
-                d,
-                np.mean((pred_y - test_y) ** 2),
+        #print rcorr
+        print "%i: TRAINMSE: %3.3f TESTMSE: %3.3f MRC: %3.3f" % (
+                boosting_iter,
+                np.mean(cur_train_y ** 2),
+                np.mean((cur_pred_y - test_y) ** 2),
                 np.mean(rcorr))
+        mdls.append(mdl)
+
+    def predict_one(x):
+        return np.sum([mdl.predict(x) for mdl in mdls])
 
     print "Validating Model against actual hardware"
     device = init_cuda(int(dev_id_str))
[…]

### 2011-12-06 paper/abstract.tex
m

@@
 Auto-tuning is a widely used and effective technique for optimizing a
-parametrized GPU code template for a particular type of computation on
+parametrized GPU code template for a particular computation on
 particular hardware.
 Its drawback is that thorough or exhaustive auto-tuning requires compiling
-many kernels, and calling each one many times; this process is slow.
+many kernels and calling each one many times; this process is slow.
 Furthermore, library abstraction boundaries provide operations such as image
@@
 How can we draw on data from previous auto-tuning of related problems on related hardware to make a
-good decision on a novel problem?
+just-in-time implementation decision for a novel problem?
 This paper presents a machine learning approach to auto-tuning, in
@@
 (b) the kernel configuration
-and (c) the problem instance are used to fit a regression model (boosted regression trees) that predicts
-how much faster an implementation will be than a reference baseline.
-Combinatorial optimization strategies that would normally be applied for auto-tuning can be applied to the regression model to find an implementation much more quickly.
-We validate our approach using the filterbank correlation kernel described in [anon].
-We find that 0.05 seconds of hill climbing on the regression model can achieve
-80\% of the improvement brought by 120 seconds of real auto-tuning.
+and (c) the problem instance are passed to a regression model (boosted
+regression trees) which predicts
+how much faster this kernel will be than a reference baseline.
+Combinatorial optimization strategies that would normally implement
+auto-tuning by evaluating kernel configurations on the real hardware are
+orders of magnitude faster when evaluating the regression model instead.
+We validate our approach using the filterbank correlation kernel described in
+[anon], where we find that 0.1 seconds of hill climbing on the regression
+model can achieve an average of
+90\% of the speed brought by 120 seconds of real auto-tuning.
 Our approach is not specific to filterbank correlation, or even to GPU auto-tuning: the approach of using a non-linear regression model on top of simple features applies to a variety of problem types, kernel types, and platforms.
-A detailed platform model can provide useful features for
-cross-platform auto-tuning, but such a model is not necessary for same-platform autotuning.
 \end{abstract}

### 2011-12-06 experiment.py

@@ -384,24 +384,34 @@ def main_figtrain():
         logscale_x = False
     elif '480' in train_result:
         logscale_x = True
+    elif '1060' in train_result:
+        logscale_x = False
 
     def maybe_log(x):
         if logscale_x:
             return np.log(x)
         else:
             return x
+    def use_this(p, t):
+        return t.valid and getpspecspeed(p, 'ref') is not None
+
+    print 'Tree Avg GFLOP/s', np.mean([t.speed() for p, t in mdl_timings.items()])
+    print 'HC Avg GFLOP/s', np.mean([getpspecspeed(p, 'gen75') for p, t in mdl_timings.items() if use_this(p, t)])
+    print 'Grid Avg GFLOP/s', np.mean([getpspecspeed(p, 'grid') for p, t in mdl_timings.items() if use_this(p, t)])
+    print 'Ref Avg GFLOP/s', np.mean([getpspecspeed(p, 'ref') for p, t in mdl_timings.items() if use_this(p, t)])
+
     tree_data = [maybe_log(t.speed() / getpspecspeed(p, 'ref'))
-        for p, t in mdl_timings.iteritems() if t.valid]
+        for p, t in mdl_timings.iteritems() if use_this(p, t)]
     hc_data = [maybe_log(getpspecspeed(p, 'gen75') / getpspecspeed(p, 'ref'))
-        for p, t in mdl_timings.iteritems() if t.valid]
+        for p, t in mdl_timings.iteritems() if use_this(p, t)]
     grid_data = [maybe_log(getpspecspeed(p, 'grid') / getpspecspeed(p, 'ref'))
-        for p, t in mdl_timings.iteritems() if t.valid]
+        for p, t in mdl_timings.iteritems() if use_this(p, t)]
 
     plt.hist([tree_data, hc_data, grid_data],
         label=[
-            'hill climbing (model)',
-            'hill climbing (actual)',
-            'grid (actual)'],
+            'Hill Climbing (model)',
+            'Hill Climbing (real)',
+            'Grid (real)'],
         color=['b', 'g', 'r'],
         bins=20,
         )
@@ -427,8 +437,19 @@ def main_figtrain():
             vline(np.exp(np.mean(np.log(tree_data))), 'b', 1.2, 17)
             vline(np.exp(np.mean(np.log(hc_data))), 'g', 1.55, 18)
             vline(np.exp(np.mean(np.log(grid_data))), 'r', 1.55, 16)
[…]

### 2011-12-07 experiment.py
m

@@ -213,41 +213,32 @@ def main_step():
         ofile.close()
 
 
-def main_dump_logspeedup_dataset():
+def build_logspeed_dataset(wisdomfile, invalid_multiplier, n_train):
     timings = {}
     refspeed = {}
-    try:
-        ofile = open(sys.argv[2])
-        print 'Please remove output file manually:', sys.argv[2]
-        return
-    except IOError:
-        pass
-    for pklfile in sys.argv[3:]:
-        assert len(sys.argv) == 4  # XXX: platformfeatures are not implemented
-        print "Loading data from ", pklfile
-        wdb, results, rng = cPickle.load(open(pklfile))
-        for ii, finding in enumerate(results):
-            ref = finding['ref']
-            ref_orig = finding['ref_orig']
-            key_ii = (ref.prob_spec, pklfile)
 
-            refspeed.setdefault(key_ii,
-                    np.mean([ref.speed(), ref_orig.speed()]))
-            timings.setdefault((ref.prob_spec, pklfile), [])
+    print "Loading data from ", wisdomfile
+    wdb, results, rng = cPickle.load(open(wisdomfile))
+    for ii, finding in enumerate(results):
+        ref = finding['ref']
+        ref_orig = finding['ref_orig']
+        key_ii = (ref.prob_spec, wisdomfile)
 
-            for k, t in finding.iteritems():
-                if not t.valid:
-                    continue
-                if not k.startswith('ref'):
-                    t.src = dict(srcfile=pklfile, pos=ii, key=k)
-                    timings[key_ii].append(t)
+        refspeed.setdefault(key_ii,
+                np.mean([ref.speed(), ref_orig.speed()]))
+        timings.setdefault((ref.prob_spec, wisdomfile), [])
 
-        print "Out of %i results" % len(results)
-        print "Loaded timings for %i problem specs" % len(timings)
+        for k, t in finding.iteritems():
+            if not k.startswith('ref'):
+                t.src = dict(srcfile=wisdomfile, pos=ii, key=k)
+                timings[key_ii].append(t)
 
-    n_keys = len(timings)
-    train_keys = timings.keys()[:int(.75 * n_keys)]
-    test_keys = timings.keys()[int(.75 * n_keys):]
[…]

### 2011-12-07 wisdom.py

@@ -641,6 +641,7 @@ class Timing(object):
             for i in xrange(N):
                 self.measure_1(fop, in_, out_, fb_, in_data, fb_data)
 
+
 def quick_winner(new_timing, timing, device, finding):
     in_data, fb_data, out_dat = new_timing.get_sample_data()
     with CudaContext(device) as context:
@@ -662,6 +663,7 @@ def quick_winner(new_timing, timing, device, finding):
         return new_timing
     else:
         return timing
+
 
 def gcg_grid_autotune(timing, device, finding):
     import fbconv3_cuda_metaparams_cherrypick

### 2011-12-07 paper/draft1.tex
m

@@
 
-\title{Boosted Regression Trees for Model-Free Auto-Tuning}
+\title{Boosted Regression Trees for Data-Driven Auto-Tuning}
 
@@
 \begin{abstract}
-Autotuning is a widely used and effective technique for optimizing a
-parametrized GPU code template for a particular type of computation on
-particular hardware.
-Its drawback is that thorough or exhaustive auto-tuning can be very slow.
+Auto-tuning is a widely used and effective technique for optimizing a
+parametrized GPU code template for a particular computation on particular
+hardware.  Its drawback is that thorough or exhaustive auto-tuning requires
+compiling many kernels and calling each one many times; this process is slow.
 Furthermore, library abstraction boundaries provide operations such as image
 filtering and matrix multiplication, which actually correspond to a large set
-of potential problem configurations with a wide variety of memory
-access patterns and computational bottlenecks.
-How can we draw on data from previous auto-tuning of related problems on related hardware to make a
-good decision on a novel problem?
-This paper presents a machine learning approach to auto-tuning, in
-which features of
-(a) the current hardware platform,
-(b) the kernel configuration
-and (c) the problem instance are used to fit a regression model (boosted regression trees) that predicts
-how much faster an implementation will be than a reference baseline.
-Combinatorial optimization strategies that would normally be applied for auto-tuning can be applied to the regression model to find an implementation much more quickly.
-We validate our approach using the filterbank correlation kernel described in [anon].
-We find that 0.05 seconds of hill climbing on the regression model can achieve
-80\% of the improvement brought by 120 seconds of real auto-tuning.
-Our approach is not specific to filterbank correlation, or even to GPU auto-tuning: the approach of using a non-linear regression model on top of simple features applies to a variety of problem types, kernel types, and platforms.
-A detailed platform model can provide useful features for
-cross-platform auto-tuning, but such a model is not necessary for same-platform autotuning.
+of potential problem configurations with a wide variety of memory access
+patterns and computational bottlenecks.  How can we draw on data from previous
+empirical auto-tuning of related problems on related hardware to make a just-in-time
+implementation decision for a novel problem?  This paper presents a machine
+learning approach to auto-tuning, in which features of (a) the current
+hardware platform, (b) the kernel configuration and (c) the problem instance
+are passed to a regression model (boosted regression trees) which predicts how
+much faster this kernel will be than a reference baseline.  Combinatorial
+optimization strategies that would normally implement auto-tuning by
+evaluating kernel configurations on the real hardware are orders of magnitude
+faster when evaluating the regression model instead.  We validate our approach
+using the filterbank correlation kernel described in [XXX], where we find
+that 0.1 seconds (XXX) of hill climbing on the regression model can achieve an
+average of 95\% of the speed brought by 120 seconds of true auto-tuning XXX.  Our
+approach is not specific to filterbank correlation, or even to GPU
+auto-tuning: the approach of using a non-linear regression model on top of
+simple features applies to a variety of problem types, kernel types, and
+platforms.
 \end{abstract}
@@
 
-\subsection{Problem Space: Filter-bank Correlation}
+\section{Filter-bank Correlation}
 
-Our experiments on automatic auto-tuning will be focused on filterbank
+Following [XXX], the experiments in this paper will focus on filterbank
 correlation.  Mathematically, we define filterbank correlation in terms of an
@@
 channels) that we call its {\em depth}. We index $x$ like $\mathbf{x}[i,j,d]$
-where $1 \leq i \leq R$ and $1 \leq j \leq C$.
-The filterbank $f$ has $K$ square filters of side-length $W$, each one of
-which has $D$ channels.
+where $0 \leq i < R$, $0 \leq j < C$, and $0 \leq d < D$.
+The filterbank $f$ has $F$ filters that are like little images: each has a
+height $H$, a width $W$, and $D$ channels.
+We will restrict ourselves to what are called {\em valid} correlations, in
+which the image is larger in both rows and columns than the filters.
 The result of filterbank correlation of $x$ with $f$ is an image-like array
-$z$ with $R-W+1$ rows, $C-W+1$ columns, and depth $K$, whose elements are
+$z$ with $R-H+1$ rows, $C-W+1$ columns, and depth $F$, whose elements are
 defined according to Equation~\ref{eq:z}:
@@
 \begin{equation}
-    \mathbf{z}[i,j,k] = \sum_{u=1}^{W} \sum_{v=1}^{W} \sum_{d=1}^D
-        \mathbf{x}[i+u, j+v, d]~ \mathbf{f}[k, u, v, d].
+    \mathbf{z}[r,c,k] = \sum_{w=0}^{W-1} \sum_{h=0}^{H-1} \sum_{d=0}^{D-1}
+        \mathbf{x}[r+h, c+h, d]~ \mathbf{f}[k, h, w, d].
         \label{eq:z}
@@
 the choices of $R$, $C$, $D$, $F$, and $W$.
+The computation of filterbank correlation on GPUs was recently treated by
+XXX, which serves as the basis for the present work.
+
+\section{Problem Configuration}
 
 We randomly sampled problem specifications from the following space:
-\begin{figure}
-\begin{center}
-\begin{minipage}{.4\textwidth}
-\begin{verbatim}
-R : one_of(256, 512, 1024, 2048, 4096)
-D : one_of(1, 4, 8, 16, 32, 64, 128, 256)
-F : one_of(1, 4, 8, 16, 32, 64, 128, 256)
-W : one_of(3, 5, 7, 9, 11)
-\end{verbatim}
-\end{minipage}
-\end{center}
-\caption{The full configuration space for filterbank correlation included a
-total of 1600 elements, but we restricted our experiments to problems that
+\begin{align*}
+R = C & \in \{ 256, 512, 1024, 2048, 4096 \} \\
+H = W & \in \{ 3, 5, 7, 9, 11 \} \\
+D &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \} \\
+F &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \}
+\end{align*}
+The problem configuration space used in our experiments. The product space
+includes 1600 elements, but we restricted our experiments to problems that
 represented between 1 and 50 gigaflops. Our problem space included 602
@@
 valid, the number of configurations would be effectively infinite.
-XXX: define R, D, F, W here
-}
-\label{fig:probspace}
-\end{figure}
+
 % Problems were considered invalid if the filters were larger than the images.
@@
 
-\begin{figure}
-\begin{center}
-\begin{minipage}{.43\textwidth}
-\begin{verbatim}
-block_w   : one_of(4, 8, 16, 32, 64, 128)
-block_h   : one_of(4, 8, 16, 32, 64, 128)
-n_filter_r: one_of(1, 2)
-n_output4s: one_of("all", 1, 2)
-spill     : one_of(False, True)
-imul_fast : one_of(False, True)
-pad_shared: one_of(False, True)
-use_tex1d : one_of(False, True)
-maxrreg   : one_of(8, 16, 20, 24, 28, 32, inf)
-fast_math : one_of(False, True)
-\end{verbatim}
-\end{minipage}
-\end{center}
-\caption{The kernel used to perform
+\begin{align*}
+\mathrm{block\_w}    & \in (4, 8, 16, 32, 64, 128) \\
+\mathrm{block\_h}    & \in (4, 8, 16, 32, 64, 128) \\
+\mathrm{n\_filter\_r} & \in (1, 2) \\
+\mathrm{n\_output4s} & \in (\mathrm{all}, 1, 2) \\
+\mathrm{spill}      & \in (False, True) \\
+\mathrm{imul\_fast}  & \in (False, True) \\
+\mathrm{pad\_shared} & \in (False, True) \\
+\mathrm{use\_tex1d}  & \in (False, True) \\
+\mathrm{maxrreg}    & \in (8, 16, 20, 24, 28, 32, \infty) \\
+\mathrm{fast\_math}  & \in (False, True)
+\end{align*}
+
+The kernel used to perform
 filterbank correlation was parametrized by 10 parameters, some of which
@@
 XXX: how to make sense of these paramters without code listing?
-}
-\end{figure}
 
@@
 }
-\label{fig:speedup_295}
+\label{fig:speedup}
+\end{figure*}
+
+\begin{figure*}
+\centering
+\includegraphics[scale=.42]{fig_ntrain_295.pdf}
+\includegraphics[scale=.42]{fig_ntrain_480.pdf}
+\caption{The effect of the invalid configuration score ($a$) and training set
+size on simulated auto-tuning.
+All candidates timed during the grid and
+hill-climbing search procedures were used as training points, so the training
+set sizes ranged from an average of 1500 (10 problem configurations) to 30
+thousand (200 problem configurations).
+Training on 10 or 25 configurations was useful (higher mean than the reference),
+but not as useful as training on 50 or more configurations.
+The results on the GTX 295 suggest that
+moderate values of $a$ between $0.25$ and $0.75$ might be
+best, but $a$ had no significant effect on the GTX 480.
+}
+\label{fig:fig_ntrain}
+\end{figure*}
+
+\begin{figure*}
+\centering
+\includegraphics[scale=.42]{fig_gflop_scatter_295.pdf}
+\includegraphics[scale=.42]{fig_gflop_scatter_480.pdf}
+\caption{Computation speed for novel problem configurations using predictive
+vs. empirical code specialization.
+The scatterplots are roughly symmetric about the axis of equality (main diagonal)
+with occasional outliers, indicating that for most problem configurations the
+predictive approach gives as good a solution as the empirical approach.
+The predictive approach typically requires about 0.1 seconds to suggest an
+implementation, whereas the empirical approach requires about 2 minutes.
+}
+\label{fig:fig_gflop_scatter}
 \end{figure*}
@@
 
+In a complete implementation of Eq.~\ref{eq:z}, there would be parameters
+related to the physical layout (e.g. strides) of $x$, $f$ and $z$ arrays in
+addition to the mathematical parameters of heights and widths and so on,
+so the total number of filterbank correlation computations that a user might
+be able to demand is astronomical.
+A typical coping mechanism would be to cast an arbitrary problem configuration into a more
+standardized form, such as by copying inputs into aligned memory buffers with
+appropriate padding, and then choosing an appropriate blocking strategy for
+the computation. At that point, auto-tuning efforts can be focussed on the
+kernel for each blocked strategy. However, on GPU hardware, the cost of
+aligning the inputs can be relatively large.  In future work we hope to apply
+our model-based technique to auto-tune in full problem configuration space, so
+that we can optimize as much of the computation as possible within the
+predictive auto-tuning framework (XXX name).
+
 \end{document}

### 2011-12-07 experiment.py

@@ -514,9 +514,9 @@ def main_fig1():
 
 
 def main_fig_gflop_scatter():
-    _python, _cmd, wisdomfile, inv_mult = sys.argv
+    _python, _cmd, wisdomfile, inv_mult, n_train = sys.argv
     import matplotlib.pyplot as plt
-    train_result = 'train_results_%s_%s' % (wisdomfile, inv_mult)
+    train_result = 'train_results_%s_%s_%s' % (wisdomfile, inv_mult, n_train)
     wdb, results, rng = cPickle.load(open(wisdomfile))
     mdl_timings = cPickle.load(open(train_result))
 
@@ -539,21 +539,103 @@ def main_fig_gflop_scatter():
 
     if '295' in wisdomfile:
         ubound =  350
+        plt.title('GTX 295')
+        figname = 'fig_gflop_scatter_295.pdf'
     elif '1060' in wisdomfile:
         ubound =  350
+        plt.title('Tesla C1060')
+        figname = 'fig_gflop_scatter_1060.pdf'
     elif '480' in wisdomfile:
         ubound =  850
+        plt.title('GTX 480')
+        figname = 'fig_gflop_scatter_480.pdf'
     elif '8600' in wisdomfile:
         ubound =  100
+        plt.title('8600GT')
+        figname = 'fig_gflop_scatter_8600.pdf'
 
     plt.plot([0, ubound], [0, ubound], c='k', ls='-')
     plt.plot([0, ubound], [0, ubound / 2.0], c='k', ls='--')
     plt.plot([0, ubound/2.0], [0, ubound], c='k', ls='--')
+    plt.text(ubound, ubound, 'equality')
+    plt.text(ubound, ubound/2.0, '2x slower')
+    plt.text(ubound/2.0, ubound, '2x faster')
 
     plt.xlabel('GFLOP/s of empirical auto-tuning')
-    plt.ylabel('GFLOP/s of model-based auto-tuning ')
+    plt.ylabel('GFLOP/s of predictive auto-tuning ')
     #plt.legend(loc='lower left')
-    plt.show()
+    if 0:
+        plt.show()
+    else:
+        plt.savefig(figname)
+
+
+def main_fig_ntrain():
+    _python, _cmd, wisdomfile = sys.argv
+    import matplotlib.pyplot as plt
+    wdb, results, rng = cPickle.load(open(wisdomfile))
+    def getpspecspeed(pspec, k):
+        for r in results:
+            if k in r and r[k].prob_spec == pspec:
[…]

### 2011-12-08 paper/draft1.tex
m

@@
 \usepackage{amssymb}
+\usepackage{natbib}
+\usepackage{myAlgorithm}
+
+\bibliographystyle{abbrvnat}
 
@@
 faster when evaluating the regression model instead.  We validate our approach
-using the filterbank correlation kernel described in [XXX], where we find
+using the filterbank correlation kernel described in \citet{pinto+cox:2011gcg}, where we find
 that 0.1 seconds (XXX) of hill climbing on the regression model can achieve an
@@
 
-Autotuning is a family of empirical techniques for finding the implementation
+Auto-tuning is a family of empirical techniques for finding the implementation
 that minimizes that runtime function, when the platform and the problem
@@
 
-\section{Filter-bank Correlation}
+\cite{volkov+demmel:2008}
+\cite{vuduc:2003} % has some stuff about ML for autotuning in ch. 9
+
+\subsection{Filter-bank Correlation}
 
@@
 The computation of filterbank correlation on GPUs was recently treated by
-XXX, which serves as the basis for the present work.
+\citet{pinto+cox:2011gcg}, which serves as the basis for the present work.
 
-\section{Problem Configuration}
+XXX: TYPICAL OF STENCIL COMPUTATIONS.
 
+``Although the auto-tuning strategy has been successfully applied to libraries,
+generalized stencil kernels are not amenable to packaging as libraries.''
+\cite{kamil+etal:2009}
+
+\cite{datta:2009}
+
+\section{Configuration Spaces}
+
+Recall from the introduction (Eq.~\ref{eq:condopt}) that we are addressing a
+conditional optimization problem, in which we condition on a problem
+configuration ($P$) and a system specification ($S$), and search for the best
+implementation $I$ for addressing problem $P$ on system $S$.
+This section describes the configuration spaces we used in our tests,
+which are described later in Section XXX.
+
+\subsection{Problem Description Space}
+
+Mathematically, filterbank correlation is an operation on an $R \times C \times
+D$ image and a $F \times H \times W \times D$ filterbank.
 We randomly sampled problem specifications from the following space:
@@
 \end{align*}
-The problem configuration space used in our experiments. The product space
+The product space
 includes 1600 elements, but we restricted our experiments to problems that
 represented between 1 and 50 gigaflops. Our problem space included 602
-configurations. In a full application in which all image and filter sizes are
-valid, the number of configurations would be effectively infinite.
+configurations. A library implementation that must support all image and
+filter sizes must handle on the order of $2^{50}$ forseable
+problem configurations (not even including variations due to strided memory
+layouts).
 
@@
 that were smaller do not fully utilize the hardware, and are handled equally
-well by many kernel settings, including the reference (XXX cite GCG).
+well by many kernel settings.
 Problems that were larger take so long to evaluate that there is negligible
@@
 a greater variety in the kinds of computations and memory transfers to
-perform.  In our experiments, we only consider the problem variations
-described in Figure ~\ref{tbl:}, which already suffice to show that automatic
-auto-tuning is a promising approach.
+perform.
 
-\subsection{Kernel Space: Metaprogramming}
+\subsection{Implementation Description Space}
+
+The strategy we use for computing filterbank correlation on the GPU
+using CUDA follows \citet{pinto+cox:2011gcg}.
+The overall strategy is to load the filterbank into some relatively fast
+memory (constant or texture), and then launch a grid of blocks that each
+computes some local piece of the output image $z$.
+Each thread computes $4 \times n\_output\_4s$ channels for some column and row of $z$.
+Each block computes a subrectangle of the output $z$,
+and the grid is used to tile the whole output image with blocks.
+Some filterbanks are too large to fit into the fast memory.
+For large filterbanks our approach permits splitting the filterbanks along
+the number-of-filters dimension ($F$) and the number-of-rows dimension ($H$).
+All the filterbanks in our study are small enough that at least half of each
+filter can always fit into the constant or texture memory.
+When splitting the filterbank in this way we must use several kernel
+executions, each of which increments the output buffer by a partial result.
+Pseudo-code for the kernel is given in Figure~\ref{fig:kernel}.
+
+\begin{figure}[!ht]
+\centerline{
+\begin{algorithm}{$\Algo{fbcorr}\big(gX, cF, gZ \big)$}
+\Aitem shared $sX \setto$ all channels of region ($\beta$) of $gX$
+\Aitem $x, y \setto$ position of this thread in output image
+\Aitem \_\_syncthreads()
+\Aitem $v[0:N] \setto 0$, for $N=4\times n\_output\_4s$
+\Aitem for $d \setto 0$~\To~$D$,
+\Aitem ~ for $h \setto 0$~\To~$H / n\_filter\_r$,
+\Aitem ~ ~ for $w \setto 0$~\To~$W$,
+\Aitem ~ ~ ~ $u \setto sX[x+h, y+w, d]$
+\Aitem ~ ~ ~ for $n \setto 0$~\To~$n\_output\_4s - 1$,
+\Aitem ~ ~ ~ ~  $v[n] \setto v[n] + cF[n, h, w, d]$
+\Aitem for $n \setto 0$~\To~$n\_output\_4s - 1$,
+\Aitem ~  gZ[x][y][4n:4n+n] = v[4n:4n+n], (float4)
+\end{algorithm}
+}
+\caption{Kernel pseudo-code for filterbank correlation.
+Input $gX$ is a pointer to $x$ in global memory, input $cF$ is a pointer to
+$f$ in either constant or texture memory, and output $gZ$ is a pointer to $z$
+in global memory.
+Each block of threads modifies $4 \times n\_output\_4s$ channels of a
+rectangle (called $\beta$ in code listing) within $z$. A grid of blocks covers all rows and columns of $z$.
+Multiple calls can be used to apply all filters of a large filterbank $f$ to
+$x$.
+}
+\label{fig:kernel}
+\end{figure}
 
@@
 \mathrm{n\_filter\_r} & \in (1, 2) \\
-\mathrm{n\_output4s} & \in (\mathrm{all}, 1, 2) \\
+\mathrm{n\_output\_4s} & \in (\mathrm{all}, 1, 2) \\
 \mathrm{spill}      & \in (False, True) \\
@@
 
-\iffalse
-\subsection{Platform Space: XXX}
+Meta-programming was implemented in Python using
+Cheetah for string processing~\citep{cheetah}
+and PyCUDA~\citep{klochner+etal:2009} for dynamic kernel compilation.
 
-XXX: GCG chapter pg 13 notes that among the grid search, the best parameter
-setting is different for the various platforms.
+XXX: analysis of READ traffic, WRITE traffic, DATA re-use and CACHEing
+strategies. FLOPS per write etc.
 
-XXX: so far we only tried 3 cards... so we're using a 1-of-N feature.
-\fi
+XXX: Why not use FFT: convolution in spectral domain?
 
-\section{Results}
+\subsection{System Description Space}
+
+ XXXX Leave for future work?
+
+\section{Auto-Tuning by Hill Climbing}
 
@@
 
+Platform Space: XXX
+
+XXX: GCG chapter pg 13 notes that among the grid search, the best parameter
+setting is different for the various platforms.
+
+XXX: so far we only tried 3 cards... so we're using a 1-of-N feature.
+
+%\renewcommand{\bibsection}{\section*{Bibliography}}
+%\setlength{\bibsep}{5pt}
+\bibliography{local}
 \end{document}

### 2011-12-08 bandit.py

@@ -44,15 +44,18 @@ class FBCorr3Bandit(Bandit):
         x = theano.tensor.nnet.conv.conv2d(imgs, filts,
                 img_shp, k_shp, border_mode='valid')
         f = theano.function([], [], updates=[(outs, x)])
-        for i in range(3):
+        dt = float('inf')
+        for i in range(5):
             t = time.time()
             f()
-            dt = time.time() - t
+            dt = min(dt, time.time() - t)
 
         fsize = probspec['fsize']
         depth = probspec['depth']
         gflop = outs.get_value(borrow=True).size * (fsize * fsize * depth * 2) / (1000.**3.)
-        print 'GFLOP/S', gflop / dt
+        rval =  gflop / dt
+        print 'GFLOP/S', rval
+        return rval
 
 
     @classmethod

### 2011-12-08 experiment.py

@@ -638,49 +638,95 @@ def main_fig_ntrain():
         plt.savefig(figfile)
 
 
-def main_fig_time():
-    _python, _cmd, wisdomfile = sys.argv
-    wdb, results, rng = cPickle.load(open(wisdomfile))
+def main_fig_gflop_scatter_theano():
+    #
+    # This script, if it runs on the 295 shows that theano is about 1/3 the
+    # speed on average. On the 480 Theano is more like 1/5 the speed.
+    #
+    _python, _cmd, wisdomfile, Nstr = sys.argv
     import matplotlib.pyplot as plt
-    for key, col, marker in [
-            ('rand25', (.50, 0, 0), '+'),
-            ('rand50', (.75, 0, 0), '+'),
-            ('rand75', (.99, 0, 0), '+'),
-            ('gen25',  (0, .50, 0), '+'),
-            ('gen50',  (0, .75, 0), '+'),
-            ('gen75',  (0, .99, 0), '+'),
-            ('tree25', (0, 0, .50), '+'),
-            ('tree50', (0, 0, .75), '+'),
-            ('tree75', (0, 0, .99), '+'),
-            ('wise25', (0, .50, .50), '+'),
-            ('wise50', (0, .75, .75), '+'),
-            ('wise75', (0, .99, .99), '+'),
-            ('grid',   (0,  0, 0), '+'),
-            ]:
-        def getspeed(r, k='ref'):
-            if k in r:
-                return r[k].prob_spec.gflops() / r[k].speed()
-            else:
-                return None
-        a = np.asarray([getspeed(r, key) for r in results]).astype('float')
-        a_orig = np.asarray([getspeed(r, key + '_orig')
-            for r in results]).astype('float')
-        plt.scatter(np.arange(len(a)).astype('float'),
-                a,
-                label=key, c=col, marker=marker)
-        if 0:
-            plt.scatter(np.arange(len(a)).astype('float')+.25,
-                    a_orig,
-                    #label=key,
-                    c=col, marker=marker)
-        #gmean = np.exp( np.log(a/b).mean())
-        #plt.axhline(gmean, c=col)
+    import bandit
+    wdb, results, rng = cPickle.load(open(wisdomfile))
 
-    plt.xlabel('random trial')
-    plt.ylabel('time')
-    plt.legend(loc='lower left')
-    plt.show()
+    ref_speeds = []
+    gen75_speeds = []
+    theano_speeds = []
+    grid_speeds = []
 
[…]

### 2011-12-08 wisdom.py

@@ -82,7 +82,7 @@ def random_op_spec(rng):
     dct = rSON2(
         "block_w" , one_of(4, 8, 16, 32, 64, 128),
         "block_h" , one_of(4, 8, 16, 32, 64, 128),
-        "n_filter_rows" , one_of(1, 2),
+        "n_filter_rows" , one_of(1, 2), #XXX: turns out this can be "all" as well
         "n_output4s" , one_of("all", 1, 2),
         "spill" , one_of(False, True),
         "imul_fast" , one_of(False, True),

### 2011-12-09 paper/draft1.tex
m

@@
 
+%XXX:  Machine Learning for Auto-tuning: A Data-Drive Approach...} ???
 \title{Boosted Regression Trees for Data-Driven Auto-Tuning}
@@
 
+%XXX: DIAGRAM OF IMAGE, FILTERBANK, OUTPUT
+
 % Problems were considered invalid if the filters were larger than the images.
@@
 using CUDA follows \citet{pinto+cox:2011gcg}.
-The overall strategy is to load the filterbank into some relatively fast
-memory (constant or texture), and then launch a grid of blocks that each
-computes some local piece of the output image $z$.
+The overall strategy is to load the filterbank into constant memory, which is
+relatively fast and visible to all threads, and then launch a grid of blocks
+that tiles the output image.
 Each thread computes $4 \times n\_output\_4s$ channels for some column and row of $z$.
-Each block computes a subrectangle of the output $z$,
-and the grid is used to tile the whole output image with blocks.
-Some filterbanks are too large to fit into the fast memory.
-For large filterbanks our approach permits splitting the filterbanks along
-the number-of-filters dimension ($F$) and the number-of-rows dimension ($H$).
-All the filterbanks in our study are small enough that at least half of each
-filter can always fit into the constant or texture memory.
-When splitting the filterbank in this way we must use several kernel
-executions, each of which increments the output buffer by a partial result.
+Each block computes $4 \times n\_output\_4s$ channels for a subrectangle of the output image ($z$).
+When there are more than $4 \times n\_output\_4s$ channels in $z$, or if the
+filterbank is too large to fit into constant memory, then multiple
+kernel executions perform the full computation.
+Our approach permits splitting the filterbank along
+the number-of-filters dimension ($F$) and the height dimension ($H$).
+All the filterbanks in our study are small enough that
+at least one row of a single filter can fit into constant memory.
 Pseudo-code for the kernel is given in Figure~\ref{fig:kernel}.
@@
 
+The kernel is parametrized by 10 parameters:
 \begin{align*}
+\mathrm{block\_h}    & \in (4, 8, 16, 32, 64, 128) \\
 \mathrm{block\_w}    & \in (4, 8, 16, 32, 64, 128) \\
-\mathrm{block\_h}    & \in (4, 8, 16, 32, 64, 128) \\
 \mathrm{n\_filter\_r} & \in (1, 2) \\
@@
 \end{align*}
+
+The block height (block\_h) and block width (block\_w) parameters control the
+number of threads that run on each multiprocessor.
+The number of filter rows (n\_filter\_r) is how many rows of each filter is
+loaded
 
@@
 XXX: Why not use FFT: convolution in spectral domain?
+
+XXXX: Talk about memory transfer requirements vs. speed vs. copy...
 

### 2011-12-09 experiment.py

@@ -729,7 +729,24 @@ def main_theano_conv_slow_so_sad():
     print foo.vs_theano()
 
 
+def main_for_memcheck():
+    _python, _cmd, dev_id_str = sys.argv
+    device = init_cuda(int(dev_id_str))
+    rng = np.random.RandomState()
+    pgen = problem_generator(rng)
+    while True:
+        prob_spec = pgen.next()
+        print prob_spec
 
+        timing = wisdom.Timing(prob_spec, wisdom.reference_op_spec())
+        timing.measure(device)
+        if not timing.valid:
+            continue
+
+        timing = wisdom.Timing(prob_spec, wisdom.random_op_spec(rng))
+        timing.measure(device)
+        if not timing.valid:
+            continue
 
 if __name__ == '__main__':
     cmd = sys.argv[1]

### 2011-12-09 wisdom.py

@@ -93,6 +93,8 @@ def random_op_spec(rng):
         ).sample(rng)
     return OpSpec(**dct)
 
+# XXX: n_output4s should offer more choices!
+# XXX: maxrregcount should go higher: 64? 128?
 
 def random_op_cross(op1, op2, rng, r=.5):
     return OpSpec(

### 2011-12-09 wisdom.py
dotransfer

@@ -562,14 +562,15 @@ class Timing(object):
             }
         return gflops_cuda[key]
 
-    def measure_1(self, fop, in_, out_, fb_, in_data, fb_data):
+    def measure_1(self, fop, in_, out_, fb_, in_data, fb_data, dotransfer=True):
         # -- upload data
         try:
             start = time.time()
-            in_[:] = in_data
-            # XXX: is writing a 0 here important for correctness?
-            out_[:] = 0
-            fb_[:] = fb_data
+            if dotransfer:
+                in_[:] = in_data
+                # XXX: writing 0 here is important for correctness!
+                out_[:] = 0
+                fb_[:] = fb_data
             end = time.time()
             t_upload = end - start
 
@@ -586,7 +587,10 @@ class Timing(object):
             t_process = end - start
 
             start = time.time()
-            out_data = out_[:]
+            if dotransfer:
+                out_data = out_[:]
+            else:
+                out_data = None
             end = time.time()
             t_download = end - start
 
@@ -658,7 +662,8 @@ def quick_winner(new_timing, timing, device, finding):
         while i < 10 and new_timing.valid:
             if i >= 3 and new_timing.speed('max') < .5 * timing.speed():
                 break
-            new_timing.measure_1(fop, in_, out_, fb_, in_data, fb_data)
+            new_timing.measure_1(fop, in_, out_, fb_, in_data, fb_data,
+                    dotransfer=(i == 0))
             i += 1
     finding['qw_%i' % len(finding)] = new_timing
     if new_timing.valid and new_timing.speed() > timing.speed():

### 2011-12-09 paper/draft1.tex
paper

@@
 evaluating kernel configurations on the real hardware are orders of magnitude
-faster when evaluating the regression model instead.  We validate our approach
+faster when evaluating the surrogate regression model instead.  We validate our approach
 using the filterbank correlation kernel described in \citet{pinto+cox:2011gcg}, where we find
-that 0.1 seconds (XXX) of hill climbing on the regression model can achieve an
-average of 95\% of the speed brought by 120 seconds of true auto-tuning XXX.  Our
+that 0.1 seconds (XXX) of hill climbing on the regression model ({\em
+predictive}
+auto-tuning) can achieve an
+average of 95\% of the speed brought by 120 seconds of empirical auto-tuning XXX.  Our
 approach is not specific to filterbank correlation, or even to GPU
-auto-tuning: the approach of using a non-linear regression model on top of
+kernel auto-tuning: the approach of using a non-linear regression model on top of
 simple features applies to a variety of problem types, kernel types, and
@@
 \citet{pinto+cox:2011gcg}, which serves as the basis for the present work.
+
+In terms of floating point operations, a filterbank correlation requires the
+inner sums to be computed for each output pixel, yielding the quantity in
+Eq.~\ref{eq:flops}:
+\begin{equation}
+\mathrm{FLOPS} = 2  K  H  W  D  (R - H + 1)( C- W + 1)
+\label{eq:flops}
+\end{equation}
+The multiplicative factor of 2 arises because we must multiply an element of
+$x$ with an element of $f$ and then add the result to an element of $z$.
+
+In terms of bandwidth, if we were to assume (optimistically) that the entire
+filterbank fit into the GPU's so-called constant memory, then the bandwidth
+required for a filterbank correlation would be the size of the filterbank,
+plus the size of the input image, plus the size of the output image. This sum
+is expressed in terms of configuration parameters in Eq.~\ref{eq:bytes}:
+\begin{align}
+\mathrm{Bytes} = 4&RCD \nonumber \\
+& + 4KHWD \nonumber \\
+& + 4(R-H+1)(C-W+1)K
+\label{eq:bytes}
+\end{align}
+We will see the bandwidth requirements of a more realistic analysis in
+Section~XXX.
 
@@
 \Aitem for $n \setto 0$~\To~$n\_output\_4s - 1$,
-\Aitem ~  gZ[x][y][4n:4n+n] = v[4n:4n+n], (float4)
+\Aitem ~  gZ[x][y][4n:4n+n] += v[4n:4n+n], (float4)
 \end{algorithm}
@@
 
-The block height (block\_h) and block width (block\_w) parameters control the
-number of threads that run on each multiprocessor.
-The number of filter rows (n\_filter\_r) is how many rows of each filter is
-loaded
+The block height (``block\_h'') and block width (``block\_w'') parameters control the
+number of threads that run within each block.
+Each kernel call loads some number of filter rows (``n\_filter\_r'') into
+constant memory and processes the correlation of the image with just those
+rows, incrementing the output buffer.
+Each thread can compute several output elements
+at once, in multiples (``n\_output\_4s'') of 4;
+this increases the efficiency of each thread, but can lead to lower occupancy.
+Registers are a precious commodity on the GPU, and this kernel accumulates
+elements of $z$ in registers.
+The ``spill'' parameter controls whether the current thread's output position in
+$z$ is stored in a register (faster access) or in shared memory (frees up
+a register).
+The ``imul\_fast'' parameter controls whether integer multiplication is done
+in 24-bit (True) or 32-bit (False) precision.
+The ``pad\_shared'' parameter controls whether the $sX$ shared memory buffer
+is padded, which wastes space in shared memory but reduces bank conflicts.
+The ``use\_tex1d'' parameter controls whether the image is loaded into shared
+memory with global pointer dereferences or texture fetches.
+The ``maxrreg'' and ``fast\_math'' parameters are passed to the nvcc compiler to limit
+the number of registers available to each thread, and to enable XXX,
+respectively.
+%XXX: what effect does fast-math have here?
+
+Often however, the entire filterbank often does not fit into the GPU's
+constant memory and so $P$ passes are necessary to compute $z$. In such cases,
+the number of bytes moved is scaled almost linearly in $P$:
+\begin{align}
+\mathrm{Bytes} = 4&RCDP \nonumber \\
+& + 4KHWD \nonumber \\
+& + 8(R-H+1)(C-W+1)KP
+% XXX: incrementing requires reading and writing $z$ right? so actually the 4
+% above should be an 8? (see related ticket on github for potential
+% optimization about this)
+\label{eq:bytesP}
+\end{align}
+
+Filterbank correlation represents a difficult code-optimization problem
+because of the number of interacting constraints: a large value of $P$ reduces
+arithmetic density, but a small value of $P$ requires that each thread do more
+work, and makes global memory access latency harder to hide. (XXX: is there a
+better reason?)
 

### 2011-12-09 paper/draft1.tex
paper

@@
 \usepackage{myAlgorithm}
+\usepackage{xcolor}
 
 \bibliographystyle{abbrvnat}
+
+\definecolor{darkred}{rgb}{.5,0,0}
 
@@
 
+Filterbank correlation is a simple spatial image filtering operation that is
+an important subroutine in many image processing applications. It has a relatively
+high arithmetic intensity which makes it a natural fit for GPU platforms.
+One might expect then, that it would be easy to implement a library providing
+this operation -- this hypothetical library could have quite a simple interface
+following the spirit of FFTW or a BLAS routine:
+\begin{verbatim}
+    gpu_fbcorr(
+        img_shape, filters_shape,
+        img, filters, output)
+\end{verbatim}
+However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
+related stencil operations XXX, it is challenging to provide an implementation
+or even an implementation strategy that provides satisfactory performance
+across the range of inputs (shapes, physical layouts) that occur even in
+typical usage.
+
+Current state-of-the art is to do empirical auto-tuning.
+XXX \vspace{12pt}
+
+Increasingly, machine learning and statistical inference strategies are
+finding application in compiler technology.
+XXX \vspace{12pt}
+
+This work shows that online auto-tuning can be accelerated by orders of
+magnitude by using a regression model built offline as a surrogate for actual
+computations on the real hardware.
+We use a powerful non-linear regression model (boosted decision trees)
+to approximate the empirical relationship between three sets of variables ($\mathcal{A},
+\mathcal{B}, \mathcal{C}$) and the observed speed-up over a
+reference implementation:
+\begin{description}
+\item[$\mathcal{A}$:] task description (e.g. shapes, physical layout)
+\item[$\mathcal{B}$:] algorithm description (i.e. auto-tuning parameters)
+\item[$\mathcal{C}$:] platform description (e.g. micro-benchmarks)
+\end{description}
+Ideally, \texttt{gpu\_fbcorr} would apply the best
+available implementation
+
+\begin{figure}[!ht]
+\centerline{
+\begin{algorithm}{$\Algo{\color{darkred}{gpu\_fbcorr\_empirical}}\big(shapes, strides \big)$}
+\Aitem $a \setto TaskFeatures(shapes, strides)$
+\Aitem $c \setto PlatformFeatures()$
+\Aitem $b^* \setto \operatorname*{argmin}_{b \in \mathcal{B}} MeasureTime(a, b, c)$
+\algoremark{\color{darkred}{slow}}
+\Aitem \Return $b^*$
+\end{algorithm}
+}
+\begin{algorithm}{$\Algo{\color{darkred}{gpu\_fbcorr\_predictive}}\big(shapes, strides \big)$}
+\Aitem $a \setto TaskFeatures(shapes, strides)$
+\Aitem $c \setto PlatformFeatures()$
+\Aitem $f \setto LoadTimingModel()$
+\Aitem $b^* \setto \operatorname*{argmin}_{b \in \mathcal{B}} f(a, b, c)$
+\algoremark{\color{darkred}{fast}}
+\Aitem \Return $b^*$
+\end{algorithm}
+\caption{Pseudo-code template for an auto-tuning implementation of
+$\texttt{gpu\_fbcorr}$. Empirical auto-tuning (above) is inevitably slow
+because dynamically-generated code must be compiled and run on a number of actual-size inputs.
+Predictive auto-tuning (below) can be orders of magnitude faster. We show that it can
+also be accurate.
+}
+\label{fig:argmin}
+\end{figure}
+
+machine learning
+
+This paper shows that a kernel with several  boosted
+
 There is a natural function from
@@
 
-Following [XXX], the experiments in this paper will focus on filterbank
+Following \citet{pinto+cox:2011gcg}, the experiments in this paper will focus on filterbank
 correlation.  Mathematically, we define filterbank correlation in terms of an
@@
 optimal strategy for doing so often depends on the shape of the problem, i.e.
-the choices of $R$, $C$, $D$, $F$, and $W$.
-The computation of filterbank correlation on GPUs was recently treated by
-\citet{pinto+cox:2011gcg}, which serves as the basis for the present work.
+the values of $R$, $C$, $D$, $F$, and $W$.
 
@@
 Section~XXX.
+
+In terms of artithmetic density, it's approximately
+\begin{align}
+\mathrm{intensity} & \approx \frac {KDHW} {2(D+K)},
+\end{align}
+because each $K$ output writes corresponds to approximately $D$ input reads
+and an inner product between $KDHW$ elements.
 
@@
 \centerline{
-\begin{algorithm}{$\Algo{fbcorr}\big(gX, cF, gZ \big)$}
+\begin{algorithm}{$\Algo{thread\_fbcorr}\big(gX, cF, gZ \big)$}
 \Aitem shared $sX \setto$ all channels of region ($\beta$) of $gX$
@@
 
-Often however, the entire filterbank often does not fit into the GPU's
-constant memory and so $P$ passes are necessary to compute $z$. In such cases,
-the number of bytes moved is scaled almost linearly in $P$:
+When the entire filterbank often does not fit into the GPU's
+constant memory and so $$P=\frac{KH}{4 \cdot n\_output\_4s \cdot
+n\_filter\_r}$$ passes are necessary to compute $z$. In such cases,
+the number of bytes moved is actually more like:
 \begin{align}
@@
 & + 4KHWD \nonumber \\
-& + 8(R-H+1)(C-W+1)KP
+& + 8(R-H+1)(C-W+1)KP.
+\nonumber
 % XXX: incrementing requires reading and writing $z$ right? so actually the 4
@@
 \end{align}
+These passes make the I/O requirements increase quadratically in $K$ and $H$.
+At the same time, the total number of floating-point operations (Eq.~\ref{eq:flops})
+is quadratic in $H$ and $W$. In our experiments, we only considered square
+filters so in our setting the total number of flops is proportional to $H^4$.
+% XXX WHERE IS ALL THIS GOING?
+
+Best arithmetic density
+
+A bandwidth analysis that takes the blocking structure into account would
+reveal that larger $H$ and $W$ require each block to read a larger input
+region, and store it in a larger shared memory region, which reduces the
+potential for high occupancy.
 
@@
 XXXX: Talk about memory transfer requirements vs. speed vs. copy...
-
-\subsection{System Description Space}
-
- XXXX Leave for future work?
 

### 2011-12-09 paper/draft1.tex
paper

@@
 computations on the real hardware.
-We use a powerful non-linear regression model (boosted decision trees)
+We use a powerful non-linear regression model (boosted regression trees)
 to approximate the empirical relationship between three sets of variables ($\mathcal{A},
@@
 
+\section{Boosted Regression Trees}
+
+The regression tree is a piece-wise constant function
+
+Boosted regression trees are
+
 \subsection{Model-based Autotuning}
 
-On the GTX 295, a single decision-tree fit to the log-speed-multiplier
+On the GTX 295, a single regression-tree fit to the log-speed-multiplier
 (LSM) function (XXX: how many problem configurations, how much vs. what

### 2011-12-09 experiment.py

@@ -16,7 +16,7 @@ from hyperopt.ht_dist2 import one_of, rSON2
 import fbconv3_cuda
 
 logger = logging.getLogger(__name__)
-logging.basicConfig(level=logging.INFO)
+logging.basicConfig(level=logging.DEBUG)
 
 from pycuda import driver
 def init_cuda(dev_id):
@@ -107,6 +107,7 @@ def main_step():
         print prob_spec
         wdb.build_dtree(rng=rng, force=True)
         wdb.print_dtree()
+        print>> sys.stderr,  "ASDF"
 
         finding = {}
         def add_finding(k, timing):
@@ -126,7 +127,9 @@ def main_step():
                     print "ERROR: INVALID TEST TIMING", k
 
         # don't reuse test_timing, it would be cheating
+        print>> sys.stderr,  "NEW TIMING"
         timing = wisdom.Timing(prob_spec, wisdom.reference_op_spec())
+        print>> sys.stderr,  "MEASURE"
         timing.measure(device)
         if not timing.valid:
             continue
@@ -709,14 +712,14 @@ def main_theano_conv_slow_so_sad():
             n_imgs=1,
             height=32 ,
             width=32 ,
-            depth=32,
-            n_filters=64,
+            depth=4,
+            n_filters=8,
             filter_height=5,
             filter_width=5,
             img_strides=None,
             filter_strides=None,
             border_mode='valid')
-    prob_spec.n_imgs = 64
+    prob_spec.n_imgs = 1
     print prob_spec.gflops()
     foo = bandit.FBCorr3Bandit(
             prob_spec.n_imgs,

### 2011-12-09 fbconv3_cuda.py

@@ -8,7 +8,7 @@ import os
 import numpy as np
 import logging
 log = logging.getLogger(__name__)
-log.setLevel(logging.WARN)
+log.setLevel(logging.DEBUG)
 
 # CUDA
 from pycuda import driver, gpuarray, compiler, tools
@@ -337,7 +337,10 @@ class Input(object):
         except driver.MemoryError, err:
             raise InvalidConfig(err)
 
-        self._garr_tmp = driver.pagelocked_empty(self._garr_l[0].shape, self.dtype)
+        if 0: # DEBUG XXX
+            self._garr_tmp = driver.pagelocked_empty(self._garr_l[0].shape, self.dtype)
+        else:
+            self._garr_tmp = np.empty(self._garr_l[0].shape, self.dtype)
         self._arr_tmp = np.empty((ngarrs,)+self._padded_shape[:2]+(self._garr_l[0].shape[-1],), dtype='float32')
 
     # -------------------------------------------------------------------------

### 2011-12-09 wisdom.py

@@ -1,3 +1,4 @@
+import sys
 import logging
 import time
 import fbconv3_cuda
@@ -7,7 +8,7 @@ from hyperopt.ht_dist2 import one_of, rSON2
 import pycuda.driver
 
 logger = logging.getLogger(__name__)
-logger.setLevel(logging.INFO)
+logger.setLevel(logging.DEBUG)
 
 class CudaContext(object):
     def __init__(self, device):
@@ -564,6 +565,8 @@ class Timing(object):
 
     def measure_1(self, fop, in_, out_, fb_, in_data, fb_data, dotransfer=True):
         # -- upload data
+        print "FOOO"
+        sys.stdout.flush()
         try:
             start = time.time()
             if dotransfer:
@@ -616,26 +619,34 @@ class Timing(object):
         return in_data, fb_data, out_data
 
     def measure_setup(self, context):
+        print>> sys.stderr,  "FOOO AAA "
         img_shp = self.prob_spec.image_shape()
         ker_shp = self.prob_spec.filters_shape()
         out_shp = self.prob_spec.output_shape()
 
+        print>> sys.stderr,  "FOOO INPUT STUFF"
         in_ = fbconv3_cuda.Input(*img_shp)
         fb_ = fbconv3_cuda.Filterbank(*ker_shp)
         out_ = fbconv3_cuda.Output(*out_shp)
 
         # -- set-up operation (i.e. compilation)
         fb_[:] = 0
+        print>> sys.stderr,  "FOOO FilterOp "
         fop = self.op_spec.FilterOp(in_, fb_, out_, ctxt=context)
+        print>> sys.stderr,  "FOOO FilterOp  RETURNED"
         return fop, in_, fb_, out_
 
     def measure(self, device, N=8):
         """
         Add N timing measurements to self.timings
         """
+        print>> sys.stderr,  "pre sample "
         in_data, fb_data, out_dat = self.get_sample_data()
+        print>> sys.stderr,  "post sample"
+        import pdb; pdb.set_trace()
 
         with CudaContext(device) as context:
+            print>> sys.stderr,  "GOT contextlD"
             try:
                 fop, in_, fb_, out_ = self.measure_setup(context)
             except (fbconv3_cuda.InvalidConfig,

### 2011-12-11 paper/draft1.tex
paper

@@
 
-Filterbank correlation is a simple spatial image filtering operation that is
-an important subroutine in many image processing applications. It has a relatively
-high arithmetic intensity which makes it a natural fit for GPU platforms.
-One might expect then, that it would be easy to implement a library providing
-this operation -- this hypothetical library could have quite a simple interface
-following the spirit of FFTW or a BLAS routine:
-\begin{verbatim}
-    gpu_fbcorr(
-        img_shape, filters_shape,
-        img, filters, output)
-\end{verbatim}
-However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
-related stencil operations XXX, it is challenging to provide an implementation
-or even an implementation strategy that provides satisfactory performance
-across the range of inputs (shapes, physical layouts) that occur even in
-typical usage.
-
-Current state-of-the art is to do empirical auto-tuning.
-XXX \vspace{12pt}
-
-Increasingly, machine learning and statistical inference strategies are
-finding application in compiler technology.
-XXX \vspace{12pt}
-
-This work shows that online auto-tuning can be accelerated by orders of
-magnitude by using a regression model built offline as a surrogate for actual
-computations on the real hardware.
-We use a powerful non-linear regression model (boosted regression trees)
-to approximate the empirical relationship between three sets of variables ($\mathcal{A},
-\mathcal{B}, \mathcal{C}$) and the observed speed-up over a
-reference implementation:
-\begin{description}
-\item[$\mathcal{A}$:] task description (e.g. shapes, physical layout)
-\item[$\mathcal{B}$:] algorithm description (i.e. auto-tuning parameters)
-\item[$\mathcal{C}$:] platform description (e.g. micro-benchmarks)
-\end{description}
-Ideally, \texttt{gpu\_fbcorr} would apply the best
-available implementation
+Current languages and software engineering practices have conspired to make it difficult
+to provide high-performance libraries for general mathematical operations on GPU devices.
+The dominant approach to implementing high-performance GPU codes has been auto-tuning,
+which is fundamentally slow. Compilation with auto-tunin requires running a large number of
+sub-compilations and representative computations to actually {\em measure} which
+of several possible implementations is fastest. General-purpose libraries such
+as CUBLAS and CUFFT XXX hide this overhead by auto-tuning offline and using
+intimate knowledge of the mathematical problem being solved as well as the
+kernel implementations to heuristically choose a good computation strategy quickly when a
+user asks for e.g. a matrix product of particular arguments.
+Designing and testing libraries of this form is time-consuming, difficult, and provides little
+reusable infrastructure to assist in the implementation of different computations.
+Libraries for stencil computations and sparse matrix operations must start from scratch.
 
@@
 }
-\label{fig:argmin}
+\label{fig:general_autotune}
 \end{figure}
+
+This work shows that auto-tuning can be accelerated by orders of
+magnitude by using a regression model built offline as a surrogate for actual
+computations on the real hardware.
+The general form of an auto-tuning based library routine is illustrated in Figure~\ref{fig:general_autotune} (top).
+We will use filterbank correlation (Section~\ref{sec:fbcorr}) as our running example, but the approach is both general and powerful.
+An auto-tuning based routine must operate on three sets of variables:
+\begin{description}
+\item[$\mathcal{A}$:] task description (argument shapes, physical layout)
+\item[$\mathcal{B}$:] implementation description (auto-tuning parameters)
+\item[$\mathcal{C}$:] platform description (compute capability, micro-benchmarks)
+\end{description}
+The hypothetical auto-tuning routine described at the top of Figure~\ref{fig:general_autotune} 
+might take many minutes or hours to perform the argmin at step 3 (during which time it computes the desired result many times!)
+so it would be all but unusable as a normal library routine.
+However, the form of the auto-tuning routine suggests the potential for enormous acceleration:
+if only there were a fast (even approximate) surrogate for the costly $MeasureTime(\cdot)$ function,
+then the argmin could be done in a fraction of a second and the routine could be used normally (Figure~{\ref:fig:general_autotune}, bottom).
+
+Our contribution of the present work is to show that boosted regression tree models,
+a powerful machine learning technique for function approximation, can meet the
+requirements of this application.
+We call this approach {\em predictive} auto-tuning to contrast it with the standard measurement-based approach which we will call {\em empirical} auto-tuning.
+We show that the kernel from the GCG chapter can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning. XXX numbers:
+how good, how fast can prediction be done how much training data is necessary?
+
+\section{Filterbank Correlation}
+\label{sec:fbcorr}
+
+Filterbank correlation is a simple spatial image filtering operation that is
+an important subroutine in many image processing applications. It has a relatively
+high arithmetic intensity which makes it a natural fit for GPU platforms.
+One might expect then, that it would be easy to implement a library providing
+this operation -- this hypothetical library could have quite a simple interface
+following the spirit of FFTW or a BLAS routine:
+\begin{verbatim}
+    gpu_fbcorr(
+        img_shape, filters_shape,
+        img, filters, output)
+\end{verbatim}
+However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
+related stencil operations XXX, it is challenging to provide an implementation
+or even an implementation strategy that provides satisfactory performance
+across the range of inputs (shapes, physical layouts) that occur even in
+typical usage.
+
+Current state-of-the art is to do empirical auto-tuning.
+XXX \vspace{12pt}
+
+Increasingly, machine learning and statistical inference strategies are
+finding application in compiler technology.
+XXX \vspace{12pt}
 

### 2011-12-11 paper/draft1.tex
paper

@@
 \centerline{
-\begin{algorithm}{$\Algo{\color{darkred}{gpu\_fbcorr\_empirical}}\big(shapes, strides \big)$}
+\begin{algorithm}{$\Algo{\color{darkred}{autotune\_empirical}}\big(shapes, strides \big)$}
 \Aitem $a \setto TaskFeatures(shapes, strides)$
@@
 }
-\begin{algorithm}{$\Algo{\color{darkred}{gpu\_fbcorr\_predictive}}\big(shapes, strides \big)$}
+\begin{algorithm}{$\Algo{\color{darkred}{autotune\_predictive}}\big(shapes, strides \big)$}
 \Aitem $a \setto TaskFeatures(shapes, strides)$
@@
 \end{algorithm}
-\caption{Pseudo-code template for an auto-tuning implementation of
-$\texttt{gpu\_fbcorr}$. Empirical auto-tuning (above) is inevitably slow
+\caption{Pseudo-code template for on-demand auto-tuning.
+Empirical auto-tuning (above) is inevitably slow
 because dynamically-generated code must be compiled and run on a number of actual-size inputs.
@@
 if only there were a fast (even approximate) surrogate for the costly $MeasureTime(\cdot)$ function,
-then the argmin could be done in a fraction of a second and the routine could be used normally (Figure~{\ref:fig:general_autotune}, bottom).
+then the argmin could be done in a fraction of a second and the routine could be used normally (Figure~\ref{fig:general_autotune}, bottom).
 
@@
 We call this approach {\em predictive} auto-tuning to contrast it with the standard measurement-based approach which we will call {\em empirical} auto-tuning.
-We show that the kernel from the GCG chapter can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning. XXX numbers:
-how good, how fast can prediction be done how much training data is necessary?
+XXX: We show that the kernel from the GCG chapter can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.
 
-\section{Filterbank Correlation}
+\cite{volkov+demmel:2008}
+\cite{vuduc:2003} % has some stuff about ML for autotuning in ch. 9
+
+\subsection{Filterbank Correlation}
 \label{sec:fbcorr}
@@
 an important subroutine in many image processing applications. It has a relatively
-high arithmetic intensity which makes it a natural fit for GPU platforms.
-One might expect then, that it would be easy to implement a library providing
-this operation -- this hypothetical library could have quite a simple interface
-following the spirit of FFTW or a BLAS routine:
-\begin{verbatim}
-    gpu_fbcorr(
-        img_shape, filters_shape,
-        img, filters, output)
-\end{verbatim}
-However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
-related stencil operations XXX, it is challenging to provide an implementation
-or even an implementation strategy that provides satisfactory performance
-across the range of inputs (shapes, physical layouts) that occur even in
-typical usage.
+high arithmetic intensity which makes it a natural fit for GPU platforms~\citep{pinto+cox:2011gcg}. XXX cite some computer vision papers.
 
-Current state-of-the art is to do empirical auto-tuning.
-XXX \vspace{12pt}
-
-Increasingly, machine learning and statistical inference strategies are
-finding application in compiler technology.
-XXX \vspace{12pt}
-
-machine learning
-
-This paper shows that a kernel with several  boosted
-
-There is a natural function from
-problem space x implementation space x platform space to runtime:
-how much wall time elapses on the given platform when solving the given
-problem with the given implementation.
-
-Auto-tuning is a family of empirical techniques for finding the implementation
-that minimizes that runtime function, when the platform and the problem
-configuration are given.
-
-XXX: some math, or possibly a picture with the three boxes that was on Pinto's
-whiteboard.
-
-This paper is about different ways of doing the minimization.
-For a particular problem (filterbank correlation) we
-look at a parameterized kernel implementation (from GCG)
-and a few hardware platforms, and compare
-random search, grid search, and a hill-climbing strategy with a good, but
-generic, reference implementation.
-Then we show that we can model the LSM function with a regression tree
-well enough to usefully perform the argmin of Equation~\ref{eq:lsm_argmin}
-on the model rather than the actual system. Whereas it may take several
-minutes to perform a hill-climbing or grid search in the original system,
-a hill-climbing search in the model requires a small fraction of a second.
-Model-based autotuning makes it possible to simulate auto-tuning quickly and
-accurately enough to be used within a single library call of the computational
-routine.
-
-\cite{volkov+demmel:2008}
-\cite{vuduc:2003} % has some stuff about ML for autotuning in ch. 9
-
-\subsection{Filter-bank Correlation}
-
-Following \citet{pinto+cox:2011gcg}, the experiments in this paper will focus on filterbank
-correlation.  Mathematically, we define filterbank correlation in terms of an
+Mathematically, we define filterbank correlation in terms of an
 image $x$ and a filterbank $f$.
@@
 
-Filterbank correlation is used extensively in image and video processing,
-where it is often the computational bottleneck.
-GPUs are well-suited to computing filterbank correlations, although the
-optimal strategy for doing so often depends on the shape of the problem, i.e.
-the values of $R$, $C$, $D$, $F$, and $W$.
-
 In terms of floating point operations, a filterbank correlation requires the
@@
 \end{equation}
-The multiplicative factor of 2 arises because we must multiply an element of
+The multiplicative factor of 2 arises because we must first multiply an element of
 $x$ with an element of $f$ and then add the result to an element of $z$.
 
-In terms of bandwidth, if we were to assume (optimistically) that the entire
-filterbank fit into the GPU's so-called constant memory, then the bandwidth
-required for a filterbank correlation would be the size of the filterbank,
-plus the size of the input image, plus the size of the output image. This sum
-is expressed in terms of configuration parameters in Eq.~\ref{eq:bytes}:
+The memory transfer requirements of filterbank correlation are more difficult to quantify.
+Assuming three kinds of non-register memory -- constant, shared, and global --
+and assuming optimistically that the entire filterbank fits into the GPU's constant memory,
+then we can establish a lower bound (Eq.~\ref{eq:bytes}) on the amount of
+memory that must be moved in order to store the computed result to global
+memory starting from arguments in global memory:
 \begin{align}
@@
 & + 4KHWD \nonumber \\
-& + 4(R-H+1)(C-W+1)K
+& + 4(R-H+1)(C-W+1)K.
 \label{eq:bytes}
 \end{align}
-We will see the bandwidth requirements of a more realistic analysis in
-Section~XXX.
+In short, we must read the filterbank and image once, and store the result.
 
-In terms of artithmetic density, it's approximately
+The arithmetic intensity of filterbank correlation, assuming our lower bound on memory transfers
+is therefore approximately
 \begin{align}
@@
 \end{align}
-because each $K$ output writes corresponds to approximately $D$ input reads
-and an inner product between $KDHW$ elements.
+for images that are large relative to filters.
+Each $K$ output writes corresponds to approximately $D$ input reads
+and $K$ inner products between $DHW$ elements.
+
+The high potential for arithmetic intensity makes the GPU an ideal platform for computing filterbank correlations,
+and and filterbank correlation is used extensively in image and video processing,
+where it is often a computational bottleneck.
+One might expect then, that it would be easy to implement a library providing
+this operation -- this hypothetical library could have quite a simple interface
+following the spirit of FFTW or a BLAS routine:
+\begin{verbatim}
+    gpu_fbcorr(
+        img_shape, filters_shape,
+        img, filters, output)
+\end{verbatim}
+However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
+related stencil operations XXX, it is challenging to provide an implementation
+or even an implementation strategy that provides satisfactory performance
+across the range of inputs (shapes, physical layouts) that occur in
+typical usage.
 
@@
 
-\section{Configuration Spaces}
-
-Recall from the introduction (Eq.~\ref{eq:condopt}) that we are addressing a
-conditional optimization problem, in which we condition on a problem
-configuration ($P$) and a system specification ($S$), and search for the best
-implementation $I$ for addressing problem $P$ on system $S$.
-This section describes the configuration spaces we used in our tests,
-which are described later in Section XXX.
-
-\subsection{Problem Description Space}
-
-Mathematically, filterbank correlation is an operation on an $R \times C \times
-D$ image and a $F \times H \times W \times D$ filterbank.
-We randomly sampled problem specifications from the following space:
-\begin{align*}
-R = C & \in \{ 256, 512, 1024, 2048, 4096 \} \\
-H = W & \in \{ 3, 5, 7, 9, 11 \} \\
-D &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \} \\
-F &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \}
-\end{align*}
-The product space
-includes 1600 elements, but we restricted our experiments to problems that
-represented between 1 and 50 gigaflops. Our problem space included 602
-configurations. A library implementation that must support all image and
-filter sizes must handle on the order of $2^{50}$ forseable
-problem configurations (not even including variations due to strided memory
-layouts).
-
-%XXX: DIAGRAM OF IMAGE, FILTERBANK, OUTPUT
-
-% Problems were considered invalid if the filters were larger than the images.
-% but in the new setting with iheight >= 256 this never happened.
-Problems were tested if they represented between 1 and 50 gigaflops.  Problems
-that were smaller do not fully utilize the hardware, and are handled equally
-well by many kernel settings.
-Problems that were larger take so long to evaluate that there is negligible
-inefficiency in implementing them via multiple calls with smaller images and fewer filters.
-The configuration space described in Figure~\ref{fig:probspace} included 602
-configurations with between 1 and 50 gigaflops.
-
-In a full, reusable, library implementation of the filter correlation
-operation, we would also provide options for arbitrary image shapes,
-arbitrary integer depth and filter shapes, and also support a variety of
-physical input layouts via strides.  These additional options would make our
-approach of automatic auto-tuning even more important, because there would be
-a greater variety in the kinds of computations and memory transfers to
-perform.
-
-\subsection{Implementation Description Space}
+\subsection{GPU Implementation}
 
@@
 
-Best arithmetic density
-
-A bandwidth analysis that takes the blocking structure into account would
-reveal that larger $H$ and $W$ require each block to read a larger input
-region, and store it in a larger shared memory region, which reduces the
-potential for high occupancy.
-
-Filterbank correlation represents a difficult code-optimization problem
-because of the number of interacting constraints: a large value of $P$ reduces
-arithmetic density, but a small value of $P$ requires that each thread do more
-work, and makes global memory access latency harder to hide. (XXX: is there a
-better reason?)
-
-The kernel used to perform
-filterbank correlation was parametrized by 10 parameters, some of which
-were binary and others of which were integer-valued. The full configuration
-space included 12000 kernels.
-XXX: how to make sense of these paramters without code listing?
-
-XXX: What to all the parameters mean?
-XXX: Point to github / GCG for full code listing.
-
-XXX: how many configurations are in the configuration space
-
-XXX: what was the reference kernel, and how was it chosen?
-
 Meta-programming was implemented in Python using
@@
 
-XXX: analysis of READ traffic, WRITE traffic, DATA re-use and CACHEing
-strategies. FLOPS per write etc.
+\section{Experiments}
 
-XXX: Why not use FFT: convolution in spectral domain?
+Recall from the introduction (Eq.~\ref{eq:condopt}) that auto-tuning can be seen
+as a conditional optimization problem. We wish to find an implementation ($\mathcal{B}$) that minimizes runtime or some other scalar-valued cost function
+for given arguments ($\mathcal{A}$) on a particular platform ($\mathcal{C}$).
+In order to perform predictive auto-tuning it is necessary to characterize these
+three types of variables with {\em features}.
+In Section~\ref{sec:modeling} we see how these features are used as inputs to
+the regression model in predictive auto-tuning.
 
-XXXX: Talk about memory transfer requirements vs. speed vs. copy...
+\subsection{Argument Description $\mathcal{A}$}
+
+We describe the arguments to a filterbank correlation with the 6-tuple
+$(R, C, D, F, H, W)$.
+We randomly sampled problem specifications from the following space:
+\begin{align*}
+R = C & \in \{ 256, 512, 1024, 2048, 4096 \} \\
+H = W & \in \{ 3, 5, 7, 9, 11 \} \\
+D &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \} \\
+F &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \}
+\end{align*}
+The product space
+includes 1600 elements, but we restricted our experiments to problems that
+represented between 1 and 50 gigaflops. Our problem space included 602
+configurations.
+
+A library implementation that must support all image and
+filter sizes must handle on the order of $2^{50}$ forseable
+problem configurations (not even including variations due to strided memory
+layouts).
+
+%XXX: DIAGRAM OF IMAGE, FILTERBANK, OUTPUT
+
+% Problems were considered invalid if the filters were larger than the images.
+% but in the new setting with iheight >= 256 this never happened.
+Problems were tested if they represented between 1 and 50 gigaflops.  Problems
+that were smaller do not fully utilize the hardware, and are handled equally
+well by many kernel settings.
+Problems that were larger take so long to evaluate that there is negligible
+inefficiency in implementing them via multiple calls with smaller images and fewer filters.
+The configuration space described in Figure~\ref{fig:probspace} included 602
+configurations with between 1 and 50 gigaflops.
+
+In a full, reusable, library implementation of the filter correlation
+operation, we would also provide options for arbitrary image shapes,
+arbitrary integer depth and filter shapes, and also support a variety of
+physical input layouts via strides.  These additional options would make our
+approach of automatic auto-tuning even more important, because there would be
+a greater variety in the kinds of computations and memory transfers to
+perform.
 
@@
 \end{document}
+
+One might expect then, that it would be easy to implement a library providing
+this operation -- this hypothetical library could have quite a simple interface
+following the spirit of FFTW or a BLAS routine:
+\begin{verbatim}
+    gpu_fbcorr(
+        img_shape, filters_shape,
+        img, filters, output)
+\end{verbatim}
+However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
+related stencil operations XXX, it is challenging to provide an implementation
+or even an implementation strategy that provides satisfactory performance
+across the range of inputs (shapes, physical layouts) that occur even in
+typical usage.
+
+Current state-of-the art is to do empirical auto-tuning.
+XXX \vspace{12pt}
+
+Increasingly, machine learning and statistical inference strategies are
+finding application in compiler technology.
+XXX \vspace{12pt}
+
+machine learning
+
+This paper shows that a kernel with several  boosted
+
+There is a natural function from
+problem space x implementation space x platform space to runtime:
+how much wall time elapses on the given platform when solving the given
+problem with the given implementation.
+
+Auto-tuning is a family of empirical techniques for finding the implementation
+that minimizes that runtime function, when the platform and the problem
+configuration are given.
+
+XXX: some math, or possibly a picture with the three boxes that was on Pinto's
+whiteboard.
+
+This paper is about different ways of doing the minimization.
+For a particular problem (filterbank correlation) we
+look at a parameterized kernel implementation (from GCG)
+and a few hardware platforms, and compare
+random search, grid search, and a hill-climbing strategy with a good, but
+generic, reference implementation.
+Then we show that we can model the LSM function with a regression tree
+well enough to usefully perform the argmin of Equation~\ref{eq:lsm_argmin}
+on the model rather than the actual system. Whereas it may take several
+minutes to perform a hill-climbing or grid search in the original system,
+a hill-climbing search in the model requires a small fraction of a second.
+Model-based autotuning makes it possible to simulate auto-tuning quickly and
+accurately enough to be used within a single library call of the computational
+routine.
+
+Best arithmetic density
+
+A bandwidth analysis that takes the blocking structure into account would
+reveal that larger $H$ and $W$ require each block to read a larger input
+region, and store it in a larger shared memory region, which reduces the
+potential for high occupancy.
+
+Filterbank correlation represents a difficult code-optimization problem
+because of the number of interacting constraints: a large value of $P$ reduces
+arithmetic density, but a small value of $P$ requires that each thread do more
+work, and makes global memory access latency harder to hide. (XXX: is there a
+better reason?)
+
+The kernel used to perform
+filterbank correlation was parametrized by 10 parameters, some of which
+were binary and others of which were integer-valued. The full configuration
+space included 12000 kernels.
+XXX: how to make sense of these paramters without code listing?
+
+XXX: What to all the parameters mean?
+XXX: Point to github / GCG for full code listing.
+
+XXX: how many configurations are in the configuration space
+
+XXX: what was the reference kernel, and how was it chosen?
+
+XXX: analysis of READ traffic, WRITE traffic, DATA re-use and CACHEing
+strategies. FLOPS per write etc.
+
+XXX: Why not use FFT: convolution in spectral domain?
+
+XXXX: Talk about memory transfer requirements vs. speed vs. copy...

### 2011-12-11 paper/draft1.tex
paper

@@
 The dominant approach to implementing high-performance GPU codes has been auto-tuning,
-which is fundamentally slow. Compilation with auto-tunin requires running a large number of
+which is fundamentally slow. Auto-tuning requires running a large number of
 sub-compilations and representative computations to actually {\em measure} which
@@
 \item[$\mathcal{B}$:] implementation description (auto-tuning parameters)
-\item[$\mathcal{C}$:] platform description (compute capability, micro-benchmarks)
+\item[$\mathcal{C}$:] platform description (capabilities, micro-benchmarks)
 \end{description}
@@
 \begin{equation}
-\mathrm{FLOPS} = 2  K  H  W  D  (R - H + 1)( C- W + 1)
+\mathrm{FLOPS} = 2  F  H  W  D  (R - H + 1)( C- W + 1)
 \label{eq:flops}
@@
 \mathrm{Bytes} = 4&RCD \nonumber \\
-& + 4KHWD \nonumber \\
-& + 4(R-H+1)(C-W+1)K.
+& + 4FHWD \nonumber \\
+& + 4(R-H+1)(C-W+1)F.
 \label{eq:bytes}
@@
 \begin{align}
-\mathrm{intensity} & \approx \frac {KDHW} {2(D+K)},
+\mathrm{intensity} & \approx \frac {FDHW} {2(D+F)},
 \end{align}
 for images that are large relative to filters.
-Each $K$ output writes corresponds to approximately $D$ input reads
-and $K$ inner products between $DHW$ elements.
+Each $F$ output writes corresponds to approximately $D$ input reads
+and $F$ inner products between $DHW$ elements.
 
@@
 One might expect then, that it would be easy to implement a library providing
-this operation -- this hypothetical library could have quite a simple interface
-following the spirit of FFTW or a BLAS routine:
-\begin{verbatim}
-    gpu_fbcorr(
-        img_shape, filters_shape,
-        img, filters, output)
-\end{verbatim}
+this operation as a simple function that takes pointers and strides for $x$, $f$, and $z$ and performs the computation.
 However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
@@
 Each thread computes $4 \times n\_output\_4s$ channels for some column and row of $z$.
-Each block computes $4 \times n\_output\_4s$ channels for a subrectangle of the output image ($z$).
+Each block of threads computes $4 \times n\_output\_4s$ channels for a subrectangle of the output image ($z$).
 When there are more than $4 \times n\_output\_4s$ channels in $z$, or if the
@@
 Registers are a precious commodity on the GPU, and this kernel accumulates
-elements of $z$ in registers.
+elements of $v$ in registers.
 The ``spill'' parameter controls whether the current thread's output position in
-$z$ is stored in a register (faster access) or in shared memory (frees up
+$gZ$ is stored in a register (faster access) or in shared memory (frees up
 a register).
@@
 
-When the entire filterbank often does not fit into the GPU's
-constant memory and so $$P=\frac{KH}{4 \cdot n\_output\_4s \cdot
-n\_filter\_r}$$ passes are necessary to compute $z$. In such cases,
-the number of bytes moved is actually more like:
+When the entire filterbank does not fit into the GPU's
+constant memory, $P$ passes are necessary to compute all of $z$, where
+$$P=\frac{FH}{4 \cdot n\_output\_4s \cdot n\_filter\_r}.$$
+In such cases, the number of bytes moved to and from global memory is much higher than
+the theoretical lower limit.
 \begin{align}
 \mathrm{Bytes} = 4&RCDP \nonumber \\
-& + 4KHWD \nonumber \\
-& + 8(R-H+1)(C-W+1)KP.
+& + 4FHWD \nonumber \\
+& + 8(R-H+1)(C-W+1)FP.
 \nonumber
@@
 \end{align}
-These passes make the I/O requirements increase quadratically in $K$ and $H$.
+These passes make the I/O requirements increase quadratically in $F$ and $H$.
 At the same time, the total number of floating-point operations (Eq.~\ref{eq:flops})
@@
 three types of variables with {\em features}.
-In Section~\ref{sec:modeling} we see how these features are used as inputs to
-the regression model in predictive auto-tuning.
+%In Section~\ref{sec:modeling} we see how these features are used as inputs to
+%the regression model in predictive auto-tuning.
 
@@
 $(R, C, D, F, H, W)$.
-We randomly sampled problem specifications from the following space:
+We randomly sampled arguments from the following space:
 \begin{align*}
@@
 \end{align*}
-The product space
-includes 1600 elements, but we restricted our experiments to problems that
-represented between 1 and 50 gigaflops. Our problem space included 602
-configurations.
 
-A library implementation that must support all image and
-filter sizes must handle on the order of $2^{50}$ forseable
-problem configurations (not even including variations due to strided memory
-layouts).
+A library implementation of this operation should support all image and
+filter sizes as well as variations due
+to strided memory layouts. In such a setting would be useful to characterize
+the arguments with
+features such as whether the inputs are Fortran-style contiguous, C-style contiguous, or row-padded to various byte alignments.
+These additional options would make our
+approach of automatic auto-tuning even more important, because there would be
+a greater variety in the kinds of computations and memory transfers to
+perform.
+Our experiments consider a somewhat simplified setting in which the arguments are always stored with depth channels being contiguous in memory, followed by columns, then rows, and then filters having the largest stride.
 
@@
 % but in the new setting with iheight >= 256 this never happened.
-Problems were tested if they represented between 1 and 50 gigaflops.  Problems
-that were smaller do not fully utilize the hardware, and are handled equally
-well by many kernel settings.
-Problems that were larger take so long to evaluate that there is negligible
+The product space in our study
+includes 1600 argument combinations, but we restricted our experiments to correlations that
+represented between 1 and 50 gigaflops of arithmetic.
+Smaller problems do not fully utilize GPU hardware and are handled equally well by many kernel settings.
+Larger take so long to evaluate that there is negligible
 inefficiency in implementing them via multiple calls with smaller images and fewer filters.
-The configuration space described in Figure~\ref{fig:probspace} included 602
-configurations with between 1 and 50 gigaflops.
-
-In a full, reusable, library implementation of the filter correlation
-operation, we would also provide options for arbitrary image shapes,
-arbitrary integer depth and filter shapes, and also support a variety of
-physical input layouts via strides.  These additional options would make our
-approach of automatic auto-tuning even more important, because there would be
-a greater variety in the kinds of computations and memory transfers to
-perform.
+With the gigaflops restriction, the argument space 
+included 602 configurations with between 1 and 50 gigaflops.
 

### 2011-12-11 paper/draft1.tex
paper

@@
 %XXX:  Machine Learning for Auto-tuning: A Data-Drive Approach...} ???
-\title{Boosted Regression Trees for Data-Driven Auto-Tuning}
+%\title{Boosted Regression Trees for Data-Driven Auto-Tuning}
+\title{Machine Learning for Predictive Auto-Tuning}
+%XXX:  Machine Learning for Auto-tuning: A Data-Drive Approach...} ???
 
@@
 typical usage.
-
-XXX: TYPICAL OF STENCIL COMPUTATIONS.
-
+\citet{kamil+etal:2009} summarize a related situation related to general stencil computations in their abstract:
 ``Although the auto-tuning strategy has been successfully applied to libraries,
 generalized stencil kernels are not amenable to packaging as libraries.''
-\cite{kamil+etal:2009}
+% XXX: out of context?
 
-\cite{datta:2009}
+XXX \cite{datta:2009}
 
-\subsection{GPU Implementation}
+This paper contributes a model-based approach to auto-tuning in which the model
+is determined almost entirely using machine learning rather than domain
+knowledge.  We show that even though our timing model has no built-in knowledge of GPU
+devices, or our kernel, or filterbank correlation, it is able to provide fast
+and accurate timing estimates.
+We show that a model built by boosting regression trees
+is fast and accurate enough to be used profitably
+in the algorithm schema shown at the bottom of
+Figure~\ref{fig:general_autotune}, which permits doing on-demand auto-tuning
+for novel (not already auto-tuned) argument combinations.
+We call this approach {\em predictive auto-tuning}.
+
+The paper is organized as follows:
+Section~\ref{sec:kernel} describes the sort of kernel we employ for our benchmarking,
+Section~\ref{sec:exp} presents the results of our benchmarking experiments, which
+compare a reference implementation to a) empirical auto-tuning over a domain-specific grid,
+b) empirical auto-tuning over a hill-climbing search, and
+c) predictive autotuning.
+%Section~\ref{sec:discussion}.
+Section~\ref{sec:previous} summarizes previous and related work involving  machine learning with performance auto-tuning.
+Section~\ref{sec:conclusion} summarizes our findings and outlines directions for future work.
+
+\section{GPU Implementation}
+\label{sec:kernel}
 
@@
 filters so in our setting the total number of flops is proportional to $H^4$.
-% XXX WHERE IS ALL THIS GOING?
 
-Meta-programming was implemented in Python using
+XXX
+Arithmetic density with passes.
+
+Critically:
+The arithmetic intensity, shared storage, and register requirements
+of this kernel change significantly and in a complicated platform-dependent way
+with the argument parameters ($R, C, D, F, H, W$) and with the implementation parameters,
+especially ``block\_w'', ``block\_h'', ``n\_output\_4s'' and ``n\_filter\_rows.''
+
+This kernel was implemented in the meta-programming style advocated
+in \citet{pinto+cox:2011gcg} in Python using
 Cheetah for string processing~\citep{cheetah}
-and PyCUDA~\citep{klochner+etal:2009} for dynamic kernel compilation.
+and PyCUDA~\citep{klochner+etal:2009} for dynamic kernel
+compilation and interfacing with CUDA.
 

### 2011-12-11 paper/draft1.tex
paper

@@
 The paper is organized as follows:
+Section~\ref{sec:brtree} describes the boosted regression tree model and the procedure for fitting it to empirical timing data.
 Section~\ref{sec:kernel} describes the sort of kernel we employ for our benchmarking,
@@
 Section~\ref{sec:conclusion} summarizes our findings and outlines directions for future work.
+
+\section{Boosted Regression Trees}
+\label{sec:brtree}
+
+A regression tree is a piece-wise constant function from one vector space to another,
+that works by recursively subdividing the input space into constant regions~\citep{breiman, textbook?}.
+They are widely used in statistics and data-mining applications because the fitting algorithm is quick and reliable, and the form of the tree can provide insight into the relevant input variables.
+We use a standard fitting procedure, which takes a set of
+$(x, y)\in \mathbb{R}^k \times \mathbb{R}$ pairs and constructs a tree with a low mean squared error.
+To construct each node of a regression tree,
+we sort the set $\mathcal{D}$ of ($x,y$) pairs along each of the $k$ features to find the best partitioning $f_{i,\gamma}$ of the input space along feature $i$ at point $\gamma$ (Eqs.~\ref{eq:figamma}, \ref{eq:treeloss}).
+\begin{align}
+    f_{i,\gamma}(x) &=
+    \begin{cases}
+        \alpha  &\text{if $x_i < \gamma$} \\
+        \beta &\text{if $x_i \geq \gamma$}
+    \end{cases}
+    \label{eq:figamma}\\
+    i^*, \gamma^* &= \operatorname*{argmin}_{i, \gamma} \mathbb{\hat E}\left[ (y - f_{i,\gamma}(x))^2 \right]
+    \label{eq:treeloss}
+\end{align}
+One disadvantage of the regression tree is that it does not make full use of broad patterns in the data -- each partition formed by the fitting procedure is fit independently in the recursive training procedure,
+so it impossible for the model to extract more than one bit of information from each training partition (XXX unclear).
+This disadvantage is mitigated to a significant extent by the practice of {\em boosting}~\citet{XXX}.
+
+Boosting is an iterative procedure for constructing an {\em ensemble} of regression trees that is coordinated to fit training examples as accurately as possible.
+In a recent empirical study of a range of machine learning regressino problems,
+boosted decision trees were found to be among the best and easiest models to apply~\cite{caruana:2010}.
+On each boosting iteration, a regression tree is fit to the residual error remaining
+after all previously-fit models have made their predictions.
+There are essentially three parameters that control the boosted regression tree
+training procedure:
+1) the depth of tree constructed on each boosting iteration,
+2) the minimum number of examples to allow at a regression tree leaf, and
+3) the number of trees constructed by boosting.
+We did not attempt a systematic study of the effect of these variables on performance.
+We chose a maximum depth of 4,
+a minimum number of examples of 10,
+and 100 iterations of boosting.
 
@@
 Recall from the introduction (Eq.~\ref{eq:condopt}) that auto-tuning can be seen
-as a conditional optimization problem. We wish to find an implementation ($\mathcal{B}$) that minimizes runtime or some other scalar-valued cost function
-for given arguments ($\mathcal{A}$) on a particular platform ($\mathcal{C}$).
-In order to perform predictive auto-tuning it is necessary to characterize these
+as a conditional optimization problem in whic we seek an implementation ($b \in \mathcal{B}$) that minimizes runtime or some other scalar-valued cost function
+for given arguments ($a \in \mathcal{A}$) on a particular platform ($c \in \mathcal{C}$).
+In order to perform predictive auto-tuning with a regression model,
+it is necessary to characterize these
 three types of variables with {\em features}.
@@
 %the regression model in predictive auto-tuning.
-
-\subsection{Argument Description $\mathcal{A}$}
-
 We describe the arguments to a filterbank correlation with the 6-tuple
 $(R, C, D, F, H, W)$.
-We randomly sampled arguments from the following space:
+We randomly sampled arguments (uniformly) from the following product space:
 \begin{align*}
@@
 
-A library implementation of this operation should support all image and
+A library implementation of this operation would ideally support all image and
 filter sizes as well as variations due
@@
 
-\section{Auto-Tuning by Hill Climbing}
+For the implementation features $b$, we directly used the integer and binary values (block\_w, block\_h, etc.) that parametrized the kernels.
+We did not use platform features ($c$) in our experiments, we leave the investigation
+of cross-platform predictive auto-tuning for future work.
 
-\subsection{Random, Grid, and Hill-Climbing Search}
+\subsection{Search Algorithms}
+
+The core of any online auto-tuning algorithm (Fig.~\ref{fig:autotune_general}) is the search procedure used to explore the space of implementations.
+Our implementation space includes 16000 (XXX) points, too many to search exhaustively.
+\cite{cox+pinto:2011gcg} recommends a particular hyper-parameter implementation
+based on empirical autotuning on several platforms, that achieves good
+performance on a variety of GPUs from older-generation cards such as the
+8600GT all the way to
+current high-end cards such as the GTX 580 and C2070.
+We call this configuration the {\em reference} implementation.
+
+Additionally, \cite{cox+pinto:2011gcg} advocate a particular grid search over
+the most promising part of the configuration space. This grid iterates over all
+combinations of XXX, but leaves the value of XXX at. In our experiments, we call this algorithm the {\em grid} search procedure.
+
+In this work, we consider a third search algorithm based on a local search heuristic.
+This {\em hill-climing} algorithm starts from the reference implementation and randomly resamples $25\%$ of the implementation parameters randomly
 
@@
 
-\begin{figure*}
+\begin{figure}
 \centering
@@
 \label{fig:speedup}
-\end{figure*}
+\end{figure}
 
@@
 \end{figure*}
-
-\section{Boosted Regression Trees}
-
-The regression tree is a piece-wise constant function
-
-Boosted regression trees are
 

### 2011-12-11 paper/draft1.tex
paper

@@
 
-This kernel was implemented in the meta-programming style advocated
-in \citet{pinto+cox:2011gcg} in Python using
-Cheetah for string processing~\citep{cheetah}
-and PyCUDA~\citep{klochner+etal:2009} for dynamic kernel
-compilation and interfacing with CUDA.
-
-\section{Experiments}
+\section{Experiment Setup}
 
@@
 We call this configuration the {\em reference} implementation.
+% XXX: what is reference?  Why was it chosen?
 
@@
 combinations of XXX, but leaves the value of XXX at. In our experiments, we call this algorithm the {\em grid} search procedure.
+The grid included $72$ points in addition to the reference implementation, for a total of $73$ points.
+XXX: isn't 73 prime?? explain this so-called ``grid''.
 
-In this work, we consider a third search algorithm based on a local search heuristic.
-This {\em hill-climing} algorithm starts from the reference implementation and randomly resamples $25\%$ of the implementation parameters randomly
+In this work, we also consider a third search algorithm based on a local search
+heuristic.  This {\em hill-climing} (HC) algorithm starts from the reference
+implementation and resamples each of the parameters
+of the current best implementation randomly with 
+probability $0.25$ (keeping the current best setting with probability $0.75$).
+On each hill-climbing iteration, if the speed of the newly sampled point is greater than the previous point, then it becomes the current point.
+We show results for search variants HC25, HC50, and HC75, which correspond do hill-climbin for 25, 50, and 75 iterations respectively.
 
-XXX: Histogram (over problems) of gigaflops/sec of random, grid, random,
-hill-climbing for each/selected platform.
-This shows that our implementations and problems are stressing the cards,
-that our implementation is pretty good, and that different problem sizes pose
-different computational challenges.
+\subsection{Platforms and Implementation}
 
-XXX: Scatterplot of gigaflops vs. gigaflops / second (of best grid point)
-should show that it's not just problem size that makes it possible to go fast.
+This kernel was implemented in the meta-programming style advocated
+in \citet{pinto+cox:2011gcg} in Python using
+Cheetah for string processing~\citep{cheetah}
+and PyCUDA~\citep{klochner+etal:2009} for dynamic kernel
+compilation and interfacing with CUDA.
 
-XXX: histogram of speedups of grid, random, hill-climbing over reference, for
-several hardware platforms. (GET ANOTHER -- macbook??)
-Sometimes the reference implementation is optimal, in many cases it is far
-from optimal: Which cases? What's wrong with the reference?
+We evaluated our approach on several GPU devices:
+XXX.
 
-TL;DR: Hill-climbing is a good way to search the implementation space, can
-search the full space and do about as well as grid search in the more
-restricted space.
+XXX Table showing theoretical peak FLOPS, theoretical peak bandwidth, core count, shared memory size, register size, etc.
+
+\section{Results}
+
+Figure~\ref{fig:speedup} shows the effectiveness of empirical auto-tuning is in this setting.  Averaging across the range of problem configurations in our study,
+implementations discovered by empirical auto-tuning are on average about 50\% (XXX check) faster than the reference implementation.
+XXX: recalculate with simpler mean speedup formula.
+On the GTX 295 speedups over the reference implementation ranged from XXX to XXX with mean XXX.
+On the GTX 480 speedups over the reference implementation ranged from XXX to XXX with mean XXX.
+XXX: Consider table for these numbers.
+XXX: Put in some timings of mixed \& matched all-stars to show that indeed the reference was a good general baseline, and in fact the auto-tuned models are special-purpose.
 
@@
 
-\begin{figure*}
+Figure~\ref{fig:speedup} also shows that random hill-climbing is at least
+competitive and often better than the hand-chosen grid approach of ~\citet{pinto+cox:2011gcg}.  Figure~\ref{fig:hcX} shows how the number of hill-climbing iterations
+affects the quality of the auto-tuned model.
+The quicker HC25 approach is clearly inferior but the quality of the 75 iteration search (HC75) is statistically similar to the shorter HC50, suggesting that XXX.
+Still, HC75 is not always better than the grid, so evidently neither search algorithm is perfect.
+
+The previous figures establish that auto-tuning on a problem-by-problem basis can achieve good performance improvements over a high-quality reference baseline, but they do not show how long it takes to find these implementations.
+A search of 75 iterations took an average of about 120 seconds XXX on a fast computer, because it requires compiling 75 CUDA kernels, and evaluating up to 750 filterbank correlations to get reliable timing information for each candidate implementation.
+
+Our contrib
+
+TL;DR: Hill-climbing is a good way to search the implementation space, can
+search the full space and do about as well as grid search in the more
+restricted space.
+
+\begin{figure}
 \centering
@@
 \label{fig:fig_ntrain}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}
+\begin{figure}
 \centering
@@
 \label{fig:fig_gflop_scatter}
-\end{figure*}
+\end{figure}
 

### 2011-12-12 paper/draft1.tex
Add some general HPC intro, moving to more motivation text

@@
 
+% HPC Landscape: from single proc to multi-cores
+
+Due to power consumption and heat dissipation concerns, scientific applications
+have shifted from computing platforms where performance had been primarily
+driven by rises in the clock frequency of a single “heavy-weight” processor
+(with complex out-of-order control and cache structures) to a platform with
+ever increasing numbers of “light-weight” cores. Interestingly, this shift is
+now not only relevant to computational sciences but to the development of all
+computer systems: from ubiquitous consumer-facing devices (e.g. phones) to
+high-end computer farms for web-scale applications (e.g. social networks).
+
+% HPC Landscape: large diversity of architectures, no consensus => push to
+% flexibility => push to sw
+
+Even though the future lies in low-power multi-core hardware designs, the field
+lacks meaningful consensus on exactly how the different subsystems (memory,
+communication and computation) should be efficiently integrated, modelled and
+programmed. As a consequence, these systems have been increasingly relying on
+flexible but low-level software-controlled cache management and parallelism
+\cite{BerkeleyViewOfHPC} in order to better control and understand the various
+trade-offs among (e.g.) performance, reliability, energy efficiency, production
+costs, etc. This evolution has profoundly altered the landscape of software
+development: application programmers are now facing a large diversity of
+low-level architectural features while being left with rapidly evolving tools
+and techniques to write high-performance \emph{and} portable code.
+
+% Better tools: motivation
+
+% XXX
+
+\subsection{Case Study}
 Current languages and software engineering practices have conspired to make it difficult

### 2011-12-12 paper/draft1.tex
Fix quotes

@@
 have shifted from computing platforms where performance had been primarily
-driven by rises in the clock frequency of a single “heavy-weight” processor
+driven by rises in the clock frequency of a single ''heavy-weight'' processor
 (with complex out-of-order control and cache structures) to a platform with
-ever increasing numbers of “light-weight” cores. Interestingly, this shift is
+ever increasing numbers of ''light-weight'' cores. Interestingly, this shift is
 now not only relevant to computational sciences but to the development of all

### 2011-12-12 paper/draft1.tex
Add more intro text

--- 2011-12-11 paper/draft1.tex
@@
 patterns and computational bottlenecks.  How can we draw on data from previous
-empirical auto-tuning of related problems on related hardware to make a just-in-time
-implementation decision for a novel problem?  This paper presents a machine
-learning approach to auto-tuning, in which features of (a) the current
-hardware platform, (b) the kernel configuration and (c) the problem instance
-are passed to a regression model (boosted regression trees) which predicts how
-much faster this kernel will be than a reference baseline.  Combinatorial
-optimization strategies that would normally implement auto-tuning by
-evaluating kernel configurations on the real hardware are orders of magnitude
-faster when evaluating the surrogate regression model instead.  We validate our approach
-using the filterbank correlation kernel described in \citet{pinto+cox:2011gcg}, where we find
-that 0.1 seconds (XXX) of hill climbing on the regression model ({\em
-predictive}
-auto-tuning) can achieve an
-average of 95\% of the speed brought by 120 seconds of empirical auto-tuning XXX.  Our
-approach is not specific to filterbank correlation, or even to GPU
-kernel auto-tuning: the approach of using a non-linear regression model on top of
-simple features applies to a variety of problem types, kernel types, and
-platforms.
-\end{abstract}
+empirical auto-tuning of related problems on related hardware to make a
+just-in-time implementation decision for a novel problem?  This paper presents
+a machine learning approach to auto-tuning, in which features of (a) the
+current hardware platform, (b) the kernel configuration and (c) the problem
+instance are passed to a regression model (boosted regression trees) which
+predicts how much faster this kernel will be than a reference baseline.
+Combinatorial optimization strategies that would normally implement auto-tuning
+by evaluating kernel configurations on the real hardware are orders of
+magnitude faster when evaluating the surrogate regression model instead.  We
+validate our approach using the filterbank correlation kernel described in
+\citet{pinto+cox:2011gcg}, where we find that 0.1 seconds (XXX) of hill
+climbing on the regression model ({\em predictive} auto-tuning) can achieve an
+average of 95\% of the speed brought by 120 seconds of empirical auto-tuning
+XXX.  Our approach is not specific to filterbank correlation, or even to GPU
+kernel auto-tuning: the approach of using a non-linear regression model on top
+of simple features applies to a variety of problem types, kernel types, and
+platforms.  \end{abstract}
 
@@
 
+% 1 Introduction
+
+% HPC Landscape: from single proc to multi-cores
+
+Due to power consumption and heat dissipation concerns, scientific applications
+have shifted from computing platforms where performance had been primarily
+driven by rises in the clock frequency of a single ‘’heavy-weight’’ processor
+(with complex out-of-order control and cache structures) to a platform with
+ever increasing numbers of ‘’light-weight’’ cores. Interestingly, this shift is
+now not only relevant to computational sciences but to the development of all
+computer systems: from ubiquitous consumer-facing devices (e.g. phones) to
+high-end computer farms for web-scale applications (e.g. social networks).
+
+% HPC Landscape: large diversity of architectures, no consensus => push to flexibility => push to sw
+
+Although the future lies in low-power multi-core hardware designs, the field
+lacks meaningful consensus on exactly how the different subsystems (memory,
+communication and computation) should be efficiently integrated, modelled and
+programmed. These systems have exhibited varying degrees of memory hierarchy
+and multi-threading complexity and, as a consequence, they have been
+increasingly relying on flexible but low-level software-controlled cache
+management and parallelism \cite{BerkeleyViewOfHPC} in order to better control
+and understand the various trade-offs among performance, reliability, energy
+efficiency, production costs, etc. This evolution has profoundly altered the
+landscape of application development: programmers are now facing a wide
+diversity of low-level architectural features to write high-performance
+\emph{and} portable code.
+
+% 1.1 Motivation
+% Better tools: motivation (compilers/tools/libs can’t keep up, it’s too difficult)
+
+In this rapidly evolving landscape, the construction of \emph{general}
+development tools and libraries that \emph{fully utilize} system resources
+remains a daunting task \cite{?}. Even within specialized architectures from
+the \emph{same} vendor, like NVIDIA’s Graphics Processing Units (GPUs) and the
+Compute Unified Device Architecture (CUDA) \cite{CUDA}, many developers default
+to massive amounts of manual labor to optimize CUDA code to specific input
+domains. In addition, hand-tuning rarely generalizes well to new hardware
+generations or different input domains, and it can also be error-prone or far
+from optimal. One of the reason is that kernels can produce staggeringly large
+optimization spaces \cite{K. Datta et al. Stencil computation optimization and
+auto-tuning on state-of-the-art multicore architectures}. The problem is
+further compounded by the fact that these spaces can be highly discontinuous
+\cite{RyooHwuCGO2008}, difficult to explore and holding quasi-optimal solutions
+at the edge of ‘’performance cliffs’’ (in part due to hard resource
+restrictions, like the register file size).
+
+% 1.2 related work
+% auto-tuning is an answer
+
+One of the strategies for addressing these challenges is to use automatic
+methods called ‘’auto-tuning’’. The literature on auto-tuning is extensive (see
+surveys in \cite{XXX}. Two different approaches to auto-tuning have emerged:
+model-driven optimization and empirical optimization.
+
+The model-driven optimization approach uses analytical abstractions to model
+the hardware architectures, the possible code transformations and their complex
+interactions \cite{XXX}. Even though highly-accurate analytical models are
+generally intractable to build, this approach has been quite successful in the
+past at accelerating serial code \cite{XXX} with simplified but general
+abstractions. However, large speed-ups for parallel code require more accurate
+high-dimensional models and since this approach is bound by the quality and
+scalability of its abstraction, it has been less suited for highly-specialized
+kernels. This approach has been dominant in the compiler community \cite{XXX}.
+As a result, it has generally been applied at compile-time where important
+run-time characteristics such as input domains are missing. These drawbacks
+render the model-driven optimization approach less attractive for
+high-performance library developers.
+
+% XXX: see gdoc and merge this below
+
+\subsection{Case Study}
 Current languages and software engineering practices have conspired to make it difficult

### 2011-12-12 paper/draft1.tex
Fix formatting issues

@@
 \title{Machine Learning for Predictive Auto-Tuning}
+% NP: I like this title !
 %XXX:  Machine Learning for Auto-tuning: A Data-Drive Approach...} ???
@@
 have shifted from computing platforms where performance had been primarily
-driven by rises in the clock frequency of a single ‘’heavy-weight’’ processor
+driven by rises in the clock frequency of a single ''heavy-weight'' processor
 (with complex out-of-order control and cache structures) to a platform with
-ever increasing numbers of ‘’light-weight’’ cores. Interestingly, this shift is
+ever increasing numbers of ''light-weight'' cores. Interestingly, this shift is
 now not only relevant to computational sciences but to the development of all
@@
 % 1.1 Motivation
-% Better tools: motivation (compilers/tools/libs can’t keep up, it’s too difficult)
+% Better tools: motivation (compilers/tools/libs can't keep up, it's too difficult)
 
@@
 remains a daunting task \cite{?}. Even within specialized architectures from
-the \emph{same} vendor, like NVIDIA’s Graphics Processing Units (GPUs) and the
+the \emph{same} vendor, like NVIDIA's Graphics Processing Units (GPUs) and the
 Compute Unified Device Architecture (CUDA) \cite{CUDA}, many developers default
@@
 \cite{RyooHwuCGO2008}, difficult to explore and holding quasi-optimal solutions
-at the edge of ‘’performance cliffs’’ (in part due to hard resource
+at the edge of ''performance cliffs'' (in part due to hard resource
 restrictions, like the register file size).
@@
 One of the strategies for addressing these challenges is to use automatic
-methods called ‘’auto-tuning’’. The literature on auto-tuning is extensive (see
+methods called ''auto-tuning''. The literature on auto-tuning is extensive (see
 surveys in \cite{XXX}. Two different approaches to auto-tuning have emerged:

### 2011-12-12 experiment.py
m

@@ -516,131 +516,6 @@ def main_fig1():
     plt.show()
 
 
-def main_fig_gflop_scatter():
-    _python, _cmd, wisdomfile, inv_mult, n_train = sys.argv
-    import matplotlib.pyplot as plt
-    train_result = 'train_results_%s_%s_%s' % (wisdomfile, inv_mult, n_train)
-    wdb, results, rng = cPickle.load(open(wisdomfile))
-    mdl_timings = cPickle.load(open(train_result))
-
-    def getpspecspeed(pspec, k):
-        for r in results:
-            if k in r and r[k].prob_spec == pspec:
-                if r[k].valid:
-                    return r[k].speed()
-                else:
-                    return None
-    def use_this(p, t):
-        return t.valid and getpspecspeed(p, 'ref') is not None
-    hc_speeds = [getpspecspeed(p, 'gen75') for p, t in mdl_timings.items() if use_this(p, t)]
-    tree_speeds = [t.speed() for p, t in mdl_timings.items()]
-    grid_speeds = [getpspecspeed(p, 'grid') for p, t in mdl_timings.items() if use_this(p, t)]
-    ref_speeds = [getpspecspeed(p, 'ref') for p, t in mdl_timings.items() if use_this(p, t)]
-
-    plt.scatter(hc_speeds, tree_speeds, c='b')
-    #plt.scatter(ref_speeds, tree_speeds, c='r')
-
-    if '295' in wisdomfile:
-        ubound =  350
-        plt.title('GTX 295')
-        figname = 'fig_gflop_scatter_295.pdf'
-    elif '1060' in wisdomfile:
-        ubound =  350
-        plt.title('Tesla C1060')
-        figname = 'fig_gflop_scatter_1060.pdf'
-    elif '480' in wisdomfile:
-        ubound =  850
-        plt.title('GTX 480')
-        figname = 'fig_gflop_scatter_480.pdf'
-    elif '8600' in wisdomfile:
-        ubound =  100
-        plt.title('8600GT')
-        figname = 'fig_gflop_scatter_8600.pdf'
-
-    plt.plot([0, ubound], [0, ubound], c='k', ls='-')
-    plt.plot([0, ubound], [0, ubound / 2.0], c='k', ls='--')
-    plt.plot([0, ubound/2.0], [0, ubound], c='k', ls='--')
-    plt.text(ubound, ubound, 'equality')
-    plt.text(ubound, ubound/2.0, '2x slower')
[…]

### 2011-12-12 fig.py

@@ -1,0 +1,173 @@
+"""
+Scripts for making figures
+
+"""
+
+import cPickle
+import logging
+import sys
+import time
+
+import numpy as np
+
+import wisdom
+import matplotlib.pyplot as plt
+
+logger = logging.getLogger(__name__)
+logging.basicConfig(level=logging.DEBUG)
+
+
+def step_timings(hostname, devicename):
+    """
+    Retrieve the results list produced by experiment.main_step
+    """
+    filename = ('timings/%(hostname)s/timing1_%(devicename)s_big.pkl' %
+            locals())
+    wdb, results, rng = cPickle.load(open(filename))
+    return results
+
+def test_timings(hostname, devicename, inv_mult, n_train):
+    """
+    Retrieve the results of predictive auto-tuning from experiment.main_train
+    """
+    filename = ('timings/%(hostname)s/train_results_timing1_%(devicename)s_big.pkl_%(inv_mult)s_%(n_train)s' % locals())
+    return cPickle.load(open(filename))
+
+
+
+def main_gflop_scatter():
+    _python, _cmd, hostname, devicename, inv_mult, n_train = sys.argv
+    results = step_timings(hostname, devicename)
+    mdl_timings = test_timings(hostname, devicename, inv_mult, n_train)
+
+    def getpspecspeed(pspec, k):
+        for r in results:
+            if k in r and r[k].prob_spec == pspec:
+                if r[k].valid:
+                    return r[k].speed()
+                else:
+                    return None
+    def use_this(p, t):
+        return t.valid and getpspecspeed(p, 'ref') is not None
+    hc_speeds = [getpspecspeed(p, 'gen75') for p, t in mdl_timings.items() if use_this(p, t)]
+    tree_speeds = [t.speed() for p, t in mdl_timings.items()]
+    grid_speeds = [getpspecspeed(p, 'grid') for p, t in mdl_timings.items() if use_this(p, t)]
+    ref_speeds = [getpspecspeed(p, 'ref') for p, t in mdl_timings.items() if use_this(p, t)]
+
+    plt.scatter(hc_speeds, tree_speeds, c='b')
+    #plt.scatter(ref_speeds, tree_speeds, c='r')
+
+    if devicename == '295':
+        ubound =  350
+        plt.title('GTX 295')
+    elif '1060'  == devicename:
+        ubound =  350
[…]

### 2011-12-12 fig.py
fig_ntrain

@@ -99,9 +99,9 @@ def main_gflop_scatter():
 
 
 def main_ntrain():
-    _python, _cmd, wisdomfile = sys.argv
-    import matplotlib.pyplot as plt
-    wdb, results, rng = cPickle.load(open(wisdomfile))
+    _python, _cmd, hostname, devicename = sys.argv
+    results = step_timings(hostname, devicename)
+
     def getpspecspeed(pspec, k):
         for r in results:
             if k in r and r[k].prob_spec == pspec:
@@ -123,9 +123,10 @@ def main_ntrain():
         yy = []
         yy_v = []
         for n_train in '10', '25', '50', '100', '200':
-            train_result = 'train_results_%s_%s_%s' % (wisdomfile, inv_mult, n_train)
+            train_result = test_timings(hostname, devicename, inv_mult,
+                    n_train)
             mdl_timings = [(p, t)
-                    for (p, t) in cPickle.load(open(train_result)).iteritems()
+                    for (p, t) in train_result.iteritems()
                     if use_this(p, t)]
             tree_speeds = [t.speed() for p, t in mdl_timings]
             hc_speeds = [getpspecspeed(p, 'gen75') for p, t in mdl_timings]
@@ -147,22 +148,24 @@ def main_ntrain():
 
     if '295'  == devicename:
         plt.title('GTX 295')
-        figfile = 'fig_ntrain_295.pdf'
     elif '480'  == devicename:
         plt.title('GTX 480')
-        figfile = 'fig_ntrain_480.pdf'
+    elif '580'  == devicename:
+        plt.title('GTX 580')
     elif '1060'  == devicename:
         plt.title('Tesla C1060')
-        figfile = 'fig_ntrain_1060.pdf'
+    elif '2070'  == devicename:
+        plt.title('Tesla C2070')
     elif '8600'  == devicename:
         plt.title('8600GT')
-        figfile = 'fig_ntrain_8600.pdf'
+    figfile = 'fig_ntrain_%s_%s.pdf' % (hostname, devicename,)
     plt.xlabel('N. auto-tuned problem configurations used for training')
     plt.ylabel('GFLOP/s')
     plt.legend(loc='lower right')
     if 0:
         plt.show()
     else:
+        print 'saving figure', figfile
         plt.savefig(figfile)
 
 

### 2011-12-13 experiment.py
adding munctional figures

@@ -366,6 +366,29 @@ def main_train():
     cPickle.dump(mdl_timings, open('train_results_%s_%s_%s' % (wisdomfile, inv_mult, n_train), 'w'))
 
 
+def main_allstars_mixup():
+    _python, _cmd, wisdomfile, dev_id_str = sys.argv
+    wdb, results, rng = cPickle.load(open(wisdomfile))
+    pspec_generator = problem_generator(rng)
+    timings = []
+    device = init_cuda(int(dev_id_str))
+    for ii, finding_dct in enumerate(results):
+        if 'gen75' not in finding_dct:
+            continue
+        gen75 = finding_dct['gen75']
+        print gen75.speed()
+        rpspec = pspec_generator.next()
+        timing = wisdom.Timing(rpspec, gen75.op_spec)
+        timing.measure(device)
+        if timing.valid:
+            print timing.speed()
+        else:
+            print 'INVALID'
+        timings.append(timing)
+        if len(timings) == 100: break
+    cPickle.dump(timings, open('%s_allstars_mixup_timings' % (wisdomfile,), 'w'))
+
+
 def main_figtrain():
     _python, _cmd, wisdomfile, inv_mult = sys.argv
     train_result = 'train_results_%s_%s' % (wisdomfile, inv_mult)

### 2011-12-13 fig.py

@@ -26,12 +26,47 @@ def step_timings(hostname, devicename):
     wdb, results, rng = cPickle.load(open(filename))
     return results
 
+
 def test_timings(hostname, devicename, inv_mult, n_train):
     """
     Retrieve the results of predictive auto-tuning from experiment.main_train
     """
     filename = ('timings/%(hostname)s/train_results_timing1_%(devicename)s_big.pkl_%(inv_mult)s_%(n_train)s' % locals())
     return cPickle.load(open(filename))
+
+
+def main_allstars():
+    """
+    Produce a scatterplot of ref vs. mismatched auto-tuned kernels
+    """
+    _python, _cmd, wisdomfile, timingfile = sys.argv
+    assert '295' in wisdomfile # otw fix save mechanism
+    _wdb, results, _rng = cPickle.load(open(wisdomfile))
+    timings = cPickle.load(open(timingfile))
+    print "LEN TIMINGS", len(timings)
+    print "N VALID", len([t for t in timings if t.valid])
+    i = 0
+    x = []
+    y = []
+    for finding_dct in results:
+        if 'gen75' in finding_dct:
+            gen75 = finding_dct['gen75']
+            if i >= len(timings):
+                break
+            timing = timings[i]
+            if timing.valid:
+                x.append(gen75.speed())
+                y.append(timing.speed())
+            i += 1
+    print x
+    print y
+    plt.scatter(x, y)
+    plt.xlabel("GFLOPS/s of reference")
+    plt.xlabel("GFLOPS/s of random auto-tuned")
+    if 0:
+        plt.show()
+    else:
+        plt.savefig('allstars_mixup_295.pdf')
 
 
 
@@ -91,7 +126,7 @@ def main_gflop_scatter():
     plt.xlabel('GFLOP/s of empirical auto-tuning')
     plt.ylabel('GFLOP/s of predictive auto-tuning ')
     #plt.legend(loc='lower left')
-    if 0:
+    if 1:
         plt.show()
     else:
         print 'saving to ', figname

### 2011-12-13 wisdom.py

@@ -565,7 +565,6 @@ class Timing(object):
 
     def measure_1(self, fop, in_, out_, fb_, in_data, fb_data, dotransfer=True):
         # -- upload data
-        print "FOOO"
         sys.stdout.flush()
         try:
             start = time.time()
@@ -619,34 +618,26 @@ class Timing(object):
         return in_data, fb_data, out_data
 
     def measure_setup(self, context):
-        print>> sys.stderr,  "FOOO AAA "
         img_shp = self.prob_spec.image_shape()
         ker_shp = self.prob_spec.filters_shape()
         out_shp = self.prob_spec.output_shape()
 
-        print>> sys.stderr,  "FOOO INPUT STUFF"
         in_ = fbconv3_cuda.Input(*img_shp)
         fb_ = fbconv3_cuda.Filterbank(*ker_shp)
         out_ = fbconv3_cuda.Output(*out_shp)
 
         # -- set-up operation (i.e. compilation)
         fb_[:] = 0
-        print>> sys.stderr,  "FOOO FilterOp "
         fop = self.op_spec.FilterOp(in_, fb_, out_, ctxt=context)
-        print>> sys.stderr,  "FOOO FilterOp  RETURNED"
         return fop, in_, fb_, out_
 
     def measure(self, device, N=8):
         """
         Add N timing measurements to self.timings
         """
-        print>> sys.stderr,  "pre sample "
         in_data, fb_data, out_dat = self.get_sample_data()
-        print>> sys.stderr,  "post sample"
-        import pdb; pdb.set_trace()
 
         with CudaContext(device) as context:
-            print>> sys.stderr,  "GOT contextlD"
             try:
                 fop, in_, fb_, out_ = self.measure_setup(context)
             except (fbconv3_cuda.InvalidConfig,

### 2011-12-13 fig.py
fig gen bug

@@ -62,7 +62,7 @@ def main_allstars():
     print y
     plt.scatter(x, y)
     plt.xlabel("GFLOPS/s of reference")
-    plt.xlabel("GFLOPS/s of random auto-tuned")
+    plt.ylabel("GFLOPS/s of random auto-tuned")
     if 0:
         plt.show()
     else:

### 2011-12-13 fig.py
fig genX

@@ -16,6 +16,13 @@ import matplotlib.pyplot as plt
 logger = logging.getLogger(__name__)
 logging.basicConfig(level=logging.DEBUG)
 
+def pretty_dname(devicename):
+    if devicename == '295': return 'GTX 295'
+    if devicename == '480': return 'GTX 480'
+    if devicename == '580': return 'GTX 580'
+    if devicename == '1060': return 'Tesla C1060'
+    if devicename == '2070': return 'Tesla C2070'
+
 
 def step_timings(hostname, devicename):
     """
@@ -35,7 +42,7 @@ def test_timings(hostname, devicename, inv_mult, n_train):
     return cPickle.load(open(filename))
 
 
-def main_allstars():
+def main_allstars_mixup():
     """
     Produce a scatterplot of ref vs. mismatched auto-tuned kernels
     """
@@ -68,6 +75,42 @@ def main_allstars():
     else:
         plt.savefig('allstars_mixup_295.pdf')
 
+
+def main_genX():
+    _python, _cmd, hostname, devicename = sys.argv
+    results = step_timings(hostname, devicename)
+
+    def getel(f):
+        if not ('ref' in f and f['ref'].valid): return None
+        if not ('gen25' in f and f['gen25'].valid): return None
+        if not ('gen50' in f and f['gen50'].valid): return None
+        if not ('gen75' in f and f['gen75'].valid): return None
+        if not ('grid' in f and f['grid'].valid): return None
+        return [f[k].speed() for k in 'ref', 'gen25', 'gen50', 'gen75',
+                'grid']
+    results = [f for f in results if getel(f) is not None]
+    ref, gen25, gen50, gen75, grid = zip(*map(getel, results))
+
+    ref = np.asarray(ref)
+    gen25 = np.asarray(gen25)
+    gen25 = np.asarray(gen50)
+    gen25 = np.asarray(gen75)
+    grid = np.asarray(grid)
+
+    lines = plt.boxplot(
+            x = [gen25/ref, gen50/ref, gen75/ref, grid/ref],
+            widths=[0.7] * 4,
+            #positions=[0, 1, 2, 3, 4],
+            #whis=10,
+            )
+    plt.xticks([1, 2, 3, 4], ['HC25', 'HC50', 'HC75', 'grid'])
+    plt.ylabel('Speedup over reference (%s)' % pretty_dname(devicename))
+    if 0:
+        plt.show()
+    else:
[…]

### 2011-12-13 fig.py
fixing genX figures

@@ -97,14 +97,24 @@ def main_genX():
     gen25 = np.asarray(gen75)
     grid = np.asarray(grid)
 
-    lines = plt.boxplot(
+    axes= plt.axes()
+    lines = axes.boxplot(
             x = [gen25/ref, gen50/ref, gen75/ref, grid/ref],
             widths=[0.7] * 4,
             #positions=[0, 1, 2, 3, 4],
-            #whis=10,
+            whis=100,
             )
-    plt.xticks([1, 2, 3, 4], ['HC25', 'HC50', 'HC75', 'grid'])
     plt.ylabel('Speedup over reference (%s)' % pretty_dname(devicename))
+    plt.xticks([1, 2, 3, 4], ['HC25', 'HC50', 'HC75', 'grid'])
+    if devicename in ('480', '580', '2070'):
+        axes.set_yscale('log')
+        axes.set_ylim(0.5, 33)
+        plt.yticks([1, 2, 4, 8, 16, 32],
+                [('%ix'%i) for i in [1, 2, 4, 8, 16, 32]])
+    else:
+        yticks = [0.5, 1, 1.5, 2, 2.5]
+        plt.yticks(yticks, [('%.1fx'%i) for i in yticks])
+
     if 0:
         plt.show()
     else:

### 2011-12-13 paper/draft1.tex
Add AT section
draft1 includes many figures

@@
 now not only relevant to computational sciences but to the development of all
-computer systems: from ubiquitous consumer-facing devices (e.g. phones) to
+computer systems: from ubiquitous consumer-facing devices (e.g. smart phones) to
 high-end computer farms for web-scale applications (e.g. social networks).
@@
 
+% COMMENT: NP, great setup! -JB
+
+\section{Auto-Tuning}
+
 % 1.2 related work
@@
 scalability of its abstraction, it has been less suited for highly-specialized
+% XXX: the bit above is confusing
 kernels. This approach has been dominant in the compiler community \cite{XXX}.
@@
 high-performance library developers.
+
+% XXX: * empirical optimization: intro / advantages
+% XXX: * empirical optimization: disadvantages
+% XXX: * hybrid optimization (to counter disadvantages) ?
+% XXX: * empirical optimization: w/ ML @ compile time (another way to “fix” empirical)
+% XXX: * empirical optimization: ML @ runtime, especially for libs
+% XXX: * empirical optimization: ML @ runtime for GPU libs why ? (w/ rtcg jit, rel. to pycuda, etc)
+
+% XXX: our contributions
+% XXX: WIIFY
+% XXX: organisation of the paper
 
@@
 perform.
-Our experiments consider a somewhat simplified setting in which the arguments are always stored with depth channels being contiguous in memory, followed by columns, then rows, and then filters having the largest stride.
+Our experiments consider a somewhat simplified setting in which the arguments are always stored
+with depth channels being contiguous in memory, followed by columns, then rows, and then filters having the largest stride.
 
@@
 
-% Problems were considered invalid if the filters were larger than the images.
-% but in the new setting with iheight >= 256 this never happened.
 The product space in our study
@@
 inefficiency in implementing them via multiple calls with smaller images and fewer filters.
-With the gigaflops restriction, the argument space
-included 602 configurations with between 1 and 50 gigaflops.
+With the experiments searched an argument space included 602 configurations with between 1 and 50 gigaflops.
 
 For the implementation features $b$, we directly used the integer and binary values (block\_w, block\_h, etc.) that parametrized the kernels.
-We did not use platform features ($c$) in our experiments, we leave the investigation
+We did not use platform features ($c$) in our experiments. We leave the investigation
 of cross-platform predictive auto-tuning for future work.
@@
 The core of any online auto-tuning algorithm (Fig.~\ref{fig:autotune_general}) is the search procedure used to explore the space of implementations.
-Our implementation space includes 16000 (XXX) points, too many to search exhaustively.
+Our implementation space includes 16000 (XXX) points, too many to search exhaustively for every input configuration.
 \cite{cox+pinto:2011gcg} recommends a particular hyper-parameter implementation
@@
 8600GT all the way to
-current high-end cards such as the GTX 580 and C2070.
+current-generation flagship cards such as the GTX 580 and C2070.
 We call this configuration the {\em reference} implementation.
@@
 Additionally, \cite{cox+pinto:2011gcg} advocate a particular grid search over
-the most promising part of the configuration space. This grid iterates over all
+what was estimated to be the most relevant part of the configuration space. This grid iterates over all
 combinations of XXX, but leaves the value of XXX at. In our experiments, we call this algorithm the {\em grid} search procedure.
@@
 XXX: Put in some timings of mixed \& matched all-stars to show that indeed the reference was a good general baseline, and in fact the auto-tuned models are special-purpose.
+\begin{figure}
+\centering
+\includegraphics[scale=.42]{allstars_mixup_295.pdf}
+\caption{Different arguments call for different kernels: each point corresponds to a randomly drawn argument configuration.
+The reference speed (horizontal) is completely uncorrelated to the speed of an optimal kernel for an unrelated argument configuration (trials performed on GTX 295).
+Out of 100 random confurations, 38 (not shown) could not even be run using a kernel that was auto-tuned for different arguments.
+Good performance across a variety of inputs requires {\em input-dependent auto-tuning}.
+}
+\label{fig:speedup}
+\end{figure}
 
@@
 \includegraphics[scale=.42]{speedup_480.pdf}
-\caption{Speedup of various auto-tuning strategies over a good reference
+\caption{Speedup of various auto-tuning strategies over the reference
 implementation.
@@
 \centering
-\includegraphics[scale=.42]{fig_ntrain_295.pdf}
-\includegraphics[scale=.42]{fig_ntrain_480.pdf}
+\includegraphics[scale=.42]{fig_ntrain_295.pdf} % on vader
+\includegraphics[scale=.42]{fig_ntrain_480.pdf} % on vader
 \caption{The effect of the invalid configuration score ($a$) and training set
@@
 \centering
+\includegraphics[scale=.42]{fig_ntrain_munctional0_1060.pdf}
+\includegraphics[scale=.42]{fig_ntrain_munctional0_2070.pdf}
+\caption{more timing on munctional0}
+\label{fig:fig_ntrain}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.42]{fig_ntrain_munctional0_480.pdf}
+\includegraphics[scale=.42]{fig_ntrain_munctional0_580.pdf}
+\caption{more timing on munctional0}
+\label{fig:fig_ntrain}
+\end{figure}
+
+\begin{figure}
+\centering
 \includegraphics[scale=.42]{fig_gflop_scatter_295.pdf}
@@
 \label{fig:fig_gflop_scatter}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.42]{fig_genX_vader_295.pdf}
+\includegraphics[scale=.42]{fig_genX_munctional0_480.pdf}
+\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
+\label{fig:speedup}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.42]{fig_genX_munctional0_580.pdf}
+\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
+\label{fig:speedup}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.42]{fig_genX_munctional0_1060.pdf}
+\includegraphics[scale=.42]{fig_genX_munctional0_2070.pdf}
+\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
+\label{fig:speedup}
 \end{figure}

### 2011-12-13 paper/draft1.tex

@@
 have shifted from computing platforms where performance had been primarily
-driven by rises in the clock frequency of a single ''heavy-weight'' processor
+driven by rises in the clock frequency of a single ``heavy-weight'' processor
 (with complex out-of-order control and cache structures) to a platform with
-ever increasing numbers of ''light-weight'' cores. Interestingly, this shift is
+ever increasing numbers of ``light-weight'' cores. Interestingly, this shift is
 now not only relevant to computational sciences but to the development of all
-computer systems: from ubiquitous consumer-facing devices (e.g. smart phones) to
+computer systems: from ubiquitous consumer-facing devices (e.g. phones) to
 high-end computer farms for web-scale applications (e.g. social networks).
 
-% HPC Landscape: large diversity of architectures, no consensus => push to flexibility => push to sw
+% HPC Landscape: large diversity of architectures,
+% no consensus => push to flexibility => push to sw
 
@@
 lacks meaningful consensus on exactly how the different subsystems (memory,
-communication and computation) should be efficiently integrated, modelled and
+communication and computation) should be efficiently integrated, modeled and
 programmed. These systems have exhibited varying degrees of memory hierarchy
@@
 increasingly relying on flexible but low-level software-controlled cache
-management and parallelism \cite{BerkeleyViewOfHPC} in order to better control
-and understand the various trade-offs among performance, reliability, energy
-efficiency, production costs, etc. This evolution has profoundly altered the
-landscape of application development: programmers are now facing a wide
-diversity of low-level architectural features to write high-performance
-\emph{and} portable code.
+management and parallelism \citep{asanovic2006landscape} in order to better
+control and understand the various trade-offs among performance, reliability,
+energy efficiency, production costs, etc. This evolution has profoundly
+altered the landscape of application development: programmers are now
+facing a wide diversity of low-level architectural features to write
+high-performance \emph{and} portable code.
 
-% 1.1 Motivation
-% Better tools: motivation (compilers/tools/libs can't keep up, it's too difficult)
+\subsection{Motivation}
+
+% Better tools: motivation (compilers/tools/libs can’t keep up, it’s too
+% difficult)
 
@@
 development tools and libraries that \emph{fully utilize} system resources
-remains a daunting task \cite{?}. Even within specialized architectures from
-the \emph{same} vendor, like NVIDIA's Graphics Processing Units (GPUs) and the
-Compute Unified Device Architecture (CUDA) \cite{CUDA}, many developers default
-to massive amounts of manual labor to optimize CUDA code to specific input
-domains. In addition, hand-tuning rarely generalizes well to new hardware
-generations or different input domains, and it can also be error-prone or far
-from optimal. One of the reason is that kernels can produce staggeringly large
-optimization spaces \cite{K. Datta et al. Stencil computation optimization and
-auto-tuning on state-of-the-art multicore architectures}. The problem is
-further compounded by the fact that these spaces can be highly discontinuous
-\cite{RyooHwuCGO2008}, difficult to explore and holding quasi-optimal solutions
-at the edge of ''performance cliffs'' (in part due to hard resource
-restrictions, like the register file size).
+remains a daunting task. Even within specialized architectures from the
+\emph{same} vendor, like NVIDIA's Graphics Processing Units (GPUs) and the
+Compute Unified Device Architecture (CUDA) \citep{nickolls2008scalable,
+nvidia2011cuda}, many developers default to massive amounts of manual labor
+to optimize CUDA code to specific input domains. In addition, hand-tuning
+rarely generalizes well to new hardware generations or different input
+domains, and it can also be error-prone or far from optimal. One of the
+reason is that kernels can produce staggeringly large optimization spaces
+\citep{datta2008stencil}. The problem is further compounded by the fact that
+these spaces can be highly discontinuous \citep{ryoo2008program}, difficult
+to explore and holding quasi-optimal solutions at the edge of ``performance
+cliffs'' (in part due to hard resource restrictions, like the register file
+size).
 
-% COMMENT: NP, great setup! -JB
+\subsection{Auto-Tuning}
+% Related Work instead ??
 
-\section{Auto-Tuning}
-
-% 1.2 related work
 % auto-tuning is an answer
@@
 One of the strategies for addressing these challenges is to use automatic
-methods called ''auto-tuning''. The literature on auto-tuning is extensive (see
-surveys in \cite{XXX}. Two different approaches to auto-tuning have emerged:
-model-driven optimization and empirical optimization.
+methods called ``auto-tuning''. Two major auto-tuning approaches have emerged
+in the extensive literature covering the subject (see surveys in
+\citep{vuduc2001statistical, demmel2005self, vuduc2005oski, williams2008auto,
+datta2008stencil, cavazos2008intelligent, li2009note, park2011evaluation}):
+analytical model-driven optimization and empirical optimization
+\citep{yotov2003comparison}.
+
+% model-driven optimization: intro
 
@@
 the hardware architectures, the possible code transformations and their complex
-interactions \cite{XXX}. Even though highly-accurate analytical models are
-generally intractable to build, this approach has been quite successful in the
-past at accelerating serial code \cite{XXX} with simplified but general
-abstractions. However, large speed-ups for parallel code require more accurate
-high-dimensional models and since this approach is bound by the quality and
-scalability of its abstraction, it has been less suited for highly-specialized
-% XXX: the bit above is confusing
-kernels. This approach has been dominant in the compiler community \cite{XXX}.
-As a result, it has generally been applied at compile-time where important
-run-time characteristics such as input domains are missing. These drawbacks
-render the model-driven optimization approach less attractive for
-high-performance library developers.
+interactions. Even though highly-accurate analytical models are generally
+intractable to build, this approach has been quite successful in the past at
+accelerating serial code with simplified but general abstractions. However,
+large speed-ups for parallel code require more accurate high-dimensional models
+and since this approach is bound by the quality and scalability of its
+abstraction, it has been less suited for highly-specialized kernels. This
+approach has been dominant in the compiler community. As a result, it has
+generally been applied at compile-time where important run-time characteristics
+such as input domains are missing. These drawbacks render the model-driven
+optimization approach less attractive for high-performance library developers.
 
-% XXX: * empirical optimization: intro / advantages
-% XXX: * empirical optimization: disadvantages
-% XXX: * hybrid optimization (to counter disadvantages) ?
-% XXX: * empirical optimization: w/ ML @ compile time (another way to “fix” empirical)
-% XXX: * empirical optimization: ML @ runtime, especially for libs
-% XXX: * empirical optimization: ML @ runtime for GPU libs why ? (w/ rtcg jit, rel. to pycuda, etc)
+% * empirical optimization: intro / advantages
+
+The empirical optimization approach, on the contrary, seeks to find the best
+performing code configuration by automatically generating many versions of a
+parametrized kernel and benchmarking them on the actual hardware (possibly at
+runtime, when contextual information about the hardware and software stack is
+the richest). This method directly optimizes the metric(s) of interest and not
+surrogates. One of its main advantage is that it allows any metric to be
+optimized without loss of generality. It is indeed possible to formulate the
+problem as a multi-objective optimization and minimize both speed \emph{and}
+power consumption \citep{rahman2011automated}, a feat that renders the
+analytical model-driven approach even more intractable. Due to its flexibility,
+empirical auto-tuning has been successfully applied to build high-performance
+domain-specific libraries including dense linear algebra
+\citep{clint2001automated, bilmes1997optimizing}, sparse linear algebra
+\citep{vuduc2005oski}, signal processing \citep{frigo1998fftw}, sorting
+\citep{li2004dynamically}, general stencil operations \citep{kamil2010auto}, etc.
+
+Even though this ``black-box'' approach does not necessarily try to build an
+explicit model per se, it could, if care is taken, help to greatly accelerate
+our understanding of the complex interactions between the set of optimizations,
+their parameters and the hardware architectural features (XXX: that's what
+we are trying to do w/ ML-based, decision trees, etc.).
+
+% empirical optimization: disadvantages
+
+The empirical approach is very sensible to the choice of instrumented
+optimizations \emph{and} to the search method. The size of the search space is
+so large that the current best empirical auto-tuners can only consider
+highly-specialized functions with a limited set of code transformations and
+compiler options, on very few input domains \citep{ganapathi2009case}. Although
+searching for good code configurations in this highly-discontinuous space can
+be made embarrassingly parallel, it remains a very difficult and prohibitively
+expensive combinatorial optimization problem as many variants of the code must
+generated, compiled, and benchmarked on specific input domains with meaningful
+statistics (that may require multiple runs). Consequently, most proposed
+methods prune the space with hard-coded heuristics that offer little
+generalization guarantees. This has been the main drawback of the empirical
+approach compared to the model-driven approach where good code configurations
+can be directly derived from the analytical model.
+
+% hybrid optimization: the bad way
+
+To alleviate this major weakness, it seems intuitive to combine the two
+approaches by first constraining the search space with an analytical model and
+then exploring the reduced space empirically \citep{chen2005combining,
+li2009note}. Unfortunately, this hybrid approach is still bound by the quality
+of the analytical model, which remains hard to build by hand.
+
+% hybrid optimization: the ML way (note that ML-based are in essence
+% model-driven too just with an empirically learned model)
+
+In this paper, we propose to \emph{learn the model}, empirically, (XXX:
+we'll describe what this means in more details later in the paper, riiiighhht
+???) instead of deriving it analytically. By learning the model, one can
+hope to get ``the best of both approaches'' and capitalize on the large body of
+literature from the machine learning community.
+
+Although various statistical prediction techniques have been applied with
+success at compile-time for general programs on various CPU architectures
+\citep{monsifrot2002machine, stephenson2003meta, yotov2003comparison,
+kulkarni2004fast, cooper2005acme, franke2005probabilistic,
+hutter2006performance, cavazos2007rapidly, cavazos2008intelligent,
+hartono2009annotation, park2011evaluation, fursin2008milepost}, there is, to
+our knowledge, no fast predictive auto-tuning work that has tried to cover
+simultaneously: (a) more recent multi-core architectures like GPUs
+\citep{schaa2009exploring}, (b) high-performance domain-specific libraries
+\citep{nukada2009auto, li2009note, kamil2010auto}, (c) at run-time
+\citep{klockner2011pycuda, pinto+cox:2011gcg}, and (d) on \emph{many} input
+domains \citep{liu2009cross, grauer2011optimizing}.
+
+XXX WIIFY here ??? (what's in their for you) XXX in text, say what's in this paper for:
+- compiler people
+- libs people
+- ml people
+- tools people
+- cv people
+
+% auto-tuning is an answer
 

### 2011-12-13 paper/draft1.tex
Edits and comments
minor intro edit

--- 2011-12-13 paper/draft1.tex
@@
 generations or different input domains, and it can also be error-prone or far
-from optimal. One of the reason is that kernels can produce staggeringly large
+from optimal. Kernels can produce staggeringly large
 optimization spaces \cite{K. Datta et al. Stencil computation optimization and
-auto-tuning on state-of-the-art multicore architectures}. The problem is
-further compounded by the fact that these spaces can be highly discontinuous
-\cite{RyooHwuCGO2008}, difficult to explore and holding quasi-optimal solutions
-at the edge of ''performance cliffs'' (in part due to hard resource
-restrictions, like the register file size).
+auto-tuning on state-of-the-art multicore architectures}, in which performance can vary
+widely and discontinously~\citep{RyooHwuCGO2008}.
+The best kernels are often at the edge of ``performance cliffs'' induced by
+device-specific hardware limits (e.g. shared memory size, register file size, block size).
 

### 2011-12-13 paper/draft1.tex

--- 2011-12-13 paper/draft1.tex
@@
 these spaces can be highly discontinuous \citep{ryoo2008program}, difficult
-to explore and holding quasi-optimal solutions at the edge of ``performance
-cliffs'' (in part due to hard resource restrictions, like the register file
-size).
+to explore and quasi-optimal solutions lie at the edge of ``performance
+cliffs'' induced by device-specific hard constraints (e.g. register file
+size and shared memory size).
 

### 2011-12-13 paper/draft1.tex
Various edits

--- 2011-12-13 paper/draft1.tex
@@
 %\title{Boosted Regression Trees for Data-Driven Auto-Tuning}
-\title{Machine Learning for Predictive Auto-Tuning}
+\title{Boosted Regression Trees for Predictive Auto-Tuning}
 % NP: I like this title !
@@
 \begin{abstract}
-Auto-tuning is a widely used and effective technique for optimizing a
+Template-based auto-tuning techniques are widely used and effective for optimizing a
 parametrized GPU code template for a particular computation on particular
-hardware.  Its drawback is that thorough or exhaustive auto-tuning requires
-compiling many kernels and calling each one many times; this process is slow.
+hardware \cite{XXXX}.  A drawback of this approach is that exhaustive, or even thorough, auto-tuning requires
+compiling many kernels and calling each one many times, and this process is slow.
 Furthermore, library abstraction boundaries provide operations such as image
@@
 just-in-time implementation decision for a novel problem?  This paper presents
-a machine learning approach to auto-tuning, in which features of (a) the
-current hardware platform, (b) the kernel configuration and (c) the problem
+a machine learning approach to auto-tuning, in which features of the
+current hardware platform, the kernel configuration and the problem
 instance are passed to a regression model (boosted regression trees) which
@@
 average of 95\% of the speed brought by 120 seconds of empirical auto-tuning
-XXX.  Our approach is not specific to filterbank correlation, or even to GPU
-kernel auto-tuning: the approach of using a non-linear regression model on top
+XXX.  Our approach is not specific to filterbank correlation, nor to GPU
+kernel auto-tuning; the approach of using a non-linear regression model on top
 of simple features applies to a variety of problem types, kernel types, and
@@
 Although the future lies in low-power multi-core hardware designs, the field
-lacks meaningful consensus on exactly how the different subsystems (memory,
+lacks  consensus on exactly how the different subsystems (memory,
 communication and computation) should be efficiently integrated, modeled and
@@
 
-One of the strategies for addressing these challenges is to use automatic
-methods called ``auto-tuning''. Two major auto-tuning approaches have emerged
+One strategy for addressing these challenges is to use one of a variety of automatic
+methods known collectively as ``auto-tuning''. Two major auto-tuning approaches have emerged
 in the extensive literature covering the subject (see surveys in
@@
 power consumption \citep{rahman2011automated}, a feat that renders the
-analytical model-driven approach even more intractable. Due to its flexibility,
+analytical model-driven approach even more difficult. Due to its flexibility,
 empirical auto-tuning has been successfully applied to build high-performance
@@
 
-Even though this ``black-box'' approach does not necessarily try to build an
-explicit model per se, it could, if care is taken, help to greatly accelerate
+Even though such a ``black-box'' approach does not try to build an
+explicit model per se, it could, if care were taken, help to greatly accelerate
 our understanding of the complex interactions between the set of optimizations,
@@
 
+% DDC: "sensible to"?  what is meant here? "Sensitive to?"
 The empirical approach is very sensible to the choice of instrumented
@@
 
-To alleviate this major weakness, it seems intuitive to combine the two
+To alleviate this major weakness, it is intuitively appealing to combine the two
 approaches by first constraining the search space with an analytical model and
@@
 ???) instead of deriving it analytically. By learning the model, one can
-hope to get ``the best of both approaches'' and capitalize on the large body of
+hope to achieve elements of the best of both approaches and capitalize on the large body of
 literature from the machine learning community.
@@
 to provide high-performance libraries for general mathematical operations on GPU devices.
-The dominant approach to implementing high-performance GPU codes has been auto-tuning,
+An important approach in implementing high-performance GPU codes has been auto-tuning,
 which is fundamentally slow. Auto-tuning requires running a large number of
-sub-compilations and representative computations to actually {\em measure} which
+sub-compilations and representative computations to {\em measure} which
 of several possible implementations is fastest. General-purpose libraries such
@@
 reusable infrastructure to assist in the implementation of different computations.
-Libraries for stencil computations and sparse matrix operations must start from scratch.
+For instance, even if one has already developed an auto-tuned library for dense matrix algebra, 
+one must basically start from scratch to develop an auto-tuned library for
+stencil computations or sparse matrix operations.
 

### 2011-12-13 paper/draft1.tex
draft1

--- 2011-12-13 paper/draft1.tex
@@
 \includegraphics[scale=.42]{allstars_mixup_295.pdf}
-\caption{Different arguments call for different kernels: each point corresponds to a randomly drawn argument configuration.
-The reference speed (horizontal) is completely uncorrelated to the speed of an optimal kernel for an unrelated argument configuration (trials performed on GTX 295).
-Out of 100 random confurations, 38 (not shown) could not even be run using a kernel that was auto-tuned for different arguments.
+\caption{Different arguments call for different kernels:
+{\em left} is the average argument-specific empirical auto-tuned performance across 100 random argument configurations,
+{\em middle} is the average speed of our reference implementation,
+{\em right} is the average speed of kernels auto-tuned for different argument configurations than the one being tested.
+For 38/100 random confurations, the kernel auto-tuned for another problem could not even run on the GTX 295 hardware.
+These points contributed a speed of 0, bringing down the average much lower than the reference.
 Good performance across a variety of inputs requires {\em input-dependent auto-tuning}.
-}
-\label{fig:speedup}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.42]{speedup_295.pdf}
-\includegraphics[scale=.42]{speedup_480.pdf}
-\caption{Speedup of various auto-tuning strategies over the reference
-implementation.
-Each strategy was evaluated the same set of 83 problem configurations, which
-was disjoint from the problem configurations used to build the model for
-model-based hill-climbing.
-The vertical dashed lines are positioned at the geometric mean speedup (XXX?)
-of each strategy. The grid and hill-climbing approaches
-tested 73 and 75 configurations respectively, and took 45?? XXX
-seconds on average.
-The model-based approach tested 0 configuration evaluations, and took 0.05
-seconds on average, even with a naive Python implementation.
 }
@@
 \label{fig:fig_gflop_scatter}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.42]{fig_genX_vader_295.pdf}
-\includegraphics[scale=.42]{fig_genX_munctional0_480.pdf}
-\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
-\label{fig:speedup}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.42]{fig_genX_munctional0_580.pdf}
-\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
-\label{fig:speedup}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.42]{fig_genX_munctional0_1060.pdf}
-\includegraphics[scale=.42]{fig_genX_munctional0_2070.pdf}
-\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
-\label{fig:speedup}
 \end{figure}
@@
 XXXX: Talk about memory transfer requirements vs. speed vs. copy...
+
+\begin{figure}
+\centering
+\includegraphics[scale=.42]{speedup_295.pdf}
+\includegraphics[scale=.42]{speedup_480.pdf}
+\caption{Speedup of various auto-tuning strategies over the reference
+implementation.
+Each strategy was evaluated the same set of 83 problem configurations, which
+was disjoint from the problem configurations used to build the model for
+model-based hill-climbing.
+The vertical dashed lines are positioned at the geometric mean speedup (XXX?)
+of each strategy. The grid and hill-climbing approaches
+tested 73 and 75 configurations respectively, and took 45?? XXX
+seconds on average.
+The model-based approach tested 0 configuration evaluations, and took 0.05
+seconds on average, even with a naive Python implementation.
+}
+\label{fig:speedup}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.42]{fig_genX_vader_295.pdf}
+\includegraphics[scale=.42]{fig_genX_munctional0_480.pdf}
+\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
+\label{fig:speedup}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.42]{fig_genX_munctional0_580.pdf}
+\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
+\label{fig:speedup}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.42]{fig_genX_munctional0_1060.pdf}
+\includegraphics[scale=.42]{fig_genX_munctional0_2070.pdf}
+\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
+\label{fig:speedup}
+\end{figure}

### 2011-12-13 paper/draft1.tex
-sensible +sensitive

@@
 
-% DDC: "sensible to"?  what is meant here? "Sensitive to?"
-The empirical approach is very sensible to the choice of instrumented
+The empirical approach is very sensitive to the choice of instrumented
 optimizations \emph{and} to the search method. The size of the search space is
@@
 
-In this paper, we propose to \emph{learn the model}, empirically, (XXX:
-we'll describe what this means in more details later in the paper, riiiighhht
-???) instead of deriving it analytically. By learning the model, one can
-hope to achieve elements of the best of both approaches and capitalize on the large body of
-literature from the machine learning community.
+In this paper, we propose to \emph{learn the model} using non-linear regression
+modelling techniques instead of constructing a model manually. By learning the model, one can
+hope to achieve elements of the best of both approaches: XXX.
+%and capitalize on the large body of
+%literature from the machine learning community.
 
-Although various statistical prediction techniques have been applied with
+Various statistical prediction techniques have been applied with
 success at compile-time for general programs on various CPU architectures
@@
 hutter2006performance, cavazos2007rapidly, cavazos2008intelligent,
-hartono2009annotation, park2011evaluation, fursin2008milepost}, there is, to
-our knowledge, no fast predictive auto-tuning work that has tried to cover
-simultaneously: (a) more recent multi-core architectures like GPUs
-\citep{schaa2009exploring}, (b) high-performance domain-specific libraries
-\citep{nukada2009auto, li2009note, kamil2010auto}, (c) at run-time
-\citep{klockner2011pycuda, pinto+cox:2011gcg}, and (d) on \emph{many} input
-domains \citep{liu2009cross, grauer2011optimizing}.
+hartono2009annotation, park2011evaluation, fursin2008milepost}.
+Relative to this work our contribution is to show
+how to do fast predictive auto-tuning that satisifies the requirements to:
+(a) handle the variety of recent multi-core architectures like GPUs \citep{schaa2009exploring},
+(b) provide high-performance domain-specific libraries \citep{nukada2009auto, li2009note, kamil2010auto},
+(c) that select good implementations at run-time \citep{klockner2011pycuda, pinto+cox:2011gcg}, and
+(d) for the full input domain of a library routine \citep{liu2009cross, grauer2011optimizing}.
+% XXX: are these last references appropriate
+% XXX: there are 4 versions of the stencil work - consolidate to just the full journal version
 
+\iffalse
 XXX WIIFY here ??? (what's in their for you) XXX in text, say what's in this paper for:
@@
 - cv people
+\fi
+
+% XXX: organisation of the paper?
 
 % auto-tuning is an answer
-
-% XXX: our contributions
-% XXX: WIIFY
-% XXX: organisation of the paper
 
@@
 of several possible implementations is fastest. General-purpose libraries such
-as CUBLAS and CUFFT XXX hide this overhead by auto-tuning offline and using
+as CUBLAS and CUFFT hide this overhead by auto-tuning offline and using
 intimate knowledge of the mathematical problem being solved as well as the
@@
 We call this approach {\em predictive} auto-tuning to contrast it with the standard measurement-based approach which we will call {\em empirical} auto-tuning.
-XXX: We show that the kernel from the GCG chapter can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.
-
-\cite{volkov+demmel:2008}
-\cite{vuduc:2003} % has some stuff about ML for autotuning in ch. 9
+We show that the kernel from the GCG chapter can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.
+%XXX: this has no flow
 

### 2011-12-13 paper/draft1.tex
xxx in intro

@@
 
-The empirical approach is very sensitive to the choice of instrumented
+% DDC: "sensible to"?  what is meant here? "Sensitive to?"
+The empirical approach is very sensible to the choice of instrumented
 optimizations \emph{and} to the search method. The size of the search space is
@@
 of several possible implementations is fastest. General-purpose libraries such
-as CUBLAS and CUFFT hide this overhead by auto-tuning offline and using
+as CUBLAS and CUFFT XXX hide this overhead by auto-tuning offline and using
 intimate knowledge of the mathematical problem being solved as well as the
@@
 We call this approach {\em predictive} auto-tuning to contrast it with the standard measurement-based approach which we will call {\em empirical} auto-tuning.
-We show that the kernel from the GCG chapter can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.
-%XXX: this has no flow
+XXX: We show that the kernel from the GCG chapter can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.
+
+\cite{volkov+demmel:2008}
+\cite{vuduc:2003} % has some stuff about ML for autotuning in ch. 9
 

### 2011-12-13 paper/draft1.tex
draft1
Some more tweaks to language in abstract and intro

--- 2011-12-13 paper/draft1.tex
@@
 \begin{abstract}
-Template-based auto-tuning techniques are widely used and effective for optimizing a
-parametrized GPU code template for a particular computation on particular
-hardware \cite{XXXX}.  A drawback of this approach is that exhaustive, or even thorough, auto-tuning requires
+Auto-tuning of parameteric, instrumented code templates has proven to be a highly effective technique for optimizing a GPU code for a particular computation on particular
+hardware \cite{XXXX}.  
+% DDC: I'm still struggling with the above sentence.  Feel free to substitute your own ideas
+However, a drawback of this approach is that exhaustive, or even thorough, auto-tuning requires
 compiling many kernels and calling each one many times, and this process is slow.
-Furthermore, library abstraction boundaries provide operations such as image
-filtering and matrix multiplication, which actually correspond to a large set
+Furthermore, it is desirable to provide users with unified library abstraction boundaries for operations such as image
+filtering and matrix multiplication, even those these operations actually correspond to a large set
 of potential problem configurations with a wide variety of memory access
@@
 predicts how much faster this kernel will be than a reference baseline.
-Combinatorial optimization strategies that would normally implement auto-tuning
-by evaluating kernel configurations on the real hardware are orders of
-magnitude faster when evaluating the surrogate regression model instead.  We
+Combinatorial optimization strategies for auto-tuning that would normally 
+require evaluating large number of kernel configurations on the real hardware, are made orders of
+magnitude faster by evaluating the surrogate regression model instead.  We
 validate our approach using the filterbank correlation kernel described in
 \citet{pinto+cox:2011gcg}, where we find that 0.1 seconds (XXX) of hill
-climbing on the regression model ({\em predictive} auto-tuning) can achieve an
-average of 95\% of the speed brought by 120 seconds of empirical auto-tuning
-XXX.  Our approach is not specific to filterbank correlation, nor to GPU
-kernel auto-tuning; the approach of using a non-linear regression model on top
-of simple features applies to a variety of problem types, kernel types, and
+climbing on the regression model (which we dub ``{\em predictive} auto-tuning'') can achieve an
+average of 95\% of the speed-up brought by 120 seconds of empirical auto-tuning
+XXX.  Our approach is not specific to filterbank correlation, nor even to GPU
+kernel auto-tuning, and can be applied to almost any templated-code optimization problem, spanning a wide variety of problem types, kernel types, and
 platforms.  \end{abstract}
@@
 
-In this rapidly evolving landscape, the construction of \emph{general}
-development tools and libraries that \emph{fully utilize} system resources
+In this rapidly evolving landscape, the construction of general
+development tools and libraries that fully utilize system resources
 remains a daunting task. Even within specialized architectures from the
@@
 abstraction, it has been less suited for highly-specialized kernels. This
-approach has been dominant in the compiler community. As a result, it has
+approach has been dominant in the compiler community, and as a result, it has
 generally been applied at compile-time where important run-time characteristics

### 2011-12-13 paper/draft1.tex
A few minor edits

@@
 \begin{abstract}
-Auto-tuning of parameteric, instrumented code templates has proven to be a highly effective technique for optimizing a GPU code for a particular computation on particular
+Auto-tuning of parameteric, instrumented code templates has proven to be a highly effective technique for optimizing a GPU kernel for a particular computation on particular
 hardware \cite{XXXX}.  
@@
 hartono2009annotation, park2011evaluation, fursin2008milepost}.
-Relative to this work our contribution is to show
+Relative to this work, our contribution is to show
 how to do fast predictive auto-tuning that satisifies the requirements to:

### 2011-12-13 paper/draft1.tex
Add some future work blurbs

--- 2011-12-13 paper/draft1.tex
@@
 
+\section{Future Work}
+
+more features (this is the hardest, and most manual part, may require
+domain-specific knowledge, but not necessariyly, e.g. microbenchmarking)
+* features derived from micro-benchmarking \citep{wong2010demystifying}
+* features derived from performance limiter analysis \citep{micikevicius2010analysis}
+* more instrumentation
+* more data-layout
+* more threading strategies to lower latency
+
+online distributed auto-tuning with learning, especially w/ HT-based
+experiments \cite{HT} our method is indeed very well suited for parallel
+explorating of multiple configurations for learning (in a cluster, e.g. in HT
+context) this may get us \emph{more} than linear speed ups for better accuracy,
+especially when more features will be present (see above)
+
+interpret the learned model and confront the performance against theoreticial
+(or approximated) models of mem/com/compute, arithmetic intensity, etc. to
+understand performance limiters, the hardware, etc.
+
 \begin{figure}

### 2011-12-13 paper/draft1.tex
Note about using using percent and not X
A few comments and minor edits

--- 2011-12-13 paper/draft1.tex
@@
 \cite{cox+pinto:2011gcg} recommends a particular hyper-parameter implementation
-based on empirical autotuning on several platforms, that achieves good
+based on empirical auto-tuning on several platforms, that achieves good
 performance on a variety of GPUs from older-generation cards such as the
@@
 % XXX: what is reference?  Why was it chosen?
+% DDC: wasn't the reference implementation hand-tuned?
+% might also want to say something like:
+% Given that parameters were hand-chosen for the reference kernel, no claims are made as to the optimality nor universality of this reference (indeed, different programmers would undoubtedly arrive at different results).
+We provide this kernel configuration simply as a reasonable indicator of typical performance made possible by ad hoc experimentation with parameters.
 
@@
 
-\subsection{Model-based Autotuning}
+\subsection{Predictive Auto-tuning}
 

### 2011-12-13 paper/draft1.tex

@@
 platforms.  \end{abstract}
+
+XXX: use 300\% boost instead of 3x
 

### 2011-12-13 paper/draft1.tex
Adding more brain dumps

--- 2011-12-13 paper/draft1.tex
@@
 %\setlength{\bibsep}{5pt}
-\bibliography{local}
-\end{document}
 
@@
 
+we believe that our approach is the most practial one to get high perf on
+recent hardware, since there is no consensus on hw features (and it won't
+change soon)
+
+auto-tuning required hardward expertise *and* application-domain knowledge, we
+want to remove these restrictions ultimately we want a ml-based approach to
+code optimization that will where a set of generic but deep code
+transformations, could be templated and injected in any domain-specific
+library. this will reduce the need for specialized expertise across the range
+of targetted computing systems.
+
 \section{Future Work}
@@
 * more threading strategies to lower latency
+(finding a good representation is key, again...)
 
@@
 \end{figure}
+
+\bibliography{local}
+\end{document}

### 2011-12-13 paper/draft1.tex
draft1

--- 2011-12-13 paper/draft1.tex
@@
 \begin{abstract}
-Auto-tuning of parameteric, instrumented code templates has proven to be a highly effective technique for optimizing a GPU kernel for a particular computation on particular
-hardware \cite{XXXX}.  
-% DDC: I'm still struggling with the above sentence.  Feel free to substitute your own ideas
-However, a drawback of this approach is that exhaustive, or even thorough, auto-tuning requires
+Template-based auto-tuning techniques are widely used and effective for optimizing a
+parametrized GPU code template for a particular computation on particular
+hardware \cite{XXXX}.  A drawback of this approach is that exhaustive, or even thorough, auto-tuning requires
 compiling many kernels and calling each one many times, and this process is slow.
-Furthermore, it is desirable to provide users with unified library abstraction boundaries for operations such as image
-filtering and matrix multiplication, even those these operations actually correspond to a large set
+Furthermore, library abstraction boundaries provide operations such as image
+filtering and matrix multiplication, which actually correspond to a large set
 of potential problem configurations with a wide variety of memory access
@@
 predicts how much faster this kernel will be than a reference baseline.
-Combinatorial optimization strategies for auto-tuning that would normally 
-require evaluating large number of kernel configurations on the real hardware, are made orders of
-magnitude faster by evaluating the surrogate regression model instead.  We
+Combinatorial optimization strategies that would normally implement auto-tuning
+by evaluating kernel configurations on the real hardware are orders of
+magnitude faster when evaluating the surrogate regression model instead.  We
 validate our approach using the filterbank correlation kernel described in
 \citet{pinto+cox:2011gcg}, where we find that 0.1 seconds (XXX) of hill
-climbing on the regression model (which we dub ``{\em predictive} auto-tuning'') can achieve an
-average of 95\% of the speed-up brought by 120 seconds of empirical auto-tuning
-XXX.  Our approach is not specific to filterbank correlation, nor even to GPU
-kernel auto-tuning, and can be applied to almost any templated-code optimization problem, spanning a wide variety of problem types, kernel types, and
+climbing on the regression model ({\em predictive} auto-tuning) can achieve an
+average of 95\% of the speed brought by 120 seconds of empirical auto-tuning
+XXX.  Our approach is not specific to filterbank correlation, nor to GPU
+kernel auto-tuning; the approach of using a non-linear regression model on top
+of simple features applies to a variety of problem types, kernel types, and
 platforms.  \end{abstract}
@@
 
-In this rapidly evolving landscape, the construction of general
-development tools and libraries that fully utilize system resources
+In this rapidly evolving landscape, the construction of \emph{general}
+development tools and libraries that \emph{fully utilize} system resources
 remains a daunting task. Even within specialized architectures from the
@@
 abstraction, it has been less suited for highly-specialized kernels. This
-approach has been dominant in the compiler community, and as a result, it has
+approach has been dominant in the compiler community. As a result, it has
 generally been applied at compile-time where important run-time characteristics
@@
 hartono2009annotation, park2011evaluation, fursin2008milepost}.
-Relative to this work, our contribution is to show
+Relative to this work our contribution is to show
 how to do fast predictive auto-tuning that satisifies the requirements to:
@@
 
-% XXX: organisation of the paper?
+The paper is organized as follows:
+Section~\ref{sec:brtree} describes the boosted regression tree model and the procedure for fitting it to empirical timing data.
+Section~\ref{sec:kernel} describes the sort of kernel we employ for our benchmarking,
+Section~\ref{sec:exp} presents the results of our benchmarking experiments, which
+compare a reference implementation to a) empirical auto-tuning over a domain-specific grid,
+b) empirical auto-tuning over a hill-climbing search, and
+c) predictive autotuning.
+%Section~\ref{sec:discussion}.
+Section~\ref{sec:previous} summarizes previous and related work involving  machine learning with performance auto-tuning.
+Section~\ref{sec:conclusion} summarizes our findings and outlines directions for future work.
 
-% auto-tuning is an answer
-
-% XXX: see gdoc and merge this below
-
-\subsection{Case Study}
-Current languages and software engineering practices have conspired to make it difficult
-to provide high-performance libraries for general mathematical operations on GPU devices.
-An important approach in implementing high-performance GPU codes has been auto-tuning,
-which is fundamentally slow. Auto-tuning requires running a large number of
-sub-compilations and representative computations to {\em measure} which
-of several possible implementations is fastest. General-purpose libraries such
-as CUBLAS and CUFFT hide this overhead by auto-tuning offline and using
-intimate knowledge of the mathematical problem being solved as well as the
-kernel implementations to heuristically choose a good computation strategy quickly when a
-user asks for e.g. a matrix product of particular arguments.
-Designing and testing libraries of this form is time-consuming, difficult, and provides little
-reusable infrastructure to assist in the implementation of different computations.
-For instance, even if one has already developed an auto-tuned library for dense matrix algebra, 
-one must basically start from scratch to develop an auto-tuned library for
-stencil computations or sparse matrix operations.
+\section{Predictive Auto-Tuning}
 
@@
 \end{algorithm}
-\caption{Pseudo-code template for on-demand auto-tuning.
+\caption{Pseudo-code template for empirical and predictive auto-tuning.
 Empirical auto-tuning (above) is inevitably slow
@@
 
-This work shows that auto-tuning can be accelerated by orders of
-magnitude by using a regression model built offline as a surrogate for actual
-computations on the real hardware.
-The general form of an auto-tuning based library routine is illustrated in Figure~\ref{fig:general_autotune} (top).
-We will use filterbank correlation (Section~\ref{sec:fbcorr}) as our running example, but the approach is both general and powerful.
-An auto-tuning based routine must operate on three sets of variables:
-\begin{description}
-\item[$\mathcal{A}$:] task description (argument shapes, physical layout)
-\item[$\mathcal{B}$:] implementation description (auto-tuning parameters)
-\item[$\mathcal{C}$:] platform description (capabilities, micro-benchmarks)
-\end{description}
-The hypothetical auto-tuning routine described at the top of Figure~\ref{fig:general_autotune}
-might take many minutes or hours to perform the argmin at step 3 (during which time it computes the desired result many times!)
-so it would be all but unusable as a normal library routine.
-However, the form of the auto-tuning routine suggests the potential for enormous acceleration:
-if only there were a fast (even approximate) surrogate for the costly $MeasureTime(\cdot)$ function,
-then the argmin could be done in a fraction of a second and the routine could be used normally (Figure~\ref{fig:general_autotune}, bottom).
+XXX
+Figure~\ref{fig:general_autotune}, which permits doing on-demand auto-tuning
+for novel (not already auto-tuned) argument combinations.
+We call this approach {\em predictive auto-tuning}.
 
-Our contribution of the present work is to show that boosted regression tree models,
-a powerful machine learning technique for function approximation, can meet the
-requirements of this application.
-We call this approach {\em predictive} auto-tuning to contrast it with the standard measurement-based approach which we will call {\em empirical} auto-tuning.
-We show that the kernel from the GCG chapter can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.
-%XXX: this has no flow
+\subsection{Learning a Regression Model}
 
-\subsection{Filterbank Correlation}
+The heart of predictive auto-tuning is a regression model that acts as surrogate
+for a hand-written hardware model or empirical timing estimates.
+In our experiments, this regression model is fit to empirically measured timing information for {\em some} elements of the configuration space, and the regression model interpolates and extrapolates that timing across the remainder of the space.
+To fit this model, we
+form a training set $\mathcal{X}, \mathcal{Y}$ where
+each point $x^{(i)} \in \mathcal{X}$ is a tuple $(a^{(i)}, b^{(i)}, c^{(i)})$
+and each target $y^{(i)} \in \mathcal{Y}$  reflects the speed of implementation $b^{(i)}$
+on inputs $a^{(i)}$ on platform $c^{(i)}$.
+
+The effectiveness of predictive auto-tuning depends on the mapping between the raw kernel timings $t(a, b, c)$ (i.e. in seconds) and the utility $y$ associated with that timing.
+Anticipating that regression involves minimizing the squared error of our predictor (see Eq.~\ref{XXXbelow}) it is important to choose $y$ so that differences of a given numerical {\em magnitude} correspond to {\em improvements} of a certain utility.
+In program optimization we are interested in improving the speedup over a reference implementation $b^{(\mathrm{ref})}$, so it is natural to choose
+\begin{equation}
+y^{(i)}
+= \log\left(\frac{\mathrm{speed}(a, b, c)}{\mathrm{speed}(a, b^{(\mathrm{ref})}, c)} \right)
+= \log\left(\frac{t(a, b^{(\mathrm{ref})}, c)}{t(a, b, c)} \right)
+\end{equation}
+
+One wrinkle compared with a standard regression framework is that not all kernel implementations $b$ are {\em valid} on all input configurations.
+
+\subsection{Regression Trees}
+\label{sec:brtree}
+
+A regression tree is a piece-wise constant function from one vector space to another,
+that works by recursively subdividing the input space into constant regions~\citep{breiman, textbook?}.
+They are widely used in statistics and data-mining applications because the fitting algorithm is quick and reliable, and the form of the tree can provide insight into the relevant input variables.
+We use a standard fitting procedure, which takes a set of
+$(x, y)\in \mathbb{R}^k \times \mathbb{R}$ pairs and constructs a tree with a low mean squared error.
+To construct each node of a regression tree,
+we sort the set $\mathcal{D}$ of ($x,y$) pairs along each of the $k$ features to find the best partitioning $f_{i,\gamma}$ of the input space along feature $i$ at point $\gamma$ (Eqs.~\ref{eq:figamma}, \ref{eq:treeloss}).
+\begin{align}
+    f_{i,\gamma}(x) &=
+    \begin{cases}
+        \alpha  &\text{if $x_i < \gamma$} \\
+        \beta &\text{if $x_i \geq \gamma$}
+    \end{cases}
+    \label{eq:figamma}\\
+    i^*, \gamma^* &= \operatorname*{argmin}_{i, \gamma} \mathbb{\hat E}\left[ (y - f_{i,\gamma}(x))^2 \right]
+    \label{eq:treeloss}
+\end{align}
+One disadvantage of the regression tree is that it does not make full use of broad patterns in the data -- each partition formed by the fitting procedure is fit independently in the recursive training procedure,
+so it impossible for the model to extract more than one bit of information from each training partition (XXX unclear).
+This disadvantage is mitigated to a significant extent by the practice of {\em boosting}~\citet{XXX}.
+
+\subsection{Boosted Regression Trees}
+
+Boosting is an iterative procedure for constructing an {\em ensemble} of regression trees that is coordinated to fit training examples as accurately as possible.
+In a recent empirical study of a range of machine learning regressino problems,
+boosted decision trees were found to be among the best and easiest models to apply~\cite{caruana:2010}.
+On each boosting iteration, a regression tree is fit to the residual error remaining
+after all previously-fit models have made their predictions.
+There are essentially three parameters that control the boosted regression tree
+training procedure:
+1) the depth of tree constructed on each boosting iteration,
+2) the minimum number of examples to allow at a regression tree leaf, and
+3) the number of trees constructed by boosting.
+We did not attempt a systematic study of the effect of these variables on performance.
+We chose a maximum depth of 4,
+a minimum number of examples of 10,
+and 100 iterations of boosting.
+
+\section{Case Study: Filterbank Correlation}
+
 \label{sec:fbcorr}
@@
 an important subroutine in many image processing applications. It has a relatively
-high arithmetic intensity which makes it a natural fit for GPU platforms~\citep{pinto+cox:2011gcg}. XXX cite some computer vision papers.
+high arithmetic intensity which makes it a natural fit for GPU platforms~\citep{pinto+cox:2011gcg}.
+%XXX cite some computer vision papers.
 
@@
 However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
-related stencil operations XXX, it is challenging to provide an implementation
+related stencil operations, %XXX
+it is challenging to provide an implementation
 or even an implementation strategy that provides satisfactory performance
@@
 
-XXX \cite{datta:2009}
+XXXXXXXXXXX
+Current languages and software engineering practices have conspired to make it difficult
+to provide high-performance libraries for general mathematical operations on GPU devices.
+An important approach in implementing high-performance GPU codes has been auto-tuning,
+which is fundamentally slow. Auto-tuning requires running a large number of
+sub-compilations and representative computations to {\em measure} which
+of several possible implementations is fastest. General-purpose libraries such
+as CUBLAS and CUFFT hide this overhead by auto-tuning offline and using
+intimate knowledge of the mathematical problem being solved as well as the
+kernel implementations to heuristically choose a good computation strategy quickly when a
+user asks for e.g. a matrix product of particular arguments.
+Designing and testing libraries of this form is time-consuming, difficult, and provides little
+reusable infrastructure to assist in the implementation of different computations.
+For instance, even if one has already developed an auto-tuned library for dense matrix algebra, 
+one must basically start from scratch to develop an auto-tuned library for
+stencil computations or sparse matrix operations.
 
-This paper contributes a model-based approach to auto-tuning in which the model
-is determined almost entirely using machine learning rather than domain
-knowledge.  We show that even though our timing model has no built-in knowledge of GPU
-devices, or our kernel, or filterbank correlation, it is able to provide fast
-and accurate timing estimates.
-We show that a model built by boosting regression trees
-is fast and accurate enough to be used profitably
-in the algorithm schema shown at the bottom of
-Figure~\ref{fig:general_autotune}, which permits doing on-demand auto-tuning
-for novel (not already auto-tuned) argument combinations.
-We call this approach {\em predictive auto-tuning}.
+This work shows that auto-tuning can be accelerated by orders of
+magnitude by using a regression model built offline as a surrogate for actual
+computations on the real hardware.
+The general form of an auto-tuning based library routine is illustrated in Figure~\ref{fig:general_autotune} (top).
+We will use filterbank correlation (Section~\ref{sec:fbcorr}) as our running example, but the approach is both general and powerful.
+An auto-tuning based routine must operate on three sets of variables:
+\begin{description}
+\item[$\mathcal{A}$:] task description (argument shapes, physical layout)
+\item[$\mathcal{B}$:] implementation description (auto-tuning parameters)
+\item[$\mathcal{C}$:] platform description (capabilities, micro-benchmarks)
+\end{description}
+The hypothetical auto-tuning routine described at the top of Figure~\ref{fig:general_autotune}
+might take many minutes or hours to perform the argmin at step 3 (during which time it computes the desired result many times!)
+so it would be all but unusable as a normal library routine.
+However, the form of the auto-tuning routine suggests the potential for enormous acceleration:
+if only there were a fast (even approximate) surrogate for the costly $MeasureTime(\cdot)$ function,
+then the argmin could be done in a fraction of a second and the routine could be used normally (Figure~\ref{fig:general_autotune}, bottom).
 
-The paper is organized as follows:
-Section~\ref{sec:brtree} describes the boosted regression tree model and the procedure for fitting it to empirical timing data.
-Section~\ref{sec:kernel} describes the sort of kernel we employ for our benchmarking,
-Section~\ref{sec:exp} presents the results of our benchmarking experiments, which
-compare a reference implementation to a) empirical auto-tuning over a domain-specific grid,
-b) empirical auto-tuning over a hill-climbing search, and
-c) predictive autotuning.
-%Section~\ref{sec:discussion}.
-Section~\ref{sec:previous} summarizes previous and related work involving  machine learning with performance auto-tuning.
-Section~\ref{sec:conclusion} summarizes our findings and outlines directions for future work.
-
-\section{Boosted Regression Trees}
-\label{sec:brtree}
-
-A regression tree is a piece-wise constant function from one vector space to another,
-that works by recursively subdividing the input space into constant regions~\citep{breiman, textbook?}.
-They are widely used in statistics and data-mining applications because the fitting algorithm is quick and reliable, and the form of the tree can provide insight into the relevant input variables.
-We use a standard fitting procedure, which takes a set of
-$(x, y)\in \mathbb{R}^k \times \mathbb{R}$ pairs and constructs a tree with a low mean squared error.
-To construct each node of a regression tree,
-we sort the set $\mathcal{D}$ of ($x,y$) pairs along each of the $k$ features to find the best partitioning $f_{i,\gamma}$ of the input space along feature $i$ at point $\gamma$ (Eqs.~\ref{eq:figamma}, \ref{eq:treeloss}).
-\begin{align}
-    f_{i,\gamma}(x) &=
-    \begin{cases}
-        \alpha  &\text{if $x_i < \gamma$} \\
-        \beta &\text{if $x_i \geq \gamma$}
-    \end{cases}
-    \label{eq:figamma}\\
-    i^*, \gamma^* &= \operatorname*{argmin}_{i, \gamma} \mathbb{\hat E}\left[ (y - f_{i,\gamma}(x))^2 \right]
-    \label{eq:treeloss}
-\end{align}
-One disadvantage of the regression tree is that it does not make full use of broad patterns in the data -- each partition formed by the fitting procedure is fit independently in the recursive training procedure,
-so it impossible for the model to extract more than one bit of information from each training partition (XXX unclear).
-This disadvantage is mitigated to a significant extent by the practice of {\em boosting}~\citet{XXX}.
-
-Boosting is an iterative procedure for constructing an {\em ensemble} of regression trees that is coordinated to fit training examples as accurately as possible.
-In a recent empirical study of a range of machine learning regressino problems,
-boosted decision trees were found to be among the best and easiest models to apply~\cite{caruana:2010}.
-On each boosting iteration, a regression tree is fit to the residual error remaining
-after all previously-fit models have made their predictions.
-There are essentially three parameters that control the boosted regression tree
-training procedure:
-1) the depth of tree constructed on each boosting iteration,
-2) the minimum number of examples to allow at a regression tree leaf, and
-3) the number of trees constructed by boosting.
-We did not attempt a systematic study of the effect of these variables on performance.
-We chose a maximum depth of 4,
-a minimum number of examples of 10,
-and 100 iterations of boosting.
+Our contribution of the present work is to show that boosted regression tree models,
+a powerful machine learning technique for function approximation, can meet the
+requirements of this application.
+We call this approach {\em predictive} auto-tuning to contrast it with the standard measurement-based approach which we will call {\em empirical} auto-tuning.
+We show that the kernel from the GCG chapter can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.
+%XXX: this has no flow
 
@@
 \cite{cox+pinto:2011gcg} recommends a particular hyper-parameter implementation
-based on empirical auto-tuning on several platforms, that achieves good
+based on empirical autotuning on several platforms, that achieves good
 performance on a variety of GPUs from older-generation cards such as the
@@
 % XXX: what is reference?  Why was it chosen?
-% DDC: wasn't the reference implementation hand-tuned?
-% might also want to say something like:
-% Given that parameters were hand-chosen for the reference kernel, no claims are made as to the optimality nor universality of this reference (indeed, different programmers would undoubtedly arrive at different results).
-We provide this kernel configuration simply as a reasonable indicator of typical performance made possible by ad hoc experimentation with parameters.
 
@@
 \includegraphics[scale=.42]{allstars_mixup_295.pdf}
+% XXX: ERROR BARS
 \caption{Different arguments call for different kernels:
@@
 
-\subsection{Predictive Auto-tuning}
+\subsection{Model-based Autotuning}
 

### 2011-12-13 paper/draft1.tex
Minor edits to the abstract

@@
 \begin{abstract}
-Auto-tuning of parameteric, instrumented code templates has proven to be a highly effective technique for optimizing a GPU kernel for a particular computation on particular
+Auto-tuning of parameteric, instrumented code templates has proven to be a
+highly effective technique for optimizing a GPU kernel for a particular
+computation on particular
 hardware \cite{XXXX}.  
@@
 compiling many kernels and calling each one many times, and this process is slow.
-Furthermore, it is desirable to provide users with unified library abstraction boundaries for operations such as image
+Furthermore, it is desirable to provide users with unified library abstraction
+boundaries for operations such as image
 filtering and matrix multiplication, even those these operations actually correspond to a large set
@@
 climbing on the regression model (which we dub ``{\em predictive} auto-tuning'') can achieve an
-average of 95\% of the speed-up brought by 120 seconds of empirical auto-tuning
+average of 95\% of the speed-up brought by minutes of empirical auto-tuning
 XXX.  Our approach is not specific to filterbank correlation, nor even to GPU
-kernel auto-tuning, and can be applied to almost any templated-code optimization problem, spanning a wide variety of problem types, kernel types, and
-platforms.  \end{abstract}
+kernel auto-tuning, and can be applied to almost any templated-code
+optimization problem, spanning a wide variety of problem types, kernel types,
+and platforms.
+
+\end{abstract}
 

### 2011-12-13 paper/draft1.tex
draft1

--- 2011-12-13 paper/draft1.tex
@@
 for a hand-written hardware model or empirical timing estimates.
-In our experiments, this regression model is fit to empirically measured timing information for {\em some} elements of the configuration space, and the regression model interpolates and extrapolates that timing across the remainder of the space.
+In our experiments this regression model fits empirically measured timing information for a {\em subset} of the configuration space and interpolates / extrapolates that timing across the remainder of the space.
 To fit this model, we

### 2011-12-13 paper/draft1.tex

--- 2011-12-13 paper/draft1.tex
@@
 
-This work shows that auto-tuning can be accelerated by orders of
-magnitude by using a regression model built offline as a surrogate for actual
-computations on the real hardware.
-The general form of an auto-tuning based library routine is illustrated in Figure~\ref{fig:general_autotune} (top).
-We will use filterbank correlation (Section~\ref{sec:fbcorr}) as our running example, but the approach is both general and powerful.
-An auto-tuning based routine must operate on three sets of variables:
-\begin{description}
-\item[$\mathcal{A}$:] task description (argument shapes, physical layout)
-\item[$\mathcal{B}$:] implementation description (auto-tuning parameters)
-\item[$\mathcal{C}$:] platform description (capabilities, micro-benchmarks)
-\end{description}
-The hypothetical auto-tuning routine described at the top of Figure~\ref{fig:general_autotune}
-might take many minutes or hours to perform the argmin at step 3 (during which time it computes the desired result many times!)
-so it would be all but unusable as a normal library routine.
-However, the form of the auto-tuning routine suggests the potential for enormous acceleration:
-if only there were a fast (even approximate) surrogate for the costly $MeasureTime(\cdot)$ function,
-then the argmin could be done in a fraction of a second and the routine could be used normally (Figure~\ref{fig:general_autotune}, bottom).
+XXX
+Figure~\ref{fig:general_autotune}, which permits doing on-demand auto-tuning
+for novel (not already auto-tuned) argument combinations.
+We call this approach {\em predictive auto-tuning}.
 
-Our contribution of the present work is to show that boosted regression tree models,
-a powerful machine learning technique for function approximation, can meet the
-requirements of this application.
-We call this approach {\em predictive} auto-tuning to contrast it with the standard measurement-based approach which we will call {\em empirical} auto-tuning.
-We show that the kernel from the GCG chapter can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.
-%XXX: this has no flow
+\subsection{Learning a Regression Model}
 
-\subsection{Filterbank Correlation}
+The heart of predictive auto-tuning is a regression model that acts as surrogate
+for a hand-written hardware model or empirical timing estimates.
+In our experiments this regression model fits empirically measured timing information for a {\em subset} of the configuration space and interpolates / extrapolates that timing across the remainder of the space.
+To fit this model, we
+form a training set $\mathcal{X}, \mathcal{Y}$ where
+each point $x^{(i)} \in \mathcal{X}$ is a tuple $(a^{(i)}, b^{(i)}, c^{(i)})$
+and each target $y^{(i)} \in \mathcal{Y}$  reflects the speed of implementation $b^{(i)}$
+on inputs $a^{(i)}$ on platform $c^{(i)}$.
+
+The effectiveness of predictive auto-tuning depends on the mapping between the raw kernel timings $t(a, b, c)$ (i.e. in seconds) and the utility $y$ associated with that timing.
+Anticipating that regression involves minimizing the squared error of our predictor (see Eq.~\ref{XXXbelow}) it is important to choose $y$ so that differences of a given numerical {\em magnitude} correspond to {\em improvements} of a certain utility.
+In program optimization we are interested in improving the speedup over a reference implementation $b^{(\mathrm{ref})}$, so it is natural to choose
+\begin{equation}
+y^{(i)}
+= \log\left(\frac{\mathrm{speed}(a, b, c)}{\mathrm{speed}(a, b^{(\mathrm{ref})}, c)} \right)
+= \log\left(\frac{t(a, b^{(\mathrm{ref})}, c)}{t(a, b, c)} \right)
+\end{equation}
+
+One wrinkle compared with a standard regression framework is that not all kernel implementations $b$ are {\em valid} on all input configurations.
+
+\subsection{Regression Trees}
+\label{sec:brtree}
+
+A regression tree is a piece-wise constant function from one vector space to another,
+that works by recursively subdividing the input space into constant regions~\citep{breiman, textbook?}.
+They are widely used in statistics and data-mining applications because the fitting algorithm is quick and reliable, and the form of the tree can provide insight into the relevant input variables.
+We use a standard fitting procedure, which takes a set of
+$(x, y)\in \mathbb{R}^k \times \mathbb{R}$ pairs and constructs a tree with a low mean squared error.
+To construct each node of a regression tree,
+we sort the set $\mathcal{D}$ of ($x,y$) pairs along each of the $k$ features to find the best partitioning $f_{i,\gamma}$ of the input space along feature $i$ at point $\gamma$ (Eqs.~\ref{eq:figamma}, \ref{eq:treeloss}).
+\begin{align}
+    f_{i,\gamma}(x) &=
+    \begin{cases}
+        \alpha  &\text{if $x_i < \gamma$} \\
+        \beta &\text{if $x_i \geq \gamma$}
+    \end{cases}
+    \label{eq:figamma}\\
+    i^*, \gamma^* &= \operatorname*{argmin}_{i, \gamma} \mathbb{\hat E}\left[ (y - f_{i,\gamma}(x))^2 \right]
+    \label{eq:treeloss}
+\end{align}
+One disadvantage of the regression tree is that it does not make full use of broad patterns in the data -- each partition formed by the fitting procedure is fit independently in the recursive training procedure,
+so it impossible for the model to extract more than one bit of information from each training partition (XXX unclear).
+This disadvantage is mitigated to a significant extent by the practice of {\em boosting}~\citet{XXX}.
+
+\subsection{Boosted Regression Trees}
+
+Boosting is an iterative procedure for constructing an {\em ensemble} of regression trees that is coordinated to fit training examples as accurately as possible.
+In a recent empirical study of a range of machine learning regressino problems,
+boosted decision trees were found to be among the best and easiest models to apply~\cite{caruana:2010}.
+On each boosting iteration, a regression tree is fit to the residual error remaining
+after all previously-fit models have made their predictions.
+There are essentially three parameters that control the boosted regression tree
+training procedure:
+1) the depth of tree constructed on each boosting iteration,
+2) the minimum number of examples to allow at a regression tree leaf, and
+3) the number of trees constructed by boosting.
+We did not attempt a systematic study of the effect of these variables on performance.
+We chose a maximum depth of 4,
+a minimum number of examples of 10,
+and 100 iterations of boosting.
+
+\section{Case Study: Filterbank Correlation}
+
 \label{sec:fbcorr}

### 2011-12-13 paper/draft1.tex
minor edits

--- 2011-12-13 paper/draft1.tex
@@
 these spaces can be highly discontinuous \citep{ryoo2008program}, difficult
-to explore and quasi-optimal solutions lie at the edge of ``performance
-cliffs'' induced by device-specific hard constraints (e.g. register file
-size and shared memory size).
+to explore, and quasi-optimal solutions lie at the edge of ``performance
+cliffs'' induced by hard device-specific constraints (e.g. register file
+size or low-latency cache size).
 

### 2011-12-13 paper/draft1.tex
Add brewer ref, if you find a way to include it
rearranging sec 2

--- 2011-12-13 paper/draft1.tex
@@
 One strategy for addressing these challenges is to use one of a variety of automatic
-methods known collectively as ``auto-tuning''. Two major auto-tuning approaches have emerged
+methods known collectively as ``auto-tuning.'' Two major auto-tuning approaches have emerged
 in the extensive literature covering the subject (see surveys in
@@
 
+This work shows that auto-tuning can be accelerated by orders of
+magnitude by using a regression model built offline as a surrogate for actual
+computations on the real hardware.
+The general form of an auto-tuning based library routine is illustrated in Figure~\ref{fig:general_autotune} (top).
+An auto-tuning based routine must operate on three sets of variables:
+\begin{description}
+\item[$\mathcal{A}$:] task description (argument shapes, physical layout)
+\item[$\mathcal{B}$:] implementation description (auto-tuning parameters)
+\item[$\mathcal{C}$:] platform description (capabilities, micro-benchmarks)
+\end{description}
+The hypothetical auto-tuning routine described at the top of Figure~\ref{fig:general_autotune}
+might take many minutes or hours to perform the argmin at step 3 (during which time it computes the desired result many times!)
+so it would not be suitable for a normal library subroutine implementation.
+However, the form of the auto-tuning routine suggests the potential for enormous acceleration:
+if only there were a fast (even approximate) surrogate for the costly $MeasureTime(\cdot)$ function,
+then the argmin could be done in a fraction of a second and the routine could be used normally (Figure~\ref{fig:general_autotune}, bottom).
+
 \begin{figure}[!ht]
@@
 \Aitem $c \setto PlatformFeatures()$
-\Aitem $f \setto LoadTimingModel()$
+\Aitem $f \setto TimingModel()$
 \Aitem $b^* \setto \operatorname*{argmin}_{b \in \mathcal{B}} f(a, b, c)$
@@
 \end{figure}
-
-XXX
-Figure~\ref{fig:general_autotune}, which permits doing on-demand auto-tuning
-for novel (not already auto-tuned) argument combinations.
-We call this approach {\em predictive auto-tuning}.
 
@@
 
-One wrinkle compared with a standard regression framework is that not all kernel implementations $b$ are {\em valid} on all input configurations.
+One unique aspect of our setting compared with standard regression
+is that not all kernel implementations ($b$) are {\em valid}
+for all input configurations. One option for dealing with these invalid configurations
+would be to simply omit them from $\mathcal{X}$ and $\mathcal{Y}$, but that would lead to a biased regression model.
+Instead, we chose to associate invalid $(a, b, c)$ tuples with a constant $y = \zeta$.
+It makes sense to choose $\zeta < 0$ so that invalid configurations are treated as being worse than the reference,
+but the question of how much worse these points should be is an empirical one.
+Our experiments evaluate $\log(0.01) \leq \zeta \leq \log(.99)$.
 
@@
 
-\section{Case Study: Filterbank Correlation}
+\section{Filterbank Correlation}
 
@@
 XXXXXXXXXXX
-Current languages and software engineering practices have conspired to make it difficult
-to provide high-performance libraries for general mathematical operations on GPU devices.
-An important approach in implementing high-performance GPU codes has been auto-tuning,
-which is fundamentally slow. Auto-tuning requires running a large number of
-sub-compilations and representative computations to {\em measure} which
-of several possible implementations is fastest. General-purpose libraries such
-as CUBLAS and CUFFT hide this overhead by auto-tuning offline and using
-intimate knowledge of the mathematical problem being solved as well as the
-kernel implementations to heuristically choose a good computation strategy quickly when a
-user asks for e.g. a matrix product of particular arguments.
-Designing and testing libraries of this form is time-consuming, difficult, and provides little
-reusable infrastructure to assist in the implementation of different computations.
-For instance, even if one has already developed an auto-tuned library for dense matrix algebra, 
-one must basically start from scratch to develop an auto-tuned library for
-stencil computations or sparse matrix operations.
-
-This work shows that auto-tuning can be accelerated by orders of
-magnitude by using a regression model built offline as a surrogate for actual
-computations on the real hardware.
-The general form of an auto-tuning based library routine is illustrated in Figure~\ref{fig:general_autotune} (top).
-We will use filterbank correlation (Section~\ref{sec:fbcorr}) as our running example, but the approach is both general and powerful.
-An auto-tuning based routine must operate on three sets of variables:
-\begin{description}
-\item[$\mathcal{A}$:] task description (argument shapes, physical layout)
-\item[$\mathcal{B}$:] implementation description (auto-tuning parameters)
-\item[$\mathcal{C}$:] platform description (capabilities, micro-benchmarks)
-\end{description}
-The hypothetical auto-tuning routine described at the top of Figure~\ref{fig:general_autotune}
-might take many minutes or hours to perform the argmin at step 3 (during which time it computes the desired result many times!)
-so it would be all but unusable as a normal library routine.
-However, the form of the auto-tuning routine suggests the potential for enormous acceleration:
-if only there were a fast (even approximate) surrogate for the costly $MeasureTime(\cdot)$ function,
-then the argmin could be done in a fraction of a second and the routine could be used normally (Figure~\ref{fig:general_autotune}, bottom).
 

### 2011-12-13 paper/draft1.tex
Minor edits, e.g. in abstract

@@
 \begin{abstract}
-Auto-tuning of parameteric, instrumented code templates has proven to be a
-highly effective technique for optimizing a GPU kernel for a particular
-computation on particular
-hardware \cite{XXXX}.  
+Auto-tuning of parameteric, instrumented code has proven to be a highly
+effective technique for tailoring an optimal compute kernel for a particular
+hardware \cite{XXX}.
 % DDC: I'm still struggling with the above sentence.  Feel free to substitute your own ideas
-However, a drawback of this approach is that exhaustive, or even thorough, auto-tuning requires
-compiling many kernels and calling each one many times, and this process is slow.
-Furthermore, it is desirable to provide users with unified library abstraction
-boundaries for operations such as image
-filtering and matrix multiplication, even those these operations actually correspond to a large set
-of potential problem configurations with a wide variety of memory access
-patterns and computational bottlenecks.  How can we draw on data from previous
-empirical auto-tuning of related problems on related hardware to make a
-just-in-time implementation decision for a novel problem?  This paper presents
-a machine learning approach to auto-tuning, in which features of the
-current hardware platform, the kernel configuration and the problem
-instance are passed to a regression model (boosted regression trees) which
-predicts how much faster this kernel will be than a reference baseline.
-Combinatorial optimization strategies for auto-tuning that would normally
-require evaluating large number of kernel configurations on the real hardware, are made orders of
-magnitude faster by evaluating the surrogate regression model instead.  We
-validate our approach using the filterbank correlation kernel described in
-\citet{pinto+cox:2011gcg}, where we find that 0.1 seconds (XXX) of hill
-climbing on the regression model (which we dub ``{\em predictive} auto-tuning'') can achieve an
-average of 95\% of the speed-up brought by minutes of empirical auto-tuning
-XXX.  Our approach is not specific to filterbank correlation, nor even to GPU
-kernel auto-tuning, and can be applied to almost any templated-code
-optimization problem, spanning a wide variety of problem types, kernel types,
-and platforms.
+However, a drawback of this approach is that exhaustive, or even thorough,
+auto-tuning requires compiling many kernels and calling each one many times,
+and this process is slow.  Furthermore, it is desirable to provide users
+with unified library abstraction boundaries for operations such as image
+filtering and matrix multiplication, even those these operations actually
+correspond to a large set of potential problem configurations with a wide
+variety of memory access patterns and computational bottlenecks.  How can we
+draw on data from previous empirical auto-tuning of related problems on related
+hardware to make a just-in-time implementation decision for a novel problem?
+This paper presents a machine learning approach to auto-tuning, in which
+features of the current hardware platform, the kernel configuration and the
+problem instance are passed to a regression model (boosted regression trees)
+which predicts how much faster this kernel will be than a reference
+baseline.  Combinatorial optimization strategies for auto-tuning that would
+normally require evaluating large number of kernel configurations on the real
+hardware, are made orders of magnitude faster by evaluating the surrogate
+regression model instead.  We validate our approach using the filterbank
+correlation kernel described in \citet{pinto+cox:2011gcg}, where we find that
+0.1 seconds (XXX) of hill climbing on the regression model (which we dub ``{\em
+predictive} auto-tuning'') can achieve an average of 95\% of the
+speed-up brought by minutes of empirical auto-tuning XXX.  Our approach is not
+specific to filterbank correlation, nor even to GPU kernel auto-tuning, and can
+be applied to almost any templated-code optimization problem, spanning a wide
+variety of problem types, kernel types, and platforms.
 
@@
 
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
@@
 
+% =============================================================================
 \subsection{Motivation}
@@
 
+% =============================================================================
 \subsection{Auto-Tuning}
 % Related Work instead ??
+% XXX: a reference to brewer1995high
 
@@
 
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Predictive Auto-Tuning}
@@
 
+% =============================================================================
 \subsection{Learning a Regression Model}
@@
 
+% =============================================================================
 \subsection{Regression Trees}
@@
 
+% =============================================================================
 \subsection{Boosted Regression Trees}
@@
 
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Filterbank Correlation}
@@
 \label{sec:fbcorr}
+
+\iffalse
+XXX: note that image processing libraries include convolutions (e.g. OpenCV ???
+even for GPU!), but only don’t exploit the _bank_ dimension of the
+_filterbank_ and thus can’t exploit parallel hardware as much.
+
+XXX: in addition, when they are optimized, it’s often for pre-defined filters
+(e.g. sobel?), and they are not suitable for encapsulation into libraries
+usable by most real-world applications (bouuuuhhhh opencv !!!).
+\iffalse
 
@@
 
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{GPU Implementation}
@@
 
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Experiment Setup}
@@
 
+% =============================================================================
 \subsection{Search Algorithms}
@@
 
+% =============================================================================
 \subsection{Platforms and Implementation}
@@
 
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Results}
@@
 
+% =============================================================================
 \subsection{Predictive Auto-tuning}
@@
 
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Discussion}
@@
 
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Future Work}

### 2011-12-13 paper/draft1.tex

@@
 
-This work shows that auto-tuning can be accelerated by orders of
-magnitude by using a regression model built offline as a surrogate for actual
-computations on the real hardware.
-The general form of an auto-tuning based library routine is illustrated in Figure~\ref{fig:general_autotune} (top).
-An auto-tuning based routine must operate on three sets of variables:
-\begin{description}
-\item[$\mathcal{A}$:] task description (argument shapes, physical layout)
-\item[$\mathcal{B}$:] implementation description (auto-tuning parameters)
-\item[$\mathcal{C}$:] platform description (capabilities, micro-benchmarks)
-\end{description}
-The hypothetical auto-tuning routine described at the top of Figure~\ref{fig:general_autotune}
-might take many minutes or hours to perform the argmin at step 3 (during which time it computes the desired result many times!)
-so it would not be suitable for a normal library subroutine implementation.
-However, the form of the auto-tuning routine suggests the potential for enormous acceleration:
-if only there were a fast (even approximate) surrogate for the costly $MeasureTime(\cdot)$ function,
-then the argmin could be done in a fraction of a second and the routine could be used normally (Figure~\ref{fig:general_autotune}, bottom).
-
 \begin{figure}[!ht]
@@
 
-% =============================================================================
+This work shows that auto-tuning can be accelerated by orders of
+magnitude by using a regression model built offline as a surrogate for actual
+computations on the real hardware.
+The general form of an auto-tuning based library routine is illustrated in Figure~\ref{fig:general_autotune} (top).
+An auto-tuning based routine must operate on three sets of variables:
+\begin{description}
+\item[$\mathcal{A}$:] task description (argument shapes, physical layout)
+\item[$\mathcal{B}$:] implementation description (auto-tuning parameters)
+\item[$\mathcal{C}$:] platform description (capabilities, micro-benchmarks)
+\end{description}
+The hypothetical auto-tuning routine described at the top of Figure~\ref{fig:general_autotune}
+might take many minutes or hours to perform the argmin at step 3 (during which time it computes the desired result many times!)
+so it would not be suitable for a normal library subroutine implementation.
+However, the form of the auto-tuning routine suggests the potential for enormous acceleration:
+if only there were a fast (even approximate) surrogate for the costly $MeasureTime(\cdot)$ function,
+then the argmin could be done in a fraction of a second and the routine could be used normally (Figure~\ref{fig:general_autotune}, bottom).
+
 \subsection{Learning a Regression Model}
@@
 A regression tree is a piece-wise constant function from one vector space to another,
-that works by recursively subdividing the input space into constant regions~\citep{breiman, textbook?}.
+that works by recursively subdividing the input space into constant regions~\citep{breiman+friedman+olshen+stone:1984, hastie+tibshirani+friedman:2001}.
 They are widely used in statistics and data-mining applications because the fitting algorithm is quick and reliable, and the form of the tree can provide insight into the relevant input variables.
@@
 One disadvantage of the regression tree is that it does not make full use of broad patterns in the data -- each partition formed by the fitting procedure is fit independently in the recursive training procedure,
-so it impossible for the model to extract more than one bit of information from each training partition (XXX unclear).
-This disadvantage is mitigated to a significant extent by the practice of {\em boosting}~\citet{XXX}.
+so it impossible for the model to extract more than one bit of information from each training partition.
+% XXX unclear
+This disadvantage is mitigated to a significant extent by the practice of {\em boosting}.
 
@@
 Boosting is an iterative procedure for constructing an {\em ensemble} of regression trees that is coordinated to fit training examples as accurately as possible.
+\citep{schapire:2001,friedman:2002}
 In a recent empirical study of a range of machine learning regressino problems,
-boosted decision trees were found to be among the best and easiest models to apply~\cite{caruana:2010}.
+boosted decision trees were found to be among the best and easiest models to
+apply~\citep{caruana+niculescu-mizil:2006}.
 On each boosting iteration, a regression tree is fit to the residual error remaining
@@
 and 100 iterations of boosting.
+
+\subsection{Search Algorithms}
+
+Once an accurate regression model has been fit to the data, it remains to be optimized for novel arguments (Fig.~\ref{fig:autotune_general}, bottom, step 4).
+An exhaustive search is the most reliable if it can be afforded, but in our experiments
+(as in general) an exhaustive search is prohibitively expensive.
+In our experiments we compare two strategies: (1) a hand-chosen grid provided by the authors of the kernel used in our experiments \citep{pinto+cox:2011gcg},
+and (2) a generic stochastic hill-climbing search.
+
+% XXX: what is reference?  Why was it chosen?
+% DDC: wasn't the reference implementation hand-tuned?
+% might also want to say something like:
+% Given that parameters were hand-chosen for the reference kernel, no claims are made as to the optimality nor universality of this reference (indeed, different programmers would undoubtedly arrive at different results).
+We provide this kernel configuration simply as a reasonable indicator of typical performance made possible by ad hoc experimentation with parameters.
+
+Additionally, \cite{pinto+cox:2011gcg} advocate a particular grid search over
+what was estimated to be the most relevant part of the configuration space. This grid iterates over all
+combinations of XXX, but leaves the value of XXX at. In our experiments, we call this algorithm the {\em grid} search procedure.
+The grid included $72$ points in addition to the reference implementation, for a total of $73$ points.
+XXX: isn't 73 prime?? explain this so-called ``grid''.
+
+In this work, we also consider a third search algorithm based on a local search
+heuristic.  This {\em hill-climing} (HC) algorithm starts from the reference
+implementation and resamples each of the parameters
+of the current best implementation randomly with
+probability $0.25$ (keeping the current best setting with probability $0.75$).
+On each hill-climbing iteration, if the speed of the newly sampled point is greater than the previous point, then it becomes the current point.
+We show results for search variants HC25, HC50, and HC75, which correspond do hill-climbin for 25, 50, and 75 iterations respectively.
 

### 2011-12-13 paper/draft1.tex
rearrange sec 2

@@
 \begin{abstract}
-Auto-tuning of parameteric, instrumented code has proven to be a highly
-effective technique for tailoring an optimal compute kernel for a particular
-hardware \cite{XXX}.
+Auto-tuning of parameteric, instrumented code templates has proven to be a
+highly effective technique for optimizing a GPU kernel for a particular
+computation on particular
+hardware \cite{XXXX}.  
 % DDC: I'm still struggling with the above sentence.  Feel free to substitute your own ideas
-However, a drawback of this approach is that exhaustive, or even thorough,
-auto-tuning requires compiling many kernels and calling each one many times,
-and this process is slow.  Furthermore, it is desirable to provide users
-with unified library abstraction boundaries for operations such as image
-filtering and matrix multiplication, even those these operations actually
-correspond to a large set of potential problem configurations with a wide
-variety of memory access patterns and computational bottlenecks.  How can we
-draw on data from previous empirical auto-tuning of related problems on related
-hardware to make a just-in-time implementation decision for a novel problem?
-This paper presents a machine learning approach to auto-tuning, in which
-features of the current hardware platform, the kernel configuration and the
-problem instance are passed to a regression model (boosted regression trees)
-which predicts how much faster this kernel will be than a reference
-baseline.  Combinatorial optimization strategies for auto-tuning that would
-normally require evaluating large number of kernel configurations on the real
-hardware, are made orders of magnitude faster by evaluating the surrogate
-regression model instead.  We validate our approach using the filterbank
-correlation kernel described in \citet{pinto+cox:2011gcg}, where we find that
-0.1 seconds (XXX) of hill climbing on the regression model (which we dub ``{\em
-predictive} auto-tuning'') can achieve an average of 95\% of the
-speed-up brought by minutes of empirical auto-tuning XXX.  Our approach is not
-specific to filterbank correlation, nor even to GPU kernel auto-tuning, and can
-be applied to almost any templated-code optimization problem, spanning a wide
-variety of problem types, kernel types, and platforms.
+However, a drawback of this approach is that exhaustive, or even thorough, auto-tuning requires
+compiling many kernels and calling each one many times, and this process is slow.
+Furthermore, it is desirable to provide users with unified library abstraction
+boundaries for operations such as image
+filtering and matrix multiplication, even those these operations actually correspond to a large set
+of potential problem configurations with a wide variety of memory access
+patterns and computational bottlenecks.  How can we draw on data from previous
+empirical auto-tuning of related problems on related hardware to make a
+just-in-time implementation decision for a novel problem?  This paper presents
+a machine learning approach to auto-tuning, in which features of the
+current hardware platform, the kernel configuration and the problem
+instance are passed to a regression model (boosted regression trees) which
+predicts how much faster this kernel will be than a reference baseline.
+Combinatorial optimization strategies for auto-tuning that would normally
+require evaluating large number of kernel configurations on the real hardware, are made orders of
+magnitude faster by evaluating the surrogate regression model instead.  We
+validate our approach using the filterbank correlation kernel described in
+\citet{pinto+cox:2011gcg}, where we find that 0.1 seconds (XXX) of hill
+climbing on the regression model (which we dub ``{\em predictive} auto-tuning'') can achieve an
+average of 95\% of the speed-up brought by minutes of empirical auto-tuning
+XXX.  Our approach is not specific to filterbank correlation, nor even to GPU
+kernel auto-tuning, and can be applied to almost any templated-code
+optimization problem, spanning a wide variety of problem types, kernel types,
+and platforms.
 
@@
 
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
@@
 
-% =============================================================================
 \subsection{Motivation}
@@
 
-% =============================================================================
 \subsection{Auto-Tuning}
 % Related Work instead ??
-% XXX: a reference to brewer1995high
 
@@
 
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Predictive Auto-Tuning}
@@
 
-% =============================================================================
 \subsection{Regression Trees}
@@
 
-% =============================================================================
 \subsection{Boosted Regression Trees}
@@
 
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Filterbank Correlation}
@@
 \label{sec:fbcorr}
-
-\iffalse
-XXX: note that image processing libraries include convolutions (e.g. OpenCV ???
-even for GPU!), but only don’t exploit the _bank_ dimension of the
-_filterbank_ and thus can’t exploit parallel hardware as much.
-
-XXX: in addition, when they are optimized, it’s often for pre-defined filters
-(e.g. sobel?), and they are not suitable for encapsulation into libraries
-usable by most real-world applications (bouuuuhhhh opencv !!!).
-\iffalse
 
@@
 
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{GPU Implementation}
@@
 
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Experiment Setup}
@@
 
-% =============================================================================
-\subsection{Search Algorithms}
-
-The core of any online auto-tuning algorithm (Fig.~\ref{fig:autotune_general}) is the search procedure used to explore the space of implementations.
-Our implementation space includes 16000 (XXX) points, too many to search exhaustively for every input configuration.
-\cite{cox+pinto:2011gcg} recommends a particular hyper-parameter implementation
-based on empirical auto-tuning on several platforms, that achieves good
-performance on a variety of GPUs from older-generation cards such as the
-8600GT all the way to
-current-generation flagship cards such as the GTX 580 and C2070.
-We call this configuration the {\em reference} implementation.
-% XXX: what is reference?  Why was it chosen?
-% DDC: wasn't the reference implementation hand-tuned?
-% might also want to say something like:
-% Given that parameters were hand-chosen for the reference kernel, no claims are made as to the optimality nor universality of this reference (indeed, different programmers would undoubtedly arrive at different results).
-We provide this kernel configuration simply as a reasonable indicator of typical performance made possible by ad hoc experimentation with parameters.
-
-Additionally, \cite{cox+pinto:2011gcg} advocate a particular grid search over
-what was estimated to be the most relevant part of the configuration space. This grid iterates over all
-combinations of XXX, but leaves the value of XXX at. In our experiments, we call this algorithm the {\em grid} search procedure.
-The grid included $72$ points in addition to the reference implementation, for a total of $73$ points.
-XXX: isn't 73 prime?? explain this so-called ``grid''.
-
-In this work, we also consider a third search algorithm based on a local search
-heuristic.  This {\em hill-climing} (HC) algorithm starts from the reference
-implementation and resamples each of the parameters
-of the current best implementation randomly with
-probability $0.25$ (keeping the current best setting with probability $0.75$).
-On each hill-climbing iteration, if the speed of the newly sampled point is greater than the previous point, then it becomes the current point.
-We show results for search variants HC25, HC50, and HC75, which correspond do hill-climbin for 25, 50, and 75 iterations respectively.
-
-% =============================================================================
 \subsection{Platforms and Implementation}
@@
 
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Results}
@@
 
-% =============================================================================
 \subsection{Predictive Auto-tuning}
@@
 
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Discussion}
@@
 
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Future Work}

### 2011-12-13 paper/draft1.tex
Last brain dump edits... ready to go to writing mode... 4096 childrennnn ! ! !

--- 2011-12-13 paper/draft1.tex
@@
 
+XXX: deliver significant performance increases at very low run-time cost (less
+than a second if the input domain has not been seen before, zero if it
+has, since it will hit the cache) compared with the reference
+implementation
+
+XXX always delivered good performance: optimality without loss of productivity?
+
+XXX such machine learning derived model-based empirical auto-tuner hold great
+promise for dev productivity, portability, performance on future multi-core hw
+with different architectural features
+
+XXX frameworks for predicting performance exist \cite{XXX Hwu An Adaptive
+Performance Modeling Tool for GPU Architectures}
+but for non-GPU architectures
+
 auto-tuning required hardward expertise *and* application-domain knowledge, we

### 2011-12-13 paper/draft1.tex

@@
 (as in general) an exhaustive search is prohibitively expensive.
-In our experiments we compare two strategies: (1) a hand-chosen grid provided by the authors of the kernel used in our experiments \citep{pinto+cox:2011gcg},
-and (2) a generic stochastic hill-climbing search.
-
-% XXX: what is reference?  Why was it chosen?
-% DDC: wasn't the reference implementation hand-tuned?
-% might also want to say something like:
-% Given that parameters were hand-chosen for the reference kernel, no claims are made as to the optimality nor universality of this reference (indeed, different programmers would undoubtedly arrive at different results).
-We provide this kernel configuration simply as a reasonable indicator of typical performance made possible by ad hoc experimentation with parameters.
-
-Additionally, \cite{pinto+cox:2011gcg} advocate a particular grid search over
-what was estimated to be the most relevant part of the configuration space. This grid iterates over all
-combinations of XXX, but leaves the value of XXX at. In our experiments, we call this algorithm the {\em grid} search procedure.
-The grid included $72$ points in addition to the reference implementation, for a total of $73$ points.
-XXX: isn't 73 prime?? explain this so-called ``grid''.
-
-In this work, we also consider a third search algorithm based on a local search
-heuristic.  This {\em hill-climing} (HC) algorithm starts from the reference
+In our experiments we compare two strategies:
+(1)
+a generic stochastic hill-climbing search,
+and (2)
+a hand-chosen grid provided by the authors of the kernel used in our experiments \citep{pinto+cox:2011gcg}.
+The {\em hill-climing} (HC) search algorithm starts from the reference
 implementation and resamples each of the parameters
@@
 We show results for search variants HC25, HC50, and HC75, which correspond do hill-climbin for 25, 50, and 75 iterations respectively.
+The {\em grid} algorithm is specific to the kernel used in our case study,
+the details of the grid are provided with our experimental results in Section~\ref{sec:refgrid}
 
@@
 
+\subsection{Reference and Grid}
+\label{sec:refgrid}
+The reference implementation XXX
+
+% XXX: what is reference?  Why was it chosen?
+% DDC: wasn't the reference implementation hand-tuned?
+% might also want to say something like:
+% Given that parameters were hand-chosen for the reference kernel, no claims are made as to the optimality nor universality of this reference (indeed, different programmers would undoubtedly arrive at different results).
+We provide this kernel configuration simply as a reasonable indicator of typical performance made possible by ad hoc experimentation with parameters.
+
+Additionally, \cite{pinto+cox:2011gcg} advocate a particular grid search over
+what was estimated to be the most relevant part of the configuration space. This grid iterates over all
+combinations of XXX, but leaves the value of XXX at. In our experiments, we call this algorithm the {\em grid} search procedure.
+The grid included $72$ points in addition to the reference implementation, for a total of $73$ points.
+XXX: isn't 73 prime?? explain this so-called ``grid''.
+
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Experiment Setup}

### 2011-12-13 paper/draft1.tex
new reference and grid sec

--- 2011-12-13 paper/draft1.tex
@@
 % Related Work instead ??
+% XXX: a reference to brewer1995high
 
@@
 especially ``block\_w'', ``block\_h'', ``n\_output\_4s'' and ``n\_filter\_rows.''
+
+\subsection{Reference and Grid}
+\label{sec:refgrid}
+The reference implementation XXX
+
+% XXX: what is reference?  Why was it chosen?
+% DDC: wasn't the reference implementation hand-tuned?
+% might also want to say something like:
+% Given that parameters were hand-chosen for the reference kernel, no claims are made as to the optimality nor universality of this reference (indeed, different programmers would undoubtedly arrive at different results).
+We provide this kernel configuration simply as a reasonable indicator of typical performance made possible by ad hoc experimentation with parameters.
+
+Additionally, \cite{pinto+cox:2011gcg} advocate a particular grid search over
+what was estimated to be the most relevant part of the configuration space. This grid iterates over all
+combinations of XXX, but leaves the value of XXX at. In our experiments, we call this algorithm the {\em grid} search procedure.
+The grid included $72$ points in addition to the reference implementation, for a total of $73$ points.
+XXX: isn't 73 prime?? explain this so-called ``grid''.
 

### 2011-12-13 paper/draft1.tex

--- 2011-12-13 paper/draft1.tex
@@
 
-XXX: deliver significant performance increases at very low run-time cost (less
-than a second if the input domain has not been seen before, zero if it
-has, since it will hit the cache) compared with the reference
-implementation
-
-XXX always delivered good performance: optimality without loss of productivity?
-
-XXX such machine learning derived model-based empirical auto-tuner hold great
-promise for dev productivity, portability, performance on future multi-core hw
-with different architectural features
-
-XXX frameworks for predicting performance exist \cite{XXX Hwu An Adaptive
-Performance Modeling Tool for GPU Architectures}
-but for non-GPU architectures
-
 auto-tuning required hardward expertise *and* application-domain knowledge, we

### 2011-12-14 fig.py
cleaning up figures

@@ -6,15 +6,48 @@ Scripts for making figures
 import cPickle
 import logging
 import sys
-import time
+from os import environ
+# -- 
+import matplotlib.pyplot as plt
+plt.rcParams["xtick.direction"] = "out"
+plt.rcParams["ytick.direction"] = "out"
+#plt.rcParams['ps.useafm'] = True 
+#plt.rcParams['ps.fonttype'] = 42
+#plt.rcParams['ps.usedistiller'] = 'xpdf'
+plt.rcParams['ps.useafm'] = True 
+plt.rcParams['ps.fonttype'] = 42
+plt.rcParams['ps.usedistiller'] = 'xpdf'
+
+PLOT = int(environ.get("PLOT", 0))
 
 import numpy as np
 
-import wisdom
 import matplotlib.pyplot as plt
 
 logger = logging.getLogger(__name__)
 logging.basicConfig(level=logging.DEBUG)
+
+
+# *try* to convince mpl to do something non-crappy
+def adjust_axis(ax):
+
+    for loc, spine in ax.spines.iteritems():
+        if loc in ['left', 'bottom']:
+            spine.set_position(('outward', 10))
+            spine.set_linewidth(0.5)
+        if loc in ['right', 'top']:
+            spine.set_color('none')
+
+    ax.yaxis.set_ticks_position('left')
+    ax.xaxis.set_ticks_position('bottom')
+
+    fontsize = 10
+    for tick in ax.xaxis.get_major_ticks() + ax.yaxis.get_major_ticks():
+        tick.label1.set_fontsize(fontsize)
+        tick.label1.set_family('sans-serif')
+    for line in ax.xaxis.get_ticklines() + ax.yaxis.get_ticklines():
+        line.set_markeredgewidth(0.5)
+
 
 def pretty_dname(devicename):
     if devicename == '295': return 'GTX 295'
@@ -79,7 +112,7 @@ def main_allstars_mixup():
         'Reference',
         'Mismatched Auto-tuned'])
     plt.ylabel("Avg. GFLOPS/s")
-    if 0:
+    if PLOT:
         plt.show()
     else:
         plt.savefig('allstars_mixup_295.pdf')
@@ -124,7 +157,7 @@ def main_genX():
         yticks = [0.5, 1, 1.5, 2, 2.5]
         plt.yticks(yticks, [('%.1fx'%i) for i in yticks])
 
-    if 0:
+    if PLOT:
         plt.show()
     else:
[…]

### 2011-12-14 paper/draft1.tex
ref and grid

@@
 \nonumber
-% XXX: incrementing requires reading and writing $z$ right? so actually the 4
-% above should be an 8? (see related ticket on github for potential
-% optimization about this)
 \label{eq:bytesP}
@@
 
-XXX
-Arithmetic density with passes.
-
 Critically:
-The arithmetic intensity, shared storage, and register requirements
+what makes this kernel interesting as a case study is that the
+arithmetic intensity, shared storage, and register requirements
 of this kernel change significantly and in a complicated platform-dependent way
@@
 \label{sec:refgrid}
-The reference implementation XXX
-
-% XXX: what is reference?  Why was it chosen?
-% DDC: wasn't the reference implementation hand-tuned?
-% might also want to say something like:
-% Given that parameters were hand-chosen for the reference kernel, no claims are made as to the optimality nor universality of this reference (indeed, different programmers would undoubtedly arrive at different results).
-We provide this kernel configuration simply as a reasonable indicator of typical performance made possible by ad hoc experimentation with parameters.
+
+\cite{pinto+cox:2011gcg} recommends as a reference implementation: block\_w 8, 
+block\_h = 8, n\_filter\_rows 1, n\_output\_4s all, spill $False$, imul\_fast $True$,
+pad\_shared $True$, 
+text1dfetch $True$,
+maxrregcount $\infty$,
+and fast\_math $False$.
+This reference implementation was chosen manually based on good performance across a range of platforms from older-generation
+cards such as the
+8600GT all the way to
+current-generation flagship cards such as the GTX 580 and C2070.
+Given that parameters were hand-chosen for the reference kernel, no claims are
+made as to the optimality nor universality of this reference (indeed, different
+programmers would undoubtedly arrive at different results).  We use this
+kernel configuration as a reasonable indicator of typical performance
+made possible by ad hoc experimentation with parameters.
 
 Additionally, \cite{pinto+cox:2011gcg} advocate a particular grid search over
-what was estimated to be the most relevant part of the configuration space. This grid iterates over all
-combinations of XXX, but leaves the value of XXX at. In our experiments, we call this algorithm the {\em grid} search procedure.
+what was estimated to be the most relevant part of the configuration space.
+This grid iterates over all
+combinations of n\_filter\_rows, n\_output\_4s, spill\_l, pad\_shared\_l for three different block\_h, block\_w choices: (16, 8), (16, 16), and (32, 8).
+In our experiments, we call this algorithm the {\em grid} search procedure.
 The grid included $72$ points in addition to the reference implementation, for a total of $73$ points.
-XXX: isn't 73 prime?? explain this so-called ``grid''.
 
@@
 % =============================================================================
-\subsection{Search Algorithms}
-
-The core of any online auto-tuning algorithm (Fig.~\ref{fig:autotune_general}) is the search procedure used to explore the space of implementations.
-Our implementation space includes 16000 (XXX) points, too many to search exhaustively for every input configuration.
-\cite{cox+pinto:2011gcg} recommends a particular hyper-parameter implementation
-based on empirical auto-tuning on several platforms, that achieves good
-performance on a variety of GPUs from older-generation cards such as the
-8600GT all the way to
-current-generation flagship cards such as the GTX 580 and C2070.
-We call this configuration the {\em reference} implementation.
-% XXX: what is reference?  Why was it chosen?
-% DDC: wasn't the reference implementation hand-tuned?
-% might also want to say something like:
-% Given that parameters were hand-chosen for the reference kernel, no claims are made as to the optimality nor universality of this reference (indeed, different programmers would undoubtedly arrive at different results).
-We provide this kernel configuration simply as a reasonable indicator of typical performance made possible by ad hoc experimentation with parameters.
-
 Additionally, \cite{cox+pinto:2011gcg} advocate a particular grid search over

### 2011-12-14 paper/draft1.tex
Cleaning up code for figures

--- 2011-12-13 paper/draft1.tex
@@
 \begin{abstract}
+
 Auto-tuning of parameteric, instrumented code has proven to be a highly
@@
 
-One strategy for addressing these challenges is to use one of a variety of automatic
-methods known collectively as ``auto-tuning.'' Two major auto-tuning approaches have emerged
-in the extensive literature covering the subject (see surveys in
+One strategy for addressing these challenges is to use one of a variety of
+automatic methods known collectively as ``auto-tuning.'' Two major auto-tuning
+approaches have emerged in the extensive literature covering the subject (see
+surveys in
 \citep{vuduc2001statistical, demmel2005self, vuduc2005oski, williams2008auto,
@@
 We call this approach {\em predictive} auto-tuning to contrast it with the standard measurement-based approach which we will call {\em empirical} auto-tuning.
-We show that the kernel from the GCG chapter can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.
+We show that the kernel from \citet{pinto+cox:2011gcg} can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.
 %XXX: this has no flow
@@
 For a particular problem (filterbank correlation) we
-look at a parameterized kernel implementation (from GCG)
+look at a parameterized kernel implementation (from \citet{pinto+cox:2011gcg})
 and a few hardware platforms, and compare
@@
 XXX: What to all the parameters mean?
-XXX: Point to github / GCG for full code listing.
 
@@
 change soon)
+
+XXX: deliver significant performance increases at very low run-time cost (less
+than a second if the input domain has not been seen before, zero if it
+has, since it will hit the cache) compared with the reference
+implementation
+
+XXX always delivered good performance: optimality without loss of productivity?
+
+XXX such machine learning derived model-based empirical auto-tuner hold great
+promise for dev productivity, portability, performance on future multi-core hw
+with different architectural features
+
+XXX frameworks for predicting performance exist \cite{XXX Hwu An Adaptive
+Performance Modeling Tool for GPU Architectures}
+but for non-GPU architectures
 
@@
 \includegraphics[scale=.42]{speedup_480.pdf}
-\caption{Speedup of various auto-tuning strategies over the reference
-implementation.
-Each strategy was evaluated the same set of 83 problem configurations, which
-was disjoint from the problem configurations used to build the model for
-model-based hill-climbing.
-The vertical dashed lines are positioned at the geometric mean speedup (XXX?)
-of each strategy. The grid and hill-climbing approaches
-tested 73 and 75 configurations respectively, and took 45?? XXX
-seconds on average.
-The model-based approach tested 0 configuration evaluations, and took 0.05
+\caption{
+Speedup of various auto-tuning strategies over the reference
+implementation.  Each strategy was evaluated the same set of 83 problem
+configurations, which was disjoint from the problem configurations used
+to build the model for model-based hill-climbing.  The vertical dashed
+lines are positioned at the geometric mean speedup (XXX?) of each
+strategy. The grid and hill-climbing approaches tested 73 and 75
+configurations respectively, and took 45?? XXX seconds on average.  The
+model-based approach tested 0 configuration evaluations, and took 0.1
 seconds on average, even with a naive Python implementation.

### 2011-12-14 x_plot_ntrain.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+python fig.py ntrain munctional0 580

### 2011-12-14 paper/draft1.tex

--- 2011-12-14 paper/draft1.tex
@@
 \citep{schapire:2001,friedman:2002}
-In a recent empirical study of a range of machine learning regressino problems,
+In a recent empirical study of a range of machine learning regression problems,
 boosted decision trees were found to be among the best and easiest models to
@@
 usable by most real-world applications (bouuuuhhhh opencv !!!).
-\iffalse
+\fi
 
@@
 
+\subsection{Software Stack}
+
+This kernel was implemented in the meta-programming style advocated
+in \citet{pinto+cox:2011gcg} in Python using
+Cheetah for string processing~\citep{cheetah}
+and PyCUDA~\citep{klochner+etal:2009} for dynamic kernel
+compilation and interfacing with CUDA.
+
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 filter sizes as well as variations due
-to strided memory layouts. In such a setting would be useful to characterize
+to strided memory layouts. In such a setting it would be useful to characterize
 the arguments with
@@
 
-% =============================================================================
-Additionally, \cite{cox+pinto:2011gcg} advocate a particular grid search over
-what was estimated to be the most relevant part of the configuration space. This grid iterates over all
-combinations of XXX, but leaves the value of XXX at. In our experiments, we call this algorithm the {\em grid} search procedure.
-The grid included $72$ points in addition to the reference implementation, for a total of $73$ points.
-XXX: isn't 73 prime?? explain this so-called ``grid''.
-
-In this work, we also consider a third search algorithm based on a local search
-heuristic.  This {\em hill-climing} (HC) algorithm starts from the reference
-implementation and resamples each of the parameters
-of the current best implementation randomly with
-probability $0.25$ (keeping the current best setting with probability $0.75$).
-On each hill-climbing iteration, if the speed of the newly sampled point is greater than the previous point, then it becomes the current point.
-We show results for search variants HC25, HC50, and HC75, which correspond do hill-climbin for 25, 50, and 75 iterations respectively.
-
-% =============================================================================
-\subsection{Platforms and Implementation}
-
-This kernel was implemented in the meta-programming style advocated
-in \citet{pinto+cox:2011gcg} in Python using
-Cheetah for string processing~\citep{cheetah}
-and PyCUDA~\citep{klochner+etal:2009} for dynamic kernel
-compilation and interfacing with CUDA.
-
-We evaluated our approach on several GPU devices:
-XXX.
-
-XXX Table showing theoretical peak FLOPS, theoretical peak bandwidth, core count, shared memory size, register size, etc.
-
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Figure~\ref{fig:speedup} shows the effectiveness of empirical auto-tuning is in this setting.  Averaging across the range of problem configurations in our study,
-implementations discovered by empirical auto-tuning are on average about 50\% (XXX check) faster than the reference implementation.
-XXX: recalculate with simpler mean speedup formula.
-On the GTX 295 speedups over the reference implementation ranged from XXX to XXX with mean XXX.
-On the GTX 480 speedups over the reference implementation ranged from XXX to XXX with mean XXX.
-XXX: Consider table for these numbers.
-XXX: Put in some timings of mixed \& matched all-stars to show that indeed the reference was a good general baseline, and in fact the auto-tuned models are special-purpose.
+Figure~\ref{fig:allstars} shows the effectiveness of empirical auto-tuning is in this setting.
+Taking the GTX 295 as an example,
+and averaging across the range of problem configurations in our study,
+we find that empirically auto-tuned implementations are on average about
+50\% faster than the reference implementation.
+The reference in turn, is about 50\% faster than implementations that were empirically auto-tuned for a [randomly chosen] different argument configuration.
+This shows that it is genereally not enough to auto-tune for particular argument configurations;
+instead it is important to choose the right kernel for the job for each unique argument configuration (input-dependent auto-tuning).
 \begin{figure}
@@
 }
-\label{fig:speedup}
+\label{fig:allstars}
 \end{figure}

### 2011-12-14 paper/draft1.tex
into results

@@
 \begin{abstract}
-
 Auto-tuning of parameteric, instrumented code has proven to be a highly
@@
 
-One strategy for addressing these challenges is to use one of a variety of
-automatic methods known collectively as ``auto-tuning.'' Two major auto-tuning
-approaches have emerged in the extensive literature covering the subject (see
-surveys in
+One strategy for addressing these challenges is to use one of a variety of automatic
+methods known collectively as ``auto-tuning.'' Two major auto-tuning approaches have emerged
+in the extensive literature covering the subject (see surveys in
 \citep{vuduc2001statistical, demmel2005self, vuduc2005oski, williams2008auto,
@@
 We call this approach {\em predictive} auto-tuning to contrast it with the standard measurement-based approach which we will call {\em empirical} auto-tuning.
-We show that the kernel from \citet{pinto+cox:2011gcg} can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.
+We show that the kernel from the GCG chapter can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.
 %XXX: this has no flow
@@
 For a particular problem (filterbank correlation) we
-look at a parameterized kernel implementation (from \citet{pinto+cox:2011gcg})
+look at a parameterized kernel implementation (from GCG)
 and a few hardware platforms, and compare
@@
 XXX: What to all the parameters mean?
+XXX: Point to github / GCG for full code listing.
 
@@
 change soon)
-
-XXX: deliver significant performance increases at very low run-time cost (less
-than a second if the input domain has not been seen before, zero if it
-has, since it will hit the cache) compared with the reference
-implementation
-
-XXX always delivered good performance: optimality without loss of productivity?
-
-XXX such machine learning derived model-based empirical auto-tuner hold great
-promise for dev productivity, portability, performance on future multi-core hw
-with different architectural features
-
-XXX frameworks for predicting performance exist \cite{XXX Hwu An Adaptive
-Performance Modeling Tool for GPU Architectures}
-but for non-GPU architectures
 
@@
 \includegraphics[scale=.42]{speedup_480.pdf}
-\caption{
-Speedup of various auto-tuning strategies over the reference
-implementation.  Each strategy was evaluated the same set of 83 problem
-configurations, which was disjoint from the problem configurations used
-to build the model for model-based hill-climbing.  The vertical dashed
-lines are positioned at the geometric mean speedup (XXX?) of each
-strategy. The grid and hill-climbing approaches tested 73 and 75
-configurations respectively, and took 45?? XXX seconds on average.  The
-model-based approach tested 0 configuration evaluations, and took 0.1
+\caption{Speedup of various auto-tuning strategies over the reference
+implementation.
+Each strategy was evaluated the same set of 83 problem configurations, which
+was disjoint from the problem configurations used to build the model for
+model-based hill-climbing.
+The vertical dashed lines are positioned at the geometric mean speedup (XXX?)
+of each strategy. The grid and hill-climbing approaches
+tested 73 and 75 configurations respectively, and took 45?? XXX
+seconds on average.
+The model-based approach tested 0 configuration evaluations, and took 0.05
 seconds on average, even with a naive Python implementation.

### 2011-12-14 paper/draft1.tex
results

@@
 
+Comparing the {\em grid} to {\em HC25}, {\em HC50} and {\em HC75} we found very little difference in performance.
+The HC25 was slightly poorer, but the grid, HC50, and HC75 algorithms delivered similar average results.
+None of the algorithms was strictly better than the others.
+In our predictive auto-tuning experiments we used HC75.
+
+% formation of training set
+
+Figure~\ref{XXX} shows how accurate predictive auto-tuning is compared with empirical auto-tuning.
+The training set $(\mathcal{X}, \mathcal{Y})$ for the regression model comprised
+all of the $<(a,b,c), y>$ pairs observed during grid search and hill-climbing search.
+So for each training argument configuration $a,c$ there were 148 different values of $b$ and thus 148 training points.
+Figure~\ref{XXX} shows that as more argument configurations are used for training,
+the performance of predictive auto-tuning on test configurations $(a', c)$ quickly
+approaches the average performance of empirical auto-tuning.
+We took care to partition the train and test sets so that there were no overlapping configurations.
+The key difference between the predictive and empirical auto-tuning,
+is that predictive auto-tuning
+typically took about 0.1 seconds per test example, whereas empirical auto-tuning took about 1-3 minutes.
+
+% backtracking
+In some cases, predictive auto-tuning yields an invalid implementation.
+We dealt with this scenario by backtracking through the various best-estimates discovered during the hill-climbing search.
+Some invalid kernels can only be discovered after compiling code and attempting to run the compiled code, so in these cases predictive auto-tuning took up to 3 seconds.
+Even in these cases the predictive model is much faster than empirical auto-tuning, because the first kernel that actually runs is a fast one.
+
 Figure~\ref{fig:speedup} also shows that random hill-climbing is at least

### 2011-12-14 paper/draft1.tex
Removing missed conflict frass

--- 2011-12-14 paper/draft1.tex
@@
 
-We're characterizing hardware by a 1-of-N feature vector, simply describing
-which hardware is the current hardware.
-To make better use of auto-tuning data from other platforms, it would be more
-useful to have precise and descriptive features such as: what compute
-capability is present, how many cores are active, what is the bandwidth
-between the various kinds of memory, and how much of various kinds of memory
-is present.  With these features a model-assisted auto-tuning approach
-might be able to make very good guesses on hardware for which no auto-tuning
-has ever been done.
-
-In a complete implementation of Eq.~\ref{eq:z}, there would be parameters
-related to the physical layout (e.g. strides) of $x$, $f$ and $z$ arrays in
-addition to the mathematical parameters of heights and widths and so on,
-so the total number of filterbank correlation computations that a user might
-be able to demand is astronomical.
-A typical coping mechanism would be to cast an arbitrary problem configuration into a more
-standardized form, such as by copying inputs into aligned memory buffers with
-appropriate padding, and then choosing an appropriate blocking strategy for
-the computation. At that point, auto-tuning efforts can be focussed on the
-kernel for each blocked strategy. However, on GPU hardware, the cost of
-aligning the inputs can be relatively large.  In future work we hope to apply
-our model-based technique to auto-tune in full problem configuration space, so
-that we can optimize as much of the computation as possible within the
-predictive auto-tuning framework (XXX name).
-
-Platform Space: XXX
-
-XXX: GCG chapter pg 13 notes that among the grid search, the best parameter
-setting is different for the various platforms.
-
-XXX: so far we only tried 3 cards... so we're using a 1-of-N feature.
-
-%\renewcommand{\bibsection}{\section*{Bibliography}}
-%\setlength{\bibsep}{5pt}
-
-One might expect then, that it would be easy to implement a library providing
-this operation -- this hypothetical library could have quite a simple interface
-following the spirit of FFTW or a BLAS routine:
-\begin{verbatim}
-    gpu_fbcorr(
-        img_shape, filters_shape,
-        img, filters, output)
-\end{verbatim}
-However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
-related stencil operations XXX, it is challenging to provide an implementation
-or even an implementation strategy that provides satisfactory performance
-across the range of inputs (shapes, physical layouts) that occur even in
-typical usage.
-
-Current state-of-the art is to do empirical auto-tuning.
-XXX \vspace{12pt}
-
-Increasingly, machine learning and statistical inference strategies are
-finding application in compiler technology.
-XXX \vspace{12pt}
-
-machine learning
-
-This paper shows that a kernel with several  boosted
-
-There is a natural function from
-problem space x implementation space x platform space to runtime:
-how much wall time elapses on the given platform when solving the given
-problem with the given implementation.
-
-Auto-tuning is a family of empirical techniques for finding the implementation
-that minimizes that runtime function, when the platform and the problem
-configuration are given.
-
-XXX: some math, or possibly a picture with the three boxes that was on Pinto's
-whiteboard.
-
-This paper is about different ways of doing the minimization.
-For a particular problem (filterbank correlation) we
-look at a parameterized kernel implementation (from GCG)
-and a few hardware platforms, and compare
-random search, grid search, and a hill-climbing strategy with a good, but
-generic, reference implementation.
-Then we show that we can model the LSM function with a regression tree
-well enough to usefully perform the argmin of Equation~\ref{eq:lsm_argmin}
-on the model rather than the actual system. Whereas it may take several
-minutes to perform a hill-climbing or grid search in the original system,
-a hill-climbing search in the model requires a small fraction of a second.
-Model-based autotuning makes it possible to simulate auto-tuning quickly and
-accurately enough to be used within a single library call of the computational
-routine.
-
-Best arithmetic density
-
-A bandwidth analysis that takes the blocking structure into account would
-reveal that larger $H$ and $W$ require each block to read a larger input
-region, and store it in a larger shared memory region, which reduces the
-potential for high occupancy.
-
-Filterbank correlation represents a difficult code-optimization problem
-because of the number of interacting constraints: a large value of $P$ reduces
-arithmetic density, but a small value of $P$ requires that each thread do more
-work, and makes global memory access latency harder to hide. (XXX: is there a
-better reason?)
-
-The kernel used to perform
-filterbank correlation was parametrized by 10 parameters, some of which
-were binary and others of which were integer-valued. The full configuration
-space included 12000 kernels.
-XXX: how to make sense of these paramters without code listing?
-
-XXX: What to all the parameters mean?
-XXX: Point to github / GCG for full code listing.
-
-XXX: how many configurations are in the configuration space
-
-XXX: what was the reference kernel, and how was it chosen?
-
-XXX: analysis of READ traffic, WRITE traffic, DATA re-use and CACHEing
-strategies. FLOPS per write etc.
-
-XXX: Why not use FFT: convolution in spectral domain?
-
-XXXX: Talk about memory transfer requirements vs. speed vs. copy...
-
-we believe that our approach is the most practial one to get high perf on
-recent hardware, since there is no consensus on hw features (and it won't
-change soon)
-
-auto-tuning required hardward expertise *and* application-domain knowledge, we
-want to remove these restrictions ultimately we want a ml-based approach to
-code optimization that will where a set of generic but deep code
-transformations, could be templated and injected in any domain-specific
-library. this will reduce the need for specialized expertise across the range
-of targetted computing systems.
+% say what we said
+In this paper, we've demonstrated a boosted regression tree-based auto-tuning method, wherein empirical performance data is used to train a machine learning model of performance for an instrumented GPU kernel.
+In contrast to traditional model-based auto-tuning, where an explicit model of performance is built on the basis of an understanding of hardware inner workings, and empirical auto-tuning, where an exhaustive set of implementation configurations are tried, the present approach generates, from scratch, a model of kernel performance on the basis of timing data from a user-definable number of kernel evaluations.
+This approach allows significant flexibility to navigate tradeoffs between offline and run-time costs, and final auto-tuning performance.
+Importantly, this method treats kernels as black-boxes, allowing the user to auto-tune in the absence of deep knowledge of hardware details (which may even been unknowable, in the case of hardware not available at the time of kernel creation).
+This approach also frees auto-tuning performance from a strong dependency on the accuracy (or inaccuracy) of a pre-defined analytical model.
+
+An important use case for the tools described here is in the development of user-facing numerical libraries.
+Such libraries are a critical component of scientific computing infrastructure, since they abstract away implementation details and make algorithms available to a much wider audience.
+However, the abstraction provided by libraries represents a double-edged sword: one hand, \emph{using} the library is easier, because it presents a unified abstraction of related functionality.
+However, at the same time, any given library routine might represent a wide range of substantially different problem configurations, each with distinct computational issues and bottlenecks.
+Auto-tuning has long provided a solution that finesses these two issues, providing multiple implementations under the hood, for multiple problem settings, and then using heuristics or explicit, hand-crafted models to select the appropriate implementation for a given set of inputs.
+The development of such auto-tuned libraries, while extremely successful, is also very difficult.
+The machine-learning-based techniques described here provide a middle ground, where a library implementor can simply create an instrumented kernel, and allow generic tools to automatically generate appropriate auto-tuned implementations, with small (and controllable) run-time costs.
+
+% DDC: the following text looks like random dumping ground
+
+% We're characterizing hardware by a 1-of-N feature vector, simply describing
+% which hardware is the current hardware.
+% To make better use of auto-tuning data from other platforms, it would be more
+% useful to have precise and descriptive features such as: what compute
+% capability is present, how many cores are active, what is the bandwidth
+% between the various kinds of memory, and how much of various kinds of memory
+% is present.  With these features a model-assisted auto-tuning approach
+% might be able to make very good guesses on hardware for which no auto-tuning
+% has ever been done.
+% 
+% In a complete implementation of Eq.~\ref{eq:z}, there would be parameters
+% related to the physical layout (e.g. strides) of $x$, $f$ and $z$ arrays in
+% addition to the mathematical parameters of heights and widths and so on,
+% so the total number of filterbank correlation computations that a user might
+% be able to demand is astronomical.
+% A typical coping mechanism would be to cast an arbitrary problem configuration into a more
+% standardized form, such as by copying inputs into aligned memory buffers with
+% appropriate padding, and then choosing an appropriate blocking strategy for
+% the computation. At that point, auto-tuning efforts can be focussed on the
+% kernel for each blocked strategy. However, on GPU hardware, the cost of
+% aligning the inputs can be relatively large.  In future work we hope to apply
+% our model-based technique to auto-tune in full problem configuration space, so
+% that we can optimize as much of the computation as possible within the
+% predictive auto-tuning framework (XXX name).
+% 
+% Platform Space: XXX
+% 
+% XXX: GCG chapter pg 13 notes that among the grid search, the best parameter
+% setting is different for the various platforms.
+% 
+% XXX: so far we only tried 3 cards... so we're using a 1-of-N feature.
+% 
+% 
+% %\renewcommand{\bibsection}{\section*{Bibliography}}
+% %\setlength{\bibsep}{5pt}
+% 
+% 
+% 
+% One might expect then, that it would be easy to implement a library providing
+% this operation -- this hypothetical library could have quite a simple interface
+% following the spirit of FFTW or a BLAS routine:
+% \begin{verbatim}
+%     gpu_fbcorr(
+%         img_shape, filters_shape,
+%         img, filters, output)
+% \end{verbatim}
+% However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
+% related stencil operations XXX, it is challenging to provide an implementation
+% or even an implementation strategy that provides satisfactory performance
+% across the range of inputs (shapes, physical layouts) that occur even in
+% typical usage.
+% 
+% Current state-of-the art is to do empirical auto-tuning.
+% XXX \vspace{12pt}
+% 
+% Increasingly, machine learning and statistical inference strategies are
+% finding application in compiler technology.
+% XXX \vspace{12pt}
+% 
+% machine learning
+% 
+% This paper shows that a kernel with several  boosted
+% 
+% 
+% There is a natural function from
+% problem space x implementation space x platform space to runtime:
+% how much wall time elapses on the given platform when solving the given
+% problem with the given implementation.
+% 
+% Auto-tuning is a family of empirical techniques for finding the implementation
+% that minimizes that runtime function, when the platform and the problem
+% configuration are given.
+% 
+% XXX: some math, or possibly a picture with the three boxes that was on Pinto's
+% whiteboard.
+% 
+% This paper is about different ways of doing the minimization.
+% For a particular problem (filterbank correlation) we
+% look at a parameterized kernel implementation (from \citet{pinto+cox:2011gcg})
+% and a few hardware platforms, and compare
+% random search, grid search, and a hill-climbing strategy with a good, but
+% generic, reference implementation.
+% Then we show that we can model the LSM function with a regression tree
+% well enough to usefully perform the argmin of Equation~\ref{eq:lsm_argmin}
+% on the model rather than the actual system. Whereas it may take several
+% minutes to perform a hill-climbing or grid search in the original system,
+% a hill-climbing search in the model requires a small fraction of a second.
+% Model-based autotuning makes it possible to simulate auto-tuning quickly and
+% accurately enough to be used within a single library call of the computational
+% routine.
+% 
+% Best arithmetic density
+% 
+% A bandwidth analysis that takes the blocking structure into account would
+% reveal that larger $H$ and $W$ require each block to read a larger input
+% region, and store it in a larger shared memory region, which reduces the
+% potential for high occupancy.
+% 
+% Filterbank correlation represents a difficult code-optimization problem
+% because of the number of interacting constraints: a large value of $P$ reduces
+% arithmetic density, but a small value of $P$ requires that each thread do more
+% work, and makes global memory access latency harder to hide. (XXX: is there a
+% better reason?)
+% 
+% 
+% The kernel used to perform
+% filterbank correlation was parametrized by 10 parameters, some of which
+% were binary and others of which were integer-valued. The full configuration
+% space included 12000 kernels.
+% XXX: how to make sense of these paramters without code listing?
+% 
+% XXX: What to all the parameters mean?
+% 
+% XXX: how many configurations are in the configuration space
+% 
+% XXX: what was the reference kernel, and how was it chosen?
+% 
+% XXX: analysis of READ traffic, WRITE traffic, DATA re-use and CACHEing
+% strategies. FLOPS per write etc.
+% 
+% XXX: Why not use FFT: convolution in spectral domain?
+% 
+% XXXX: Talk about memory transfer requirements vs. speed vs. copy...
+% 
+% we believe that our approach is the most practial one to get high perf on
+% recent hardware, since there is no consensus on hw features (and it won't
+% change soon)
+% 
+% XXX: deliver significant performance increases at very low run-time cost (less
+% than a second if the input domain has not been seen before, zero if it
+% has, since it will hit the cache) compared with the reference
+% implementation
+% 
+% XXX always delivered good performance: optimality without loss of productivity?
+% 
+% XXX such machine learning derived model-based empirical auto-tuner hold great
+% promise for dev productivity, portability, performance on future multi-core hw
+% with different architectural features
+% 
+% XXX frameworks for predicting performance exist \cite{XXX Hwu An Adaptive
+% Performance Modeling Tool for GPU Architectures}
+% but for non-GPU architectures
+% 
+% auto-tuning required hardward expertise *and* application-domain knowledge, we
+% want to remove these restrictions ultimately we want a ml-based approach to
+% code optimization that will where a set of generic but deep code
+% transformations, could be templated and injected in any domain-specific
+% library. this will reduce the need for specialized expertise across the range
+% of targeted computing systems.
 
@@
 
-more features (this is the hardest, and most manual part, may require
-domain-specific knowledge, but not necessariyly, e.g. microbenchmarking)
-* features derived from micro-benchmarking \citep{wong2010demystifying}
-* features derived from performance limiter analysis \citep{micikevicius2010analysis}
-* more instrumentation
-* more data-layout
-* more threading strategies to lower latency
-(finding a good representation is key, again...)
-
-online distributed auto-tuning with learning, especially w/ HT-based
-experiments \cite{HT} our method is indeed very well suited for parallel
-explorating of multiple configurations for learning (in a cluster, e.g. in HT
-context) this may get us \emph{more} than linear speed ups for better accuracy,
-especially when more features will be present (see above)
-
-interpret the learned model and confront the performance against theoreticial
-(or approximated) models of mem/com/compute, arithmetic intensity, etc. to
-understand performance limiters, the hardware, etc.
+While the present work serves as a basic demonstration of the value of using machine learning models to predict optimal implementation parameters, there are many avenues for taking these ideas further.
+A natural route to extend of our current approach is to include more features as inputs to the predictive model.
+In addition to further instrumentation of the kernel in question, input features could include a much broader range of hardware-related information, from more detailed information about device capabilities to micro-benchmarks  \citep{wong2010demystifying}, and results from performance limiter analyses \citep{micikevicius2010analysis}.
+Such additional features would increase the predictive model's ability to adapt to a wide range of different kinds of hardware, including new devices not available at the time of kernel creation.
+
+Another potentially interesting avenue of research is in interpreting the model learned by predictive auto-tuning.
+While traditional model-based auto-tuning approaches by design assume a given model, and empirical auto-tuning approaches are completely model free, the predictive approach described here \emph{generates} a model from performance data.
+Because this generated model can be interrogated by a variety of means, a significant opportunity exists to learn about the factors that drive the performance of a given kernel.
+These insights can be used to further guide the development and instrumentation of the kernel, potentially yielding even greater gains.
+
+% more features (this is the hardest, and most manual part, may require
+% domain-specific knowledge, but not necessariyly, e.g. microbenchmarking)
+% * features derived from micro-benchmarking \citep{wong2010demystifying}
+% * features derived from performance limiter analysis \citep{micikevicius2010analysis}
+
+% DDC: these aren't terribly general / interesting to a broad audience
+
+% * more instrumentation
+% * more data-layout
+% * more threading strategies to lower latency
+% (finding a good representation is key, again...)
+% 
+% online distributed auto-tuning with learning, especially w/ HT-based
+% experiments \cite{HT} our method is indeed very well suited for parallel
+% explorating of multiple configurations for learning (in a cluster, e.g. in HT
+% context) this may get us \emph{more} than linear speed ups for better accuracy,
+% especially when more features will be present (see above)
+% 
+% interpret the learned model and confront the performance against theoreticial
+% (or approximated) models of mem/com/compute, arithmetic intensity, etc. to
+% understand performance limiters, the hardware, etc.
 

### 2011-12-14 paper/draft1.tex
moved discussion and bad figs -> scratch

--- 2011-12-14 paper/draft1.tex
@@
 
-% =============================================================================
-\subsection{Predictive Auto-tuning}
-
-On the GTX 295, a single regression-tree fit to the log-speed-multiplier
-(LSM) function (XXX: how many problem configurations, how much vs. what
-kind of training data) achieves an average Spearman correlation of 0.9 on
-held-out test data (XXX: what kind of test data).
-
-Now: the big test is that if we use the genetic search on the model instead of
-the original function, do we get good performance?
-
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
-
-We're characterizing hardware by a 1-of-N feature vector, simply describing
-which hardware is the current hardware.
-To make better use of auto-tuning data from other platforms, it would be more
-useful to have precise and descriptive features such as: what compute
-capability is present, how many cores are active, what is the bandwidth
-between the various kinds of memory, and how much of various kinds of memory
-is present.  With these features a model-assisted auto-tuning approach
-might be able to make very good guesses on hardware for which no auto-tuning
-has ever been done.
-
-In a complete implementation of Eq.~\ref{eq:z}, there would be parameters
-related to the physical layout (e.g. strides) of $x$, $f$ and $z$ arrays in
-addition to the mathematical parameters of heights and widths and so on,
-so the total number of filterbank correlation computations that a user might
-be able to demand is astronomical.
-A typical coping mechanism would be to cast an arbitrary problem configuration into a more
-standardized form, such as by copying inputs into aligned memory buffers with
-appropriate padding, and then choosing an appropriate blocking strategy for
-the computation. At that point, auto-tuning efforts can be focussed on the
-kernel for each blocked strategy. However, on GPU hardware, the cost of
-aligning the inputs can be relatively large.  In future work we hope to apply
-our model-based technique to auto-tune in full problem configuration space, so
-that we can optimize as much of the computation as possible within the
-predictive auto-tuning framework (XXX name).
-
-Platform Space: XXX
-
-XXX: GCG chapter pg 13 notes that among the grid search, the best parameter
-setting is different for the various platforms.
-
-XXX: so far we only tried 3 cards... so we're using a 1-of-N feature.
-
-%\renewcommand{\bibsection}{\section*{Bibliography}}
-%\setlength{\bibsep}{5pt}
-
-One might expect then, that it would be easy to implement a library providing
-this operation -- this hypothetical library could have quite a simple interface
-following the spirit of FFTW or a BLAS routine:
-\begin{verbatim}
-    gpu_fbcorr(
-        img_shape, filters_shape,
-        img, filters, output)
-\end{verbatim}
-However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
-related stencil operations XXX, it is challenging to provide an implementation
-or even an implementation strategy that provides satisfactory performance
-across the range of inputs (shapes, physical layouts) that occur even in
-typical usage.
-
-Current state-of-the art is to do empirical auto-tuning.
-XXX \vspace{12pt}
-
-Increasingly, machine learning and statistical inference strategies are
-finding application in compiler technology.
-XXX \vspace{12pt}
-
-machine learning
-
-This paper shows that a kernel with several  boosted
-
-There is a natural function from
-problem space x implementation space x platform space to runtime:
-how much wall time elapses on the given platform when solving the given
-problem with the given implementation.
-
-Auto-tuning is a family of empirical techniques for finding the implementation
-that minimizes that runtime function, when the platform and the problem
-configuration are given.
-
-XXX: some math, or possibly a picture with the three boxes that was on Pinto's
-whiteboard.
-
-This paper is about different ways of doing the minimization.
-For a particular problem (filterbank correlation) we
-look at a parameterized kernel implementation (from GCG)
-and a few hardware platforms, and compare
-random search, grid search, and a hill-climbing strategy with a good, but
-generic, reference implementation.
-Then we show that we can model the LSM function with a regression tree
-well enough to usefully perform the argmin of Equation~\ref{eq:lsm_argmin}
-on the model rather than the actual system. Whereas it may take several
-minutes to perform a hill-climbing or grid search in the original system,
-a hill-climbing search in the model requires a small fraction of a second.
-Model-based autotuning makes it possible to simulate auto-tuning quickly and
-accurately enough to be used within a single library call of the computational
-routine.
-
-Best arithmetic density
-
-A bandwidth analysis that takes the blocking structure into account would
-reveal that larger $H$ and $W$ require each block to read a larger input
-region, and store it in a larger shared memory region, which reduces the
-potential for high occupancy.
-
-Filterbank correlation represents a difficult code-optimization problem
-because of the number of interacting constraints: a large value of $P$ reduces
-arithmetic density, but a small value of $P$ requires that each thread do more
-work, and makes global memory access latency harder to hide. (XXX: is there a
-better reason?)
-
-The kernel used to perform
-filterbank correlation was parametrized by 10 parameters, some of which
-were binary and others of which were integer-valued. The full configuration
-space included 12000 kernels.
-XXX: how to make sense of these paramters without code listing?
-
-XXX: What to all the parameters mean?
-XXX: Point to github / GCG for full code listing.
-
-XXX: how many configurations are in the configuration space
-
-XXX: what was the reference kernel, and how was it chosen?
-
-XXX: analysis of READ traffic, WRITE traffic, DATA re-use and CACHEing
-strategies. FLOPS per write etc.
-
-XXX: Why not use FFT: convolution in spectral domain?
-
-XXXX: Talk about memory transfer requirements vs. speed vs. copy...
-
-we believe that our approach is the most practial one to get high perf on
-recent hardware, since there is no consensus on hw features (and it won't
-change soon)
-
-XXX: deliver significant performance increases at very low run-time cost (less
-than a second if the input domain has not been seen before, zero if it
-has, since it will hit the cache) compared with the reference
-implementation
-
-XXX always delivered good performance: optimality without loss of productivity?
-
-XXX such machine learning derived model-based empirical auto-tuner hold great
-promise for dev productivity, portability, performance on future multi-core hw
-with different architectural features
-
-XXX frameworks for predicting performance exist \cite{XXX Hwu An Adaptive
-Performance Modeling Tool for GPU Architectures}
-but for non-GPU architectures
-
-auto-tuning required hardward expertise *and* application-domain knowledge, we
-want to remove these restrictions ultimately we want a ml-based approach to
-code optimization that will where a set of generic but deep code
-transformations, could be templated and injected in any domain-specific
-library. this will reduce the need for specialized expertise across the range
-of targetted computing systems.
-
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\begin{figure}
-\centering
-\includegraphics[scale=.42]{speedup_295.pdf}
-\includegraphics[scale=.42]{speedup_480.pdf}
-\caption{Speedup of various auto-tuning strategies over the reference
-implementation.
-Each strategy was evaluated the same set of 83 problem configurations, which
-was disjoint from the problem configurations used to build the model for
-model-based hill-climbing.
-The vertical dashed lines are positioned at the geometric mean speedup (XXX?)
-of each strategy. The grid and hill-climbing approaches
-tested 73 and 75 configurations respectively, and took 45?? XXX
-seconds on average.
-The model-based approach tested 0 configuration evaluations, and took 0.05
-seconds on average, even with a naive Python implementation.
-}
-\label{fig:speedup}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.42]{fig_genX_vader_295.pdf}
-\includegraphics[scale=.42]{fig_genX_munctional0_480.pdf}
-\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
-\label{fig:speedup}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.42]{fig_genX_munctional0_580.pdf}
-\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
-\label{fig:speedup}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.42]{fig_genX_munctional0_1060.pdf}
-\includegraphics[scale=.42]{fig_genX_munctional0_2070.pdf}
-\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
-\label{fig:speedup}
-\end{figure}
+\section{Conclusion}
 
@@
 \end{document}
+
+% FOR Extra material see scratch.tex

### 2011-12-14 paper/scratch.tex
moved discussion and bad figs -> scratch

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

We're characterizing hardware by a 1-of-N feature vector, simply describing
which hardware is the current hardware.
To make better use of auto-tuning data from other platforms, it would be more
useful to have precise and descriptive features such as: what compute
capability is present, how many cores are active, what is the bandwidth
between the various kinds of memory, and how much of various kinds of memory
is present.  With these features a model-assisted auto-tuning approach
might be able to make very good guesses on hardware for which no auto-tuning
has ever been done.

In a complete implementation of Eq.~\ref{eq:z}, there would be parameters
related to the physical layout (e.g. strides) of $x$, $f$ and $z$ arrays in
addition to the mathematical parameters of heights and widths and so on,
so the total number of filterbank correlation computations that a user might
be able to demand is astronomical.
A typical coping mechanism would be to cast an arbitrary problem configuration into a more
standardized form, such as by copying inputs into aligned memory buffers with
appropriate padding, and then choosing an appropriate blocking strategy for
the computation. At that point, auto-tuning efforts can be focussed on the
kernel for each blocked strategy. However, on GPU hardware, the cost of
aligning the inputs can be relatively large.  In future work we hope to apply
our model-based technique to auto-tune in full problem configuration space, so
that we can optimize as much of the computation as possible within the
predictive auto-tuning framework (XXX name).

Platform Space: XXX

XXX: GCG chapter pg 13 notes that among the grid search, the best parameter
setting is different for the various platforms.

XXX: so far we only tried 3 cards... so we're using a 1-of-N feature.

%\renewcommand{\bibsection}{\section*{Bibliography}}
%\setlength{\bibsep}{5pt}

One might expect then, that it would be easy to implement a library providing
this operation -- this hypothetical library could have quite a simple interface
following the spirit of FFTW or a BLAS routine:
\begin{verbatim}
    gpu_fbcorr(
        img_shape, filters_shape,
        img, filters, output)
\end{verbatim}
However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
related stencil operations XXX, it is challenging to provide an implementation
or even an implementation strategy that provides satisfactory performance
across the range of inputs (shapes, physical layouts) that occur even in
typical usage.

Current state-of-the art is to do empirical auto-tuning.
XXX \vspace{12pt}

Increasingly, machine learning and statistical inference strategies are
finding application in compiler technology.
XXX \vspace{12pt}

machine learning

This paper shows that a kernel with several  boosted

There is a natural function from
problem space x implementation space x platform space to runtime:
how much wall time elapses on the given platform when solving the given
problem with the given implementation.

Auto-tuning is a family of empirical techniques for finding the implementation
that minimizes that runtime function, when the platform and the problem
configuration are given.

XXX: some math, or possibly a picture with the three boxes that was on Pinto's
whiteboard.

This paper is about different ways of doing the minimization.
For a particular problem (filterbank correlation) we
look at a parameterized kernel implementation (from GCG)
and a few hardware platforms, and compare
random search, grid search, and a hill-climbing strategy with a good, but
generic, reference implementation.
Then we show that we can model the LSM function with a regression tree
well enough to usefully perform the argmin of Equation~\ref{eq:lsm_argmin}
on the model rather than the actual system. Whereas it may take several
minutes to perform a hill-climbing or grid search in the original system,
a hill-climbing search in the model requires a small fraction of a second.
Model-based autotuning makes it possible to simulate auto-tuning quickly and
accurately enough to be used within a single library call of the computational
routine.

Best arithmetic density

A bandwidth analysis that takes the blocking structure into account would
reveal that larger $H$ and $W$ require each block to read a larger input
region, and store it in a larger shared memory region, which reduces the
potential for high occupancy.

Filterbank correlation represents a difficult code-optimization problem
because of the number of interacting constraints: a large value of $P$ reduces
arithmetic density, but a small value of $P$ requires that each thread do more
work, and makes global memory access latency harder to hide. (XXX: is there a
better reason?)

The kernel used to perform
filterbank correlation was parametrized by 10 parameters, some of which
were binary and others of which were integer-valued. The full configuration
space included 12000 kernels.
XXX: how to make sense of these paramters without code listing?

XXX: What to all the parameters mean?
XXX: Point to github / GCG for full code listing.

XXX: how many configurations are in the configuration space

XXX: what was the reference kernel, and how was it chosen?

XXX: analysis of READ traffic, WRITE traffic, DATA re-use and CACHEing
strategies. FLOPS per write etc.

XXX: Why not use FFT: convolution in spectral domain?

XXXX: Talk about memory transfer requirements vs. speed vs. copy...

we believe that our approach is the most practial one to get high perf on
recent hardware, since there is no consensus on hw features (and it won't
change soon)

XXX: deliver significant performance increases at very low run-time cost (less
than a second if the input domain has not been seen before, zero if it
has, since it will hit the cache) compared with the reference
implementation

XXX always delivered good performance: optimality without loss of productivity?

XXX such machine learning derived model-based empirical auto-tuner hold great
promise for dev productivity, portability, performance on future multi-core hw
with different architectural features

\iffalse
XXX frameworks for predicting performance exist \cite{XXX Hwu An Adaptive
Performance Modeling Tool for GPU Architectures}
but for non-GPU architectures
\fi

auto-tuning required hardward expertise *and* application-domain knowledge, we
want to remove these restrictions ultimately we want a ml-based approach to
code optimization that will where a set of generic but deep code
transformations, could be templated and injected in any domain-specific
library. this will reduce the need for specialized expertise across the range
of targetted computing systems.

\begin{figure}
\centering
\includegraphics[scale=.42]{speedup_295.pdf}
\includegraphics[scale=.42]{speedup_480.pdf}
\caption{Speedup of various auto-tuning strategies over the reference
implementation.
Each strategy was evaluated the same set of 83 problem configurations, which
was disjoint from the problem configurations used to build the model for
model-based hill-climbing.
The vertical dashed lines are positioned at the geometric mean speedup (XXX?)
of each strategy. The grid and hill-climbing approaches
tested 73 and 75 configurations respectively, and took 45?? XXX
seconds on average.
The model-based approach tested 0 configuration evaluations, and took 0.05
seconds on average, even with a naive Python implementation.
}
\label{fig:speedup}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=.42]{fig_genX_vader_295.pdf}
\includegraphics[scale=.42]{fig_genX_munctional0_480.pdf}
\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
\label{fig:speedup}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=.42]{fig_genX_munctional0_580.pdf}
\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
\label{fig:speedup}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=.42]{fig_genX_munctional0_1060.pdf}
\includegraphics[scale=.42]{fig_genX_munctional0_2070.pdf}
\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
\label{fig:speedup}
\end{figure}

### 2011-12-14 paper/draft1.tex

@@
 
-more features (this is the hardest, and most manual part, may require
-domain-specific knowledge, but not necessariyly, e.g. microbenchmarking)
-* features derived from micro-benchmarking \citep{wong2010demystifying}
-* features derived from performance limiter analysis \citep{micikevicius2010analysis}
-* more instrumentation
-* more data-layout
-* more threading strategies to lower latency
-(finding a good representation is key, again...)
+While the present work serves as a basic demonstration of the value of using machine learning models to predict optimal implementation parameters, there are many avenues for taking these ideas further.
+A natural route to extend of our current approach is to include more features as inputs to the predictive model.
+In addition to further instrumentation of the kernel in question, input features could include a much broader range of hardware-related information, from more detailed information about device capabilities to micro-benchmarks  \citep{wong2010demystifying}, and results from performance limiter analyses \citep{micikevicius2010analysis}.
+Such additional features would increase the predictive model's ability to adapt to a wide range of different kinds of hardware, including new devices not available at the time of kernel creation.
 
-online distributed auto-tuning with learning, especially w/ HT-based
-experiments \cite{HT} our method is indeed very well suited for parallel
-explorating of multiple configurations for learning (in a cluster, e.g. in HT
-context) this may get us \emph{more} than linear speed ups for better accuracy,
-especially when more features will be present (see above)
+Another potentially interesting avenue of research is in interpreting the model learned by predictive auto-tuning.
+While traditional model-based auto-tuning approaches by design assume a given model, and empirical auto-tuning approaches are completely model free, the predictive approach described here \emph{generates} a model from performance data.
+Because this generated model can be interrogated by a variety of means, a significant opportunity exists to learn about the factors that drive the performance of a given kernel.
+These insights can be used to further guide the development and instrumentation of the kernel, potentially yielding even greater gains.
 
-interpret the learned model and confront the performance against theoreticial
-(or approximated) models of mem/com/compute, arithmetic intensity, etc. to
-understand performance limiters, the hardware, etc.
+% more features (this is the hardest, and most manual part, may require
+% domain-specific knowledge, but not necessariyly, e.g. microbenchmarking)
+% * features derived from micro-benchmarking \citep{wong2010demystifying}
+% * features derived from performance limiter analysis \citep{micikevicius2010analysis}
+
+% DDC: these aren't terribly general / interesting to a broad audience
+
+% * more instrumentation
+% * more data-layout
+% * more threading strategies to lower latency
+% (finding a good representation is key, again...)
+% 
+% online distributed auto-tuning with learning, especially w/ HT-based
+% experiments \cite{HT} our method is indeed very well suited for parallel
+% explorating of multiple configurations for learning (in a cluster, e.g. in HT
+% context) this may get us \emph{more} than linear speed ups for better accuracy,
+% especially when more features will be present (see above)
+% 
+% interpret the learned model and confront the performance against theoreticial
+% (or approximated) models of mem/com/compute, arithmetic intensity, etc. to
+% understand performance limiters, the hardware, etc.
 

### 2011-12-14 paper/scratch.tex

@@
 
-\iffalse
-XXX frameworks for predicting performance exist \cite{XXX Hwu An Adaptive
-Performance Modeling Tool for GPU Architectures}
-but for non-GPU architectures
-\fi
+XXX frameworks for predicting performance exist \cite{XXX}
+% Hwu An Adaptive Performance Modeling Tool for GPU Architectures}
+% but for non-GPU architectures
 
@@
 \includegraphics[scale=.42]{speedup_480.pdf}
-\caption{Speedup of various auto-tuning strategies over the reference
-implementation.
-Each strategy was evaluated the same set of 83 problem configurations, which
-was disjoint from the problem configurations used to build the model for
-model-based hill-climbing.
-The vertical dashed lines are positioned at the geometric mean speedup (XXX?)
-of each strategy. The grid and hill-climbing approaches
-tested 73 and 75 configurations respectively, and took 45?? XXX
-seconds on average.
-The model-based approach tested 0 configuration evaluations, and took 0.05
+\caption{
+Speedup of various auto-tuning strategies over the reference
+implementation.  Each strategy was evaluated the same set of 83 problem
+configurations, which was disjoint from the problem configurations used
+to build the model for model-based hill-climbing.  The vertical dashed
+lines are positioned at the geometric mean speedup (XXX?) of each
+strategy. The grid and hill-climbing approaches tested 73 and 75
+configurations respectively, and took 45?? XXX seconds on average.
+The model-based approach tested 0 configuration evaluations, and took 0.1
 seconds on average, even with a naive Python implementation.

### 2011-12-14 paper/draft1.tex
linewrapping future work

@@
 
-While the present work serves as a basic demonstration of the value of using machine learning models to predict optimal implementation parameters, there are many avenues for taking these ideas further.
-A natural route to extend of our current approach is to include more features as inputs to the predictive model.
-In addition to further instrumentation of the kernel in question, input features could include a much broader range of hardware-related information, from more detailed information about device capabilities to micro-benchmarks  \citep{wong2010demystifying}, and results from performance limiter analyses \citep{micikevicius2010analysis}.
-Such additional features would increase the predictive model's ability to adapt to a wide range of different kinds of hardware, including new devices not available at the time of kernel creation.
+While the present work serves as a basic demonstration of the value of using
+machine learning models to predict optimal implementation parameters, there are
+many avenues for taking these ideas further. A natural route to extend of our
+current approach is to include more features as inputs to the predictive model.
+In addition to further instrumentation of the kernel in question, input
+features could include a much broader range of hardware-related information,
+from more detailed information about device capabilities to micro-benchmarks
+\citep{wong2010demystifying}, and results from performance limiter analyses
+\citep{micikevicius2010analysis}. Such additional features would increase the
+predictive model's ability to adapt to a wide range of different kinds of
+hardware, including new devices not available at the time of kernel creation.
 
-Another potentially interesting avenue of research is in interpreting the model learned by predictive auto-tuning.
-While traditional model-based auto-tuning approaches by design assume a given model, and empirical auto-tuning approaches are completely model free, the predictive approach described here \emph{generates} a model from performance data.
-Because this generated model can be interrogated by a variety of means, a significant opportunity exists to learn about the factors that drive the performance of a given kernel.
-These insights can be used to further guide the development and instrumentation of the kernel, potentially yielding even greater gains.
+Another potentially interesting avenue of research is in interpreting the model
+learned by predictive auto-tuning.  While traditional model-based auto-tuning
+approaches by design assume a given model, and empirical auto-tuning approaches
+are completely model free, the predictive approach described here
+\emph{generates} a model from performance data.  Because this generated model
+can be interrogated by a variety of means, a significant opportunity exists to
+learn about the factors that drive the performance of a given kernel.  These
+insights can be used to further guide the development and instrumentation of
+the kernel, potentially yielding even greater gains.
 

### 2011-12-14 paper/draft1.tex

@@
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Discussion}
+
+% say what we said
+In this paper, we've demonstrated a boosted regression tree-based auto-tuning
+method, wherein empirical performance data is used to train a machine learning
+model of performance for an instrumented GPU kernel.  In contrast to
+traditional model-based auto-tuning, where an explicit model of performance is
+built on the basis of an understanding of hardware inner workings, and
+empirical auto-tuning, where an exhaustive set of implementation configurations
+are tried, the present approach generates, from scratch, a model of kernel
+performance on the basis of timing data from a user-definable number of kernel
+evaluations.  This approach allows significant flexibility to navigate
+tradeoffs between offline and run-time costs, and final auto-tuning
+performance.  Importantly, this method treats kernels as black-boxes, allowing
+the user to auto-tune in the absence of deep knowledge of hardware details
+(which may even been unknowable, in the case of hardware not available at the
+time of kernel creation).  This approach also frees auto-tuning performance
+from a strong dependency on the accuracy (or inaccuracy) of a pre-defined
+analytical model.
+
+An important use case for the tools described here is in the development of
+user-facing numerical libraries.  Such libraries are a critical component of
+scientific computing infrastructure, since they abstract away implementation
+details and make algorithms available to a much wider audience.  However, the
+abstraction provided by libraries represents a double-edged sword: one hand,
+\emph{using} the library is easier, because it presents a unified abstraction
+of related functionality.  However, at the same time, any given library routine
+might represent a wide range of substantially different problem configurations,
+each with distinct computational issues and bottlenecks.  Auto-tuning has long
+provided a solution that finesses these two issues, providing multiple
+implementations under the hood, for multiple problem settings, and then using
+heuristics or explicit, hand-crafted models to select the appropriate
+implementation for a given set of inputs.  The development of such auto-tuned
+libraries, while extremely successful, is also very difficult.  The
+machine-learning-based techniques described here provide a middle ground, where
+a library implementor can simply create an instrumented kernel, and allow
+generic tools to automatically generate appropriate auto-tuned implementations,
+with small (and controllable) run-time costs.
+
+% DDC: the following text looks like random dumping ground
+
+% We're characterizing hardware by a 1-of-N feature vector, simply describing
+% which hardware is the current hardware.
+% To make better use of auto-tuning data from other platforms, it would be more
+% useful to have precise and descriptive features such as: what compute
+% capability is present, how many cores are active, what is the bandwidth
+% between the various kinds of memory, and how much of various kinds of memory
+% is present.  With these features a model-assisted auto-tuning approach
+% might be able to make very good guesses on hardware for which no auto-tuning
+% has ever been done.
+% 
+% In a complete implementation of Eq.~\ref{eq:z}, there would be parameters
+% related to the physical layout (e.g. strides) of $x$, $f$ and $z$ arrays in
+% addition to the mathematical parameters of heights and widths and so on,
+% so the total number of filterbank correlation computations that a user might
+% be able to demand is astronomical.
+% A typical coping mechanism would be to cast an arbitrary problem configuration into a more
+% standardized form, such as by copying inputs into aligned memory buffers with
+% appropriate padding, and then choosing an appropriate blocking strategy for
+% the computation. At that point, auto-tuning efforts can be focussed on the
+% kernel for each blocked strategy. However, on GPU hardware, the cost of
+% aligning the inputs can be relatively large.  In future work we hope to apply
+% our model-based technique to auto-tune in full problem configuration space, so
+% that we can optimize as much of the computation as possible within the
+% predictive auto-tuning framework (XXX name).
+% 
+% Platform Space: XXX
+% 
+% XXX: GCG chapter pg 13 notes that among the grid search, the best parameter
+% setting is different for the various platforms.
+% 
+% XXX: so far we only tried 3 cards... so we're using a 1-of-N feature.
+% 
+% 
+% %\renewcommand{\bibsection}{\section*{Bibliography}}
+% %\setlength{\bibsep}{5pt}
+% 
+% 
+% 
+% One might expect then, that it would be easy to implement a library providing
+% this operation -- this hypothetical library could have quite a simple interface
+% following the spirit of FFTW or a BLAS routine:
+% \begin{verbatim}
+%     gpu_fbcorr(
+%         img_shape, filters_shape,
+%         img, filters, output)
+% \end{verbatim}
+% However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
+% related stencil operations XXX, it is challenging to provide an implementation
+% or even an implementation strategy that provides satisfactory performance
+% across the range of inputs (shapes, physical layouts) that occur even in
+% typical usage.
+% 
+% Current state-of-the art is to do empirical auto-tuning.
+% XXX \vspace{12pt}
+% 
+% Increasingly, machine learning and statistical inference strategies are
+% finding application in compiler technology.
+% XXX \vspace{12pt}
+% 
+% machine learning
+% 
+% This paper shows that a kernel with several  boosted
+% 
+% 
+% There is a natural function from
+% problem space x implementation space x platform space to runtime:
+% how much wall time elapses on the given platform when solving the given
+% problem with the given implementation.
+% 
+% Auto-tuning is a family of empirical techniques for finding the implementation
+% that minimizes that runtime function, when the platform and the problem
+% configuration are given.
+% 
+% XXX: some math, or possibly a picture with the three boxes that was on Pinto's
+% whiteboard.
+% 
+% This paper is about different ways of doing the minimization.
+% For a particular problem (filterbank correlation) we
+% look at a parameterized kernel implementation (from \citet{pinto+cox:2011gcg})
+% and a few hardware platforms, and compare
+% random search, grid search, and a hill-climbing strategy with a good, but
+% generic, reference implementation.
+% Then we show that we can model the LSM function with a regression tree
+% well enough to usefully perform the argmin of Equation~\ref{eq:lsm_argmin}
+% on the model rather than the actual system. Whereas it may take several
+% minutes to perform a hill-climbing or grid search in the original system,
+% a hill-climbing search in the model requires a small fraction of a second.
+% Model-based autotuning makes it possible to simulate auto-tuning quickly and
+% accurately enough to be used within a single library call of the computational
+% routine.
+% 
+% Best arithmetic density
+% 
+% A bandwidth analysis that takes the blocking structure into account would
+% reveal that larger $H$ and $W$ require each block to read a larger input
+% region, and store it in a larger shared memory region, which reduces the
+% potential for high occupancy.
+% 
+% Filterbank correlation represents a difficult code-optimization problem
+% because of the number of interacting constraints: a large value of $P$ reduces
+% arithmetic density, but a small value of $P$ requires that each thread do more
+% work, and makes global memory access latency harder to hide. (XXX: is there a
+% better reason?)
+% 
+% 
+% The kernel used to perform
+% filterbank correlation was parametrized by 10 parameters, some of which
+% were binary and others of which were integer-valued. The full configuration
+% space included 12000 kernels.
+% XXX: how to make sense of these paramters without code listing?
+% 
+% XXX: What to all the parameters mean?
+% 
+% XXX: how many configurations are in the configuration space
+% 
+% XXX: what was the reference kernel, and how was it chosen?
+% 
+% XXX: analysis of READ traffic, WRITE traffic, DATA re-use and CACHEing
+% strategies. FLOPS per write etc.
+% 
+% XXX: Why not use FFT: convolution in spectral domain?
+% 
+% XXXX: Talk about memory transfer requirements vs. speed vs. copy...
+% 
+% we believe that our approach is the most practial one to get high perf on
+% recent hardware, since there is no consensus on hw features (and it won't
+% change soon)
+% 
+% XXX: deliver significant performance increases at very low run-time cost (less
+% than a second if the input domain has not been seen before, zero if it
+% has, since it will hit the cache) compared with the reference
+% implementation
+% 
+% XXX always delivered good performance: optimality without loss of productivity?
+% 
+% XXX such machine learning derived model-based empirical auto-tuner hold great
+% promise for dev productivity, portability, performance on future multi-core hw
+% with different architectural features
+% 
+% XXX frameworks for predicting performance exist \cite{XXX Hwu An Adaptive
+% Performance Modeling Tool for GPU Architectures}
+% but for non-GPU architectures
+% 
+% auto-tuning required hardward expertise *and* application-domain knowledge, we
+% want to remove these restrictions ultimately we want a ml-based approach to
+% code optimization that will where a set of generic but deep code
+% transformations, could be templated and injected in any domain-specific
+% library. this will reduce the need for specialized expertise across the range
+% of targeted computing systems.
+
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Future Work}

### 2011-12-14 paper/draft1.tex
Add main figure into result section
not sure

@@
 
-%XXX:  Machine Learning for Auto-tuning: A Data-Drive Approach...} ???
+%XXX:  Machine Learning for Auto-tuning: A Data-Driven Approach...} ???
 %\title{Boosted Regression Trees for Data-Driven Auto-Tuning}
@@
 % NP: I like this title !
-%XXX:  Machine Learning for Auto-tuning: A Data-Drive Approach...} ???
+%XXX:  Machine Learning for Auto-tuning: A Data-Driven Approach...} ???
 
@@
 correlation kernel described in \citet{pinto+cox:2011gcg}, where we find that
-0.1 seconds (XXX) of hill climbing on the regression model (which we dub ``{\em
+0.1 seconds of hill climbing on the regression model (which we dub ``{\em
 predictive} auto-tuning'') can achieve an average of 95\% of the
-speed-up brought by minutes of empirical auto-tuning XXX.  Our approach is not
+speed-up brought by minutes of empirical auto-tuning.  Our approach is not
 specific to filterbank correlation, nor even to GPU kernel auto-tuning, and can
@@
 
-XXX: use 300\% boost instead of 3x
+%XXX: use 300\% boost instead of 3x
 
@@
 
+\iffalse
+% XXX: paragraph has XXX in it and seems slightly tangential
 Even though such a ``black-box'' approach does not try to build an
@@
 we are trying to do w/ ML-based, decision trees, etc.).
+\fi
 
@@
 modelling techniques instead of constructing a model manually. By learning the model, one can
-hope to achieve elements of the best of both approaches: XXX.
-%and capitalize on the large body of
-%literature from the machine learning community.
+hope to achieve elements of the best of both approaches: the search speed of model-based auto-tuning with the broad applicability
+and ease of implementation of empirical auto-tuning.
 
@@
 
+Our
+
 \bibliography{local}

### 2011-12-14 paper/draft1.tex
Minor section name edit
Assorted typo fixes, word changes, XXX removals

--- 2011-12-14 paper/draft1.tex
@@
 effective technique for tailoring an optimal compute kernel for a particular
-hardware \cite{XXX}.
+hardware.
 % DDC: I'm still struggling with the above sentence.  Feel free to substitute your own ideas
@@
 
-XXX: use 300\% boost instead of 3x
+% XXX: use 300\% boost instead of 3x
 
@@
 
-Even though such a ``black-box'' approach does not try to build an
-explicit model per se, it could, if care were taken, help to greatly accelerate
+Even though such a ``black-box'' approach does not explicitly try to build an
+interpretable model, per se, it could nonetheless help to greatly accelerate
 our understanding of the complex interactions between the set of optimizations,
-their parameters and the hardware architectural features (XXX: that's what
-we are trying to do w/ ML-based, decision trees, etc.).
+their parameters and the hardware architectural features.
+% (XXX: that's what we are trying to do w/ ML-based, decision trees, etc.).
 
@@
 
-To alleviate this major weakness, it is intuitively appealing to combine the two
+To ameliorate this weakness, it is intuitively appealing to combine the two
 approaches by first constraining the search space with an analytical model and
 then exploring the reduced space empirically \citep{chen2005combining,
-li2009note}. Unfortunately, this hybrid approach is still bound by the quality
+li2009note}. Unfortunately, such a hybrid approach is still bound by the quality
 of the analytical model, which remains hard to build by hand.
@@
 
-In this paper, we propose to \emph{learn the model} using non-linear regression
-modelling techniques instead of constructing a model manually. By learning the model, one can
-hope to achieve elements of the best of both approaches: XXX.
+In this paper, we propose to \emph{learn} the model using non-linear regression
+modeling techniques, rather than constructing a model manually. By learning the model, one can
+hope to achieve elements of the best of both approaches. %: XXX.
 %and capitalize on the large body of
@@
 Relative to this work, our contribution is to show
-how to do fast predictive auto-tuning that satisifies the requirements to:
+how to do fast predictive auto-tuning that satisfies the requirements to:
 (a) handle the variety of recent multi-core architectures like GPUs \citep{schaa2009exploring},
@@
 The heart of predictive auto-tuning is a regression model that acts as surrogate
-for a hand-written hardware model or empirical timing estimates.
-In our experiments this regression model fits empirically measured timing information for a {\em subset} of the configuration space and interpolates / extrapolates that timing across the remainder of the space.
+for a hand-crafted hardware model or empirical timing estimates.
+In our experiments, this regression model fits empirically measured timing information for a {\em subset} of the configuration space and interpolates / extrapolates that timing across the remainder of the space.
 To fit this model, we
@@
 
+\begin{figure*}[!ht]
+\centering
+\includegraphics[width=.9\linewidth]{illustrator/fig_main_R1.pdf}
+\caption{
+XXX UPDATE THIS WITH (a) (b) (c) XXX
+Computation speed for novel problem configurations using predictive
+vs. empirical code specialization.
+The scatterplots are roughly symmetric about the axis of equality (main diagonal)
+with occasional outliers, indicating that for most problem configurations the
+predictive approach gives as good a solution as the empirical approach.
+The predictive approach typically requires about 0.1 seconds to suggest an
+implementation, whereas the empirical approach requires about 2 minutes.
+}
+\label{fig:fig_gflop_scatter}
+\end{figure*}
+
 Figure~\ref{fig:allstars} shows the effectiveness of empirical auto-tuning is in this setting.
@@
 
-\begin{figure}
-\centering
-\includegraphics[scale=.42]{fig_gflop_scatter_295.pdf}
-\includegraphics[scale=.42]{fig_gflop_scatter_480.pdf}
-\caption{Computation speed for novel problem configurations using predictive
-vs. empirical code specialization.
-The scatterplots are roughly symmetric about the axis of equality (main diagonal)
-with occasional outliers, indicating that for most problem configurations the
-predictive approach gives as good a solution as the empirical approach.
-The predictive approach typically requires about 0.1 seconds to suggest an
-implementation, whereas the empirical approach requires about 2 minutes.
-}
-\label{fig:fig_gflop_scatter}
-\end{figure}
-
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2011-12-14 paper/draft1.tex

@@
 
-% XXX: use 300\% boost instead of 3x
+%XXX: use 300\% boost instead of 3x
 
@@
 
-Even though such a ``black-box'' approach does not explicitly try to build an
-interpretable model, per se, it could nonetheless help to greatly accelerate
+\iffalse
+% XXX: paragraph has XXX in it and seems slightly tangential
+Even though such a ``black-box'' approach does not try to build an
+explicit model per se, it could, if care were taken, help to greatly accelerate
 our understanding of the complex interactions between the set of optimizations,
-their parameters and the hardware architectural features.
-% (XXX: that's what we are trying to do w/ ML-based, decision trees, etc.).
+their parameters and the hardware architectural features (XXX: that's what
+we are trying to do w/ ML-based, decision trees, etc.).
+\fi
 
@@
 
-In this paper, we propose to \emph{learn} the model using non-linear regression
-modeling techniques, rather than constructing a model manually. By learning the model, one can
-hope to achieve elements of the best of both approaches. %: XXX.
-%and capitalize on the large body of
-%literature from the machine learning community.
+In this paper, we propose to \emph{learn the model} using non-linear regression
+modelling techniques instead of constructing a model manually. By learning the model, one can
+hope to achieve elements of the best of both approaches: the search speed of model-based auto-tuning with the broad applicability
+and ease of implementation of empirical auto-tuning.
 
@@
 
-\iffalse
-XXX: note that image processing libraries include convolutions (e.g. OpenCV ???
-even for GPU!), but only don’t exploit the _bank_ dimension of the
-_filterbank_ and thus can’t exploit parallel hardware as much.
-
-XXX: in addition, when they are optimized, it’s often for pre-defined filters
-(e.g. sobel?), and they are not suitable for encapsulation into libraries
-usable by most real-world applications (bouuuuhhhh opencv !!!).
-\fi
-
 Filterbank correlation is a simple spatial image filtering operation that is
@@
 high arithmetic intensity which makes it a natural fit for GPU platforms~\citep{pinto+cox:2011gcg}.
-%XXX cite some computer vision papers.
 
@@
 this operation as a simple function that takes pointers and strides for $x$, $f$, and $z$ and performs the computation.
-However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
-related stencil operations, %XXX
+However, as shown in \citet{pinto+cox:2011gcg} 
 it is challenging to provide an implementation
@@
 generalized stencil kernels are not amenable to packaging as libraries.''
-% XXX: out of context?
-
-XXXXXXXXXXX
-
-Our contribution of the present work is to show that boosted regression tree models,
-a powerful machine learning technique for function approximation, can meet the
-requirements of this application.
-We call this approach {\em predictive} auto-tuning to contrast it with the standard measurement-based approach which we will call {\em empirical} auto-tuning.
-We show that the kernel from \citet{pinto+cox:2011gcg} can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.
-%XXX: this has no flow
 
@@
 The ``maxrreg'' and ``fast\_math'' parameters are passed to the nvcc compiler to limit
-the number of registers available to each thread, and to enable XXX,
-respectively.
-%XXX: what effect does fast-math have here?
+the number of registers available to each thread, and to enable more aggressive instruction selection, respectively.
 
@@
 
-\section{Experiment Setup}
+\section{Experimental Setup}
 
@@
 with depth channels being contiguous in memory, followed by columns, then rows, and then filters having the largest stride.
-
-%XXX: DIAGRAM OF IMAGE, FILTERBANK, OUTPUT
 
@@
 % backtracking
-In some cases, predictive auto-tuning yields an invalid implementation.
-We dealt with this scenario by backtracking through the various best-estimates discovered during the hill-climbing search.
-Some invalid kernels can only be discovered after compiling code and attempting to run the compiled code, so in these cases predictive auto-tuning took up to 3 seconds.
-Even in these cases the predictive model is much faster than empirical auto-tuning, because the first kernel that actually runs is a fast one.
+In some cases, predictive auto-tuning yields an invalid implementation.  We
+dealt with this scenario by backtracking through the various best-estimates
+discovered during the hill-climbing search.  Some invalid kernels can only be
+discovered after compiling code and attempting to run the compiled code, so in
+these cases predictive auto-tuning took up to 3 seconds.  Even in these cases
+the predictive model is much faster than empirical auto-tuning, because the
+first kernel that actually runs is a fast one.
 
@@
 
-The previous figures establish that auto-tuning on a problem-by-problem basis can achieve good performance improvements over a high-quality reference baseline, but they do not show how long it takes to find these implementations.
-A search of 75 iterations took an average of about 120 seconds XXX on a fast computer, because it requires compiling 75 CUDA kernels, and evaluating up to 750 filterbank correlations to get reliable timing information for each candidate implementation.
-
-Our contrib
-
-TL;DR: Hill-climbing is a good way to search the implementation space, can
-search the full space and do about as well as grid search in the more
-restricted space.
+The previous figures establish that auto-tuning on a problem-by-problem basis
+can achieve good performance improvements over a high-quality reference
+baseline, but they do not show how long it takes to find these implementations.
+A search of 75 iterations took an average of about 120 seconds XXX on a fast
+computer, because it requires compiling 75 CUDA kernels, and evaluating up to
+750 filterbank correlations to get reliable timing information for each
+candidate implementation.
 

### 2011-12-14 paper/scratch.tex
m

@@
+Our contribution of the present work is to show that boosted regression tree models,
+a powerful machine learning technique for function approximation, can meet the
+requirements of this application.
+We call this approach {\em predictive} auto-tuning to contrast it with the standard measurement-based approach which we will call {\em empirical} auto-tuning.
+We show that the kernel from \citet{pinto+cox:2011gcg} can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.
+%XXX: this has no flow
+
+\iffalse
+XXX: note that image processing libraries include convolutions (e.g. OpenCV ???
+even for GPU!), but only don’t exploit the _bank_ dimension of the
+_filterbank_ and thus can’t exploit parallel hardware as much.
+
+XXX: in addition, when they are optimized, it’s often for pre-defined filters
+(e.g. sobel?), and they are not suitable for encapsulation into libraries
+usable by most real-world applications (bouuuuhhhh opencv !!!).
+\fi
+
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2011-12-14 paper/draft1.tex
Shorten refs

@@
 
-%XXX:  Machine Learning for Auto-tuning: A Data-Driven Approach...} ???
+%XXX:  Machine Learning for Auto-tuning: A Data-Drive Approach...} ???
 %\title{Boosted Regression Trees for Data-Driven Auto-Tuning}
@@
 % NP: I like this title !
-%XXX:  Machine Learning for Auto-tuning: A Data-Driven Approach...} ???
+%XXX:  Machine Learning for Auto-tuning: A Data-Drive Approach...} ???
 
@@
 correlation kernel described in \citet{pinto+cox:2011gcg}, where we find that
-0.1 seconds of hill climbing on the regression model (which we dub ``{\em
+0.1 seconds (XXX) of hill climbing on the regression model (which we dub ``{\em
 predictive} auto-tuning'') can achieve an average of 95\% of the
-speed-up brought by minutes of empirical auto-tuning.  Our approach is not
+speed-up brought by minutes of empirical auto-tuning XXX.  Our approach is not
 specific to filterbank correlation, nor even to GPU kernel auto-tuning, and can
@@
 
-%XXX: use 300\% boost instead of 3x
+XXX: use 300\% boost instead of 3x
 
@@
 
-\iffalse
-% XXX: paragraph has XXX in it and seems slightly tangential
 Even though such a ``black-box'' approach does not try to build an
@@
 we are trying to do w/ ML-based, decision trees, etc.).
-\fi
 
@@
 modelling techniques instead of constructing a model manually. By learning the model, one can
-hope to achieve elements of the best of both approaches: the search speed of model-based auto-tuning with the broad applicability
-and ease of implementation of empirical auto-tuning.
+hope to achieve elements of the best of both approaches: XXX.
+%and capitalize on the large body of
+%literature from the machine learning community.
 
@@
 
+\iffalse
+XXX: note that image processing libraries include convolutions (e.g. OpenCV ???
+even for GPU!), but only don’t exploit the _bank_ dimension of the
+_filterbank_ and thus can’t exploit parallel hardware as much.
+
+XXX: in addition, when they are optimized, it’s often for pre-defined filters
+(e.g. sobel?), and they are not suitable for encapsulation into libraries
+usable by most real-world applications (bouuuuhhhh opencv !!!).
+\fi
+
 Filterbank correlation is a simple spatial image filtering operation that is
@@
 high arithmetic intensity which makes it a natural fit for GPU platforms~\citep{pinto+cox:2011gcg}.
+%XXX cite some computer vision papers.
 
@@
 this operation as a simple function that takes pointers and strides for $x$, $f$, and $z$ and performs the computation.
-However, as shown in \citet{pinto+cox:2011gcg} 
+However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
+related stencil operations, %XXX
 it is challenging to provide an implementation
@@
 generalized stencil kernels are not amenable to packaging as libraries.''
+% XXX: out of context?
+
+XXXXXXXXXXX
+
+Our contribution of the present work is to show that boosted regression tree models,
+a powerful machine learning technique for function approximation, can meet the
+requirements of this application.
+We call this approach {\em predictive} auto-tuning to contrast it with the standard measurement-based approach which we will call {\em empirical} auto-tuning.
+We show that the kernel from \citet{pinto+cox:2011gcg} can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.
+%XXX: this has no flow
 
@@
 The ``maxrreg'' and ``fast\_math'' parameters are passed to the nvcc compiler to limit
-the number of registers available to each thread, and to enable more aggressive instruction selection, respectively.
+the number of registers available to each thread, and to enable XXX,
+respectively.
+%XXX: what effect does fast-math have here?
 
@@
 
-This kernel was implemented in the meta-programming style advocated
-in \citet{pinto+cox:2011gcg} in Python using
-Cheetah for string processing~\citep{cheetah}
-and PyCUDA~\citep{klochner+etal:2009} for dynamic kernel
-compilation and interfacing with CUDA.
+This kernel was implemented in the meta-programming style advocated in
+\citet{pinto+cox:2011gcg} in Python using Cheetah for string
+processing~\citep{cheetah} and PyCUDA~\citep{klockner2011pycuda} for dynamic
+kernel compilation and interfacing with CUDA.
 
@@
 
+%XXX: DIAGRAM OF IMAGE, FILTERBANK, OUTPUT
+
 The product space in our study
@@
 \section{Results}
+
+\begin{figure*}[!ht]
+\centering
+\includegraphics[width=.9\linewidth]{illustrator/fig_main_R1.pdf}
+\caption{
+XXX UPDATE THIS WITH (a) (b) (c) XXX
+Computation speed for novel problem configurations using predictive
+vs. empirical code specialization.
+The scatterplots are roughly symmetric about the axis of equality (main diagonal)
+with occasional outliers, indicating that for most problem configurations the
+predictive approach gives as good a solution as the empirical approach.
+The predictive approach typically requires about 0.1 seconds to suggest an
+implementation, whereas the empirical approach requires about 2 minutes.
+}
+\label{fig:fig_gflop_scatter}
+\end{figure*}
 
@@
 % backtracking
-In some cases, predictive auto-tuning yields an invalid implementation.  We
-dealt with this scenario by backtracking through the various best-estimates
-discovered during the hill-climbing search.  Some invalid kernels can only be
-discovered after compiling code and attempting to run the compiled code, so in
-these cases predictive auto-tuning took up to 3 seconds.  Even in these cases
-the predictive model is much faster than empirical auto-tuning, because the
-first kernel that actually runs is a fast one.
+In some cases, predictive auto-tuning yields an invalid implementation.
+We dealt with this scenario by backtracking through the various best-estimates discovered during the hill-climbing search.
+Some invalid kernels can only be discovered after compiling code and attempting to run the compiled code, so in these cases predictive auto-tuning took up to 3 seconds.
+Even in these cases the predictive model is much faster than empirical auto-tuning, because the first kernel that actually runs is a fast one.
 
@@
 
-The previous figures establish that auto-tuning on a problem-by-problem basis
-can achieve good performance improvements over a high-quality reference
-baseline, but they do not show how long it takes to find these implementations.
-A search of 75 iterations took an average of about 120 seconds XXX on a fast
-computer, because it requires compiling 75 CUDA kernels, and evaluating up to
-750 filterbank correlations to get reliable timing information for each
-candidate implementation.
+The previous figures establish that auto-tuning on a problem-by-problem basis can achieve good performance improvements over a high-quality reference baseline, but they do not show how long it takes to find these implementations.
+A search of 75 iterations took an average of about 120 seconds XXX on a fast computer, because it requires compiling 75 CUDA kernels, and evaluating up to 750 filterbank correlations to get reliable timing information for each candidate implementation.
+
+Our contrib
+
+TL;DR: Hill-climbing is a good way to search the implementation space, can
+search the full space and do about as well as grid search in the more
+restricted space.
 
@@
 
-Our
-
+{\small
     \bibliography{local}
+}
 \end{document}

### 2011-12-14 paper/draft1.tex

--- 2011-12-14 paper/draft1.tex
@@
 \caption{
-XXX UPDATE THIS WITH (a) (b) (c) XXX
 Computation speed for novel problem configurations using predictive
-vs. empirical code specialization.
-The scatterplots are roughly symmetric about the axis of equality (main diagonal)
-with occasional outliers, indicating that for most problem configurations the
-predictive approach gives as good a solution as the empirical approach.
-The predictive approach typically requires about 0.1 seconds to suggest an
-implementation, whereas the empirical approach requires about 2 minutes.
+vs. empirical auto-tuning.
+(a) As more problem configurations are used to train the regression model,
+the performance of predictive auto-tuning approaches that of empirical auto-tuning
+despite taking only 0.1 seconds instead of 2 minutes.
+(b) and (c) For a wide range of devices, implementation speeds found by predictive auto-tuning correlate tightly with
+implementation speeds found by empirical auto-tuning.
+%The boosted regression trees used for predictive auto-tuning generalize correctly across the implementation space.
 }
@@
 
-Figure~\ref{XXX} shows how accurate predictive auto-tuning is compared with empirical auto-tuning.
+Figure~\ref{fig:fig_gflop_scatter} shows how accurate predictive auto-tuning is compared with empirical auto-tuning.
 The training set $(\mathcal{X}, \mathcal{Y})$ for the regression model comprised
@@
 So for each training argument configuration $a,c$ there were 148 different values of $b$ and thus 148 training points.
-Figure~\ref{XXX} shows that as more argument configurations are used for training,
+Figure~\ref{fig:fig_gflop_scatter} (a) shows that as more argument configurations are used for training,
 the performance of predictive auto-tuning on test configurations $(a', c)$ quickly
@@
 typically took about 0.1 seconds per test example, whereas empirical auto-tuning took about 1-3 minutes.
+
+Figure~\ref{fig:fig_gflop_scatter} (b) and (c) show that the implementation speeds found by predictive auto-tuning correlates very well with the speed of empirical auto-tuning across all of the devices we tested: GTX 580, GTX 480, GTX 296, Tesla C2070, and Tesla C1060 which span three generations of CUDA platforms.
 
@@
 
+% DDC: XXX are these figures still in?
 Figure~\ref{fig:speedup} also shows that random hill-climbing is at least
@@
 
-\section{Conclusion}
-
-Our
-
+% DDC: XXX is "Conclusion" a "required" section in any sense.  If not, I think we could probably do without
+%\section{Conclusion}
+
+%Our
+
+{\small
     \bibliography{local}
+}
 \end{document}

### 2011-12-14 paper/draft1.tex

@@
 b) empirical auto-tuning over a hill-climbing search, and
-c) predictive autotuning.
+c) predictive auto-tuning.
 %Section~\ref{sec:discussion}.
@@
 \includegraphics[scale=.42]{allstars_mixup_295.pdf}
-% XXX: ERROR BARS
+% XXX: NPTODO: ERROR BARS
+% XXX: NPTODO: fix fonts for better proportions
 \caption{Different arguments call for different kernels:
-{\em left} is the average argument-specific empirical auto-tuned performance across 100 random argument configurations,
+{\em left} is the average argument-specific empirical
+auto-tuned performance across 100 random argument configurations,
 {\em middle} is the average speed of our reference implementation,
@@
 
-Figure~\ref{fig:fig_gflop_scatter} (b) and (c) show that the implementation speeds found by predictive auto-tuning correlates very well with the speed of empirical auto-tuning across all of the devices we tested: GTX 580, GTX 480, GTX 296, Tesla C2070, and Tesla C1060 which span three generations of CUDA platforms.
+Figure~\ref{fig:fig_gflop_scatter} (b) and (c) show that the implementation speeds found by predictive auto-tuning correlates very well with the speed of empirical auto-tuning across all of the devices we tested: GTX 580, GTX 480, GTX 295, Tesla C2070, and Tesla C1060 which span three generations of CUDA platforms.
 
@@
 Figure~\ref{fig:speedup} also shows that random hill-climbing is at least
-competitive and often better than the hand-chosen grid approach of ~\citet{pinto+cox:2011gcg}.  Figure~\ref{fig:hcX} shows how the number of hill-climbing iterations
+competitive and often better than the hand-chosen grid approach
+of~\citet{pinto+cox:2011gcg}.  Figure~\ref{fig:hcX} shows how the number of hill-climbing iterations
 affects the quality of the auto-tuned model.
@@
 \centering
-\includegraphics[scale=.42]{fig_ntrain_295.pdf} % on vader
-\includegraphics[scale=.42]{fig_ntrain_480.pdf} % on vader
+\includegraphics[scale=.42]{fig_ntrain_munctional0_580.pdf} % on vader
 \caption{The effect of the invalid configuration score ($a$) and training set
@@
 }
-\label{fig:fig_ntrain}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.42]{fig_ntrain_munctional0_1060.pdf}
-\includegraphics[scale=.42]{fig_ntrain_munctional0_2070.pdf}
-\caption{more timing on munctional0}
-\label{fig:fig_ntrain}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.42]{fig_ntrain_munctional0_480.pdf}
-\includegraphics[scale=.42]{fig_ntrain_munctional0_580.pdf}
-\caption{more timing on munctional0}
 \label{fig:fig_ntrain}
@@
 % a hill-climbing search in the model requires a small fraction of a second.
-% Model-based autotuning makes it possible to simulate auto-tuning quickly and
+% Model-based auto-tuning makes it possible to simulate auto-tuning quickly and
 % accurately enough to be used within a single library call of the computational

### 2011-12-14 paper/draft1.tex
Reverting back to an older version of the first sentence

@@
 
-Auto-tuning of parameteric, instrumented code has proven to be a highly
-effective technique for tailoring an optimal compute kernel for a particular
+Auto-tuning techniques are widely used and effective for optimizing a
+parametrized GPU code template for a particular computation on particular
 hardware.
-% DDC: I'm still struggling with the above sentence.  Feel free to substitute your own ideas
-However, a drawback of this approach is that exhaustive, or even thorough,
+However, a drawback of this approach is that thorough or exhaustive
 auto-tuning requires compiling many kernels and calling each one many times,

### 2011-12-14 paper/draft1.tex
text for nicos compound plot

--- 2011-12-14 paper/draft1.tex
@@
 effective technique for tailoring an optimal compute kernel for a particular
-hardware.
+hardware \cite{XXX}.
 % DDC: I'm still struggling with the above sentence.  Feel free to substitute your own ideas
@@
 
-To ameliorate this weakness, it is intuitively appealing to combine the two
+To alleviate this major weakness, it is intuitively appealing to combine the two
 approaches by first constraining the search space with an analytical model and
 then exploring the reduced space empirically \citep{chen2005combining,
-li2009note}. Unfortunately, such a hybrid approach is still bound by the quality
+li2009note}. Unfortunately, this hybrid approach is still bound by the quality
 of the analytical model, which remains hard to build by hand.
@@
 Relative to this work, our contribution is to show
-how to do fast predictive auto-tuning that satisfies the requirements to:
+how to do fast predictive auto-tuning that satisifies the requirements to:
 (a) handle the variety of recent multi-core architectures like GPUs \citep{schaa2009exploring},
@@
 The heart of predictive auto-tuning is a regression model that acts as surrogate
-for a hand-crafted hardware model or empirical timing estimates.
-In our experiments, this regression model fits empirically measured timing information for a {\em subset} of the configuration space and interpolates / extrapolates that timing across the remainder of the space.
+for a hand-written hardware model or empirical timing estimates.
+In our experiments this regression model fits empirically measured timing information for a {\em subset} of the configuration space and interpolates / extrapolates that timing across the remainder of the space.
 To fit this model, we
@@
 
-This kernel was implemented in the meta-programming style advocated in
-\citet{pinto+cox:2011gcg} in Python using Cheetah for string
-processing~\citep{cheetah} and PyCUDA~\citep{klockner2011pycuda} for dynamic
-kernel compilation and interfacing with CUDA.
+This kernel was implemented in the meta-programming style advocated
+in \citet{pinto+cox:2011gcg} in Python using
+Cheetah for string processing~\citep{cheetah}
+and PyCUDA~\citep{klochner+etal:2009} for dynamic kernel
+compilation and interfacing with CUDA.
 
@@
 
-% DDC: XXX are these figures still in?
 Figure~\ref{fig:speedup} also shows that random hill-climbing is at least
@@
 
-% DDC: XXX is "Conclusion" a "required" section in any sense.  If not, I think we could probably do without
-%\section{Conclusion}
-
-%Our
-
-{\small
+\section{Conclusion}
+
+Our
+
 \bibliography{local}
-}
 \end{document}

### 2011-12-14 paper/draft1.tex

--- 2011-12-14 paper/draft1.tex
@@
 b) empirical auto-tuning over a hill-climbing search, and
-c) predictive auto-tuning.
+c) predictive autotuning.
 %Section~\ref{sec:discussion}.
@@
 \includegraphics[scale=.42]{allstars_mixup_295.pdf}
-% XXX: NPTODO: ERROR BARS
-% XXX: NPTODO: fix fonts for better proportions
+% XXX: ERROR BARS
 \caption{Different arguments call for different kernels:
-{\em left} is the average argument-specific empirical
-auto-tuned performance across 100 random argument configurations,
+{\em left} is the average argument-specific empirical auto-tuned performance across 100 random argument configurations,
 {\em middle} is the average speed of our reference implementation,
@@
 
-Figure~\ref{fig:fig_gflop_scatter} (b) and (c) show that the implementation speeds found by predictive auto-tuning correlates very well with the speed of empirical auto-tuning across all of the devices we tested: GTX 580, GTX 480, GTX 295, Tesla C2070, and Tesla C1060 which span three generations of CUDA platforms.
+Figure~\ref{fig:fig_gflop_scatter} (b) and (c) show that the implementation speeds found by predictive auto-tuning correlates very well with the speed of empirical auto-tuning across all of the devices we tested: GTX 580, GTX 480, GTX 296, Tesla C2070, and Tesla C1060 which span three generations of CUDA platforms.
 
@@
 Figure~\ref{fig:speedup} also shows that random hill-climbing is at least
-competitive and often better than the hand-chosen grid approach
-of~\citet{pinto+cox:2011gcg}.  Figure~\ref{fig:hcX} shows how the number of hill-climbing iterations
+competitive and often better than the hand-chosen grid approach of ~\citet{pinto+cox:2011gcg}.  Figure~\ref{fig:hcX} shows how the number of hill-climbing iterations
 affects the quality of the auto-tuned model.
@@
 \centering
-\includegraphics[scale=.42]{fig_ntrain_munctional0_580.pdf} % on vader
+\includegraphics[scale=.42]{fig_ntrain_295.pdf} % on vader
+\includegraphics[scale=.42]{fig_ntrain_480.pdf} % on vader
 \caption{The effect of the invalid configuration score ($a$) and training set
@@
 }
+\label{fig:fig_ntrain}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.42]{fig_ntrain_munctional0_1060.pdf}
+\includegraphics[scale=.42]{fig_ntrain_munctional0_2070.pdf}
+\caption{more timing on munctional0}
+\label{fig:fig_ntrain}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.42]{fig_ntrain_munctional0_480.pdf}
+\includegraphics[scale=.42]{fig_ntrain_munctional0_580.pdf}
+\caption{more timing on munctional0}
 \label{fig:fig_ntrain}
@@
 % a hill-climbing search in the model requires a small fraction of a second.
-% Model-based auto-tuning makes it possible to simulate auto-tuning quickly and
+% Model-based autotuning makes it possible to simulate auto-tuning quickly and
 % accurately enough to be used within a single library call of the computational
@@
 
-{\small
 \bibliography{local}
-}
 \end{document}

### 2011-12-14 x_allstars_mixup.sh
na

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+python fig.py allstars_mixup timings/munctional0/timing1_580_big.pkl timings/munctional0/timing1_580_big.pkl_allstars_mixup_timings

### 2011-12-14 x_allstars_mixup.sh
na

@@ -1,3 +1,5 @@
 #!/bin/bash
 
 python fig.py allstars_mixup timings/munctional0/timing1_580_big.pkl timings/munctional0/timing1_580_big.pkl_allstars_mixup_timings
+
+mv -vf fig_allstars_mixup_580.pdf paper

### 2011-12-14 paper/draft1.tex
cutting most line figures

--- 2011-12-14 paper/draft1.tex
@@
 
-Auto-tuning techniques are widely used and effective for optimizing a
-parametrized GPU code template for a particular computation on particular
+Auto-tuning of parameteric, instrumented code has proven to be a highly
+effective technique for tailoring an optimal compute kernel for a particular
 hardware.
-However, a drawback of this approach is that thorough or exhaustive
+% DDC: I'm still struggling with the above sentence.  Feel free to substitute your own ideas
+However, a drawback of this approach is that exhaustive, or even thorough,
 auto-tuning requires compiling many kernels and calling each one many times,

### 2011-12-14 paper/draft1.tex
m

@@
 
-% DDC: XXX are these figures still in?
-Figure~\ref{fig:speedup} also shows that random hill-climbing is at least
-competitive and often better than the hand-chosen grid approach
-of~\citet{pinto+cox:2011gcg}.  Figure~\ref{fig:hcX} shows how the number of hill-climbing iterations
-affects the quality of the auto-tuned model.
-The quicker HC25 approach is clearly inferior but the quality of the 75 iteration search (HC75) is statistically similar to the shorter HC50, suggesting that XXX.
-Still, HC75 is not always better than the grid, so evidently neither search algorithm is perfect.
-
-The previous figures establish that auto-tuning on a problem-by-problem basis
-can achieve good performance improvements over a high-quality reference
-baseline, but they do not show how long it takes to find these implementations.
-A search of 75 iterations took an average of about 120 seconds XXX on a fast
-computer, because it requires compiling 75 CUDA kernels, and evaluating up to
-750 filterbank correlations to get reliable timing information for each
-candidate implementation.
-
 \begin{figure}
@@
 \includegraphics[scale=.42]{fig_ntrain_munctional0_580.pdf} % on vader
-\caption{The effect of the invalid configuration score ($a$) and training set
+\caption{The effect of the invalid configuration score ($\zeta$) and training set
 size on simulated auto-tuning.
 All candidates timed during the grid and
-hill-climbing search procedures were used as training points, so the training
-set sizes ranged from an average of 1500 (10 problem configurations) to 30
-thousand (200 problem configurations).
-Training on 10 or 25 configurations was useful (higher mean than the reference),
+hill-climbing search procedures were used as training examples, so the training
+set sizes ranged from an average of 1480 (10 problem configurations) to 29 600
+(200 problem configurations).
+Training from the largest training sets took approximately 30 seconds
+Models trained on as few as 25 configurations outperformed the reference.
+Models trained
 but not as useful as training on 50 or more configurations.
-The results on the GTX 295 suggest that
-moderate values of $a$ between $0.25$ and $0.75$ might be
-best, but $a$ had no significant effect on the GTX 480.
+Moderate values of $\zeta$ between $\log(0.25)$ and $\log(0.75)$ are best.
 }

### 2011-12-14 paper/scratch.tex
m

@@
+% DDC: XXX are these figures still in?
+Figure~\ref{fig:speedup} also shows that random hill-climbing is at least
+competitive and often better than the hand-chosen grid approach
+of~\citet{pinto+cox:2011gcg}.  Figure~\ref{fig:hcX} shows how the number of hill-climbing iterations
+affects the quality of the auto-tuned model.
+The quicker HC25 approach is clearly inferior but the quality of the 75 iteration search (HC75) is statistically similar to the shorter HC50, suggesting that XXX.
+Still, HC75 is not always better than the grid, so evidently neither search algorithm is perfect.
+
+The previous figures establish that auto-tuning on a problem-by-problem basis
+can achieve good performance improvements over a high-quality reference
+baseline, but they do not show how long it takes to find these implementations.
+A search of 75 iterations took an average of about 120 seconds XXX on a fast
+computer, because it requires compiling 75 CUDA kernels, and evaluating up to
+750 filterbank correlations to get reliable timing information for each
+candidate implementation.
+
 Our contribution of the present work is to show that boosted regression tree models,

### 2011-12-14 paper/draft1.tex

@@
 
+{\small
     \bibliography{local}
+}
 \end{document}

### 2011-12-14 paper/draft1.tex
More edits

--- 2011-12-14 paper/draft1.tex
@@
 correlation kernel described in \citet{pinto+cox:2011gcg}, where we find that
-0.1 seconds of hill climbing on the regression model (which we dub ``{\em
+0.1 seconds of hill climbing on the regression model (``{\em
 predictive} auto-tuning'') can achieve an average of 95\% of the
@@
 altered the landscape of application development: programmers are now
-facing a wide diversity of low-level architectural features to write
-high-performance \emph{and} portable code.
+facing a wide diversity of low-level architectural issues
+that must be carefully balanced if one is to write code that is both
+high-performance \emph{and} portable.
 
@@
 remains a daunting task. Even within specialized architectures from the
-\emph{same} vendor, like NVIDIA's Graphics Processing Units (GPUs) and the
+\emph{same} vendor, such as NVIDIA's Graphics Processing Units (GPUs) and the
 Compute Unified Device Architecture (CUDA) \citep{nickolls2008scalable,
@@
 The model-driven optimization approach uses analytical abstractions to model
-the hardware architectures, the possible code transformations and their complex
+the hardware architectures, in order to identify possible code transformations and their complex
 interactions. Even though highly-accurate analytical models are generally
-intractable to build, this approach has been quite successful in the past at
-accelerating serial code with simplified but general abstractions. However,
+difficult to build, this approach has been quite successful in the past, especially for 
+accelerating serial code, utilizing simplified but general abstractions. However,
 large speed-ups for parallel code require more accurate high-dimensional models
@@
 abstraction, it has been less suited for highly-specialized kernels. This
-approach has been dominant in the compiler community, and as a result, it has
-generally been applied at compile-time where important run-time characteristics
-such as input domains are missing. These drawbacks render the model-driven
-optimization approach less attractive for high-performance library developers.
+approach has been well-developed in the compiler community, and as a result, it has
+most often been applied at compile-time where important run-time characteristics
+such as input domains may be missing. These limitations render the model-driven
+optimization approach less attractive in many high-performance library development settings.
 
@@
 
-The empirical optimization approach, on the contrary, seeks to find the best
+The empirical optimization approach, in contrast, seeks to find the best
 performing code configuration by automatically generating many versions of a
@@
 runtime, when contextual information about the hardware and software stack is
-the richest). This method directly optimizes the metric(s) of interest and not
-surrogates. One of its main advantage is that it allows any metric to be
-optimized without loss of generality. It is indeed possible to formulate the
-problem as a multi-objective optimization and minimize both speed \emph{and}
-power consumption \citep{rahman2011automated}, a feat that renders the
-analytical model-driven approach even more difficult. Due to its flexibility,
-empirical auto-tuning has been successfully applied to build high-performance
+the richest). This method directly optimizes the metric(s) of interest (e.g. performance) and does not rely on
+surrogates.  A significant advantage of such approaches is that they allow any metric to be
+optimized without loss of generality. Indeed, it is possible to formulate a multi-objective optimization that minimize both run time \emph{and}
+power consumption \citep{rahman2011automated}, a feat that would be even more difficult using an
+analytical model-driven approach. Due to its flexibility,
+empirical auto-tuning has been successfully applied to build a variety of high-performance
 domain-specific libraries including dense linear algebra
@@
 optimizations \emph{and} to the search method. The size of the search space is
-so large that the current best empirical auto-tuners can only consider
+often so large that the current best empirical auto-tuners typically only consider
 highly-specialized functions with a limited set of code transformations and
-compiler options, on very few input domains \citep{ganapathi2009case}. Although
-searching for good code configurations in this highly-discontinuous space can
-be made embarrassingly parallel, it remains a very difficult and prohibitively
-expensive combinatorial optimization problem as many variants of the code must
+compiler options, on a limited set of input domains \citep{ganapathi2009case}. Although
+searches for good code configurations in highly-discontinuous spaces can
+be made ``embarrassingly'' parallel, and thus benefit from parallel execution across many devices, it remains a prohibitively
+expensive combinatorial optimization problem, as many variants of the code must
 generated, compiled, and benchmarked on specific input domains with meaningful

### 2011-12-14 paper/draft1.tex
Minor wording edits

@@
 methods prune the space with hard-coded heuristics that offer little
-generalization guarantees. This has been the main drawback of the empirical
-approach compared to the model-driven approach where good code configurations
+generalization guarantees. This has been a key drawback of the empirical
+approach as compared to the model-driven approach, where good code configurations
 can be directly derived from the analytical model.

### 2011-12-14 paper/draft1.tex
m

--- 2011-12-14 paper/draft1.tex
@@
 correlation kernel described in \citet{pinto+cox:2011gcg}, where we find that
-0.1 seconds of hill climbing on the regression model (``{\em
+0.1 seconds of hill climbing on the regression model (which we dub ``{\em
 predictive} auto-tuning'') can achieve an average of 95\% of the
@@
 altered the landscape of application development: programmers are now
-facing a wide diversity of low-level architectural issues
-that must be carefully balanced if one is to write code that is both
-high-performance \emph{and} portable.
+facing a wide diversity of low-level architectural features to write
+high-performance \emph{and} portable code.
 
@@
 remains a daunting task. Even within specialized architectures from the
-\emph{same} vendor, such as NVIDIA's Graphics Processing Units (GPUs) and the
+\emph{same} vendor, like NVIDIA's Graphics Processing Units (GPUs) and the
 Compute Unified Device Architecture (CUDA) \citep{nickolls2008scalable,
@@
 The model-driven optimization approach uses analytical abstractions to model
-the hardware architectures, in order to identify possible code transformations and their complex
+the hardware architectures, the possible code transformations and their complex
 interactions. Even though highly-accurate analytical models are generally
-difficult to build, this approach has been quite successful in the past, especially for 
-accelerating serial code, utilizing simplified but general abstractions. However,
+intractable to build, this approach has been quite successful in the past at
+accelerating serial code with simplified but general abstractions. However,
 large speed-ups for parallel code require more accurate high-dimensional models
@@
 abstraction, it has been less suited for highly-specialized kernels. This
-approach has been well-developed in the compiler community, and as a result, it has
-most often been applied at compile-time where important run-time characteristics
-such as input domains may be missing. These limitations render the model-driven
-optimization approach less attractive in many high-performance library development settings.
+approach has been dominant in the compiler community, and as a result, it has
+generally been applied at compile-time where important run-time characteristics
+such as input domains are missing. These drawbacks render the model-driven
+optimization approach less attractive for high-performance library developers.
 
@@
 
-The empirical optimization approach, in contrast, seeks to find the best
+The empirical optimization approach, on the contrary, seeks to find the best
 performing code configuration by automatically generating many versions of a
@@
 runtime, when contextual information about the hardware and software stack is
-the richest). This method directly optimizes the metric(s) of interest (e.g. performance) and does not rely on
-surrogates.  A significant advantage of such approaches is that they allow any metric to be
-optimized without loss of generality. Indeed, it is possible to formulate a multi-objective optimization that minimize both run time \emph{and}
-power consumption \citep{rahman2011automated}, a feat that would be even more difficult using an
-analytical model-driven approach. Due to its flexibility,
-empirical auto-tuning has been successfully applied to build a variety of high-performance
+the richest). This method directly optimizes the metric(s) of interest and not
+surrogates. One of its main advantage is that it allows any metric to be
+optimized without loss of generality. It is indeed possible to formulate the
+problem as a multi-objective optimization and minimize both speed \emph{and}
+power consumption \citep{rahman2011automated}, a feat that renders the
+analytical model-driven approach even more difficult. Due to its flexibility,
+empirical auto-tuning has been successfully applied to build high-performance
 domain-specific libraries including dense linear algebra
@@
 
-\iffalse
-% XXX: paragraph has XXX in it and seems slightly tangential
-Even though such a ``black-box'' approach does not try to build an
-explicit model per se, it could, if care were taken, help to greatly accelerate
-our understanding of the complex interactions between the set of optimizations,
-their parameters and the hardware architectural features (XXX: that's what
-we are trying to do w/ ML-based, decision trees, etc.).
-\fi
-
 % empirical optimization: disadvantages
@@
 optimizations \emph{and} to the search method. The size of the search space is
-often so large that the current best empirical auto-tuners typically only consider
+so large that the current best empirical auto-tuners can only consider
 highly-specialized functions with a limited set of code transformations and
-compiler options, on a limited set of input domains \citep{ganapathi2009case}. Although
-searches for good code configurations in highly-discontinuous spaces can
-be made ``embarrassingly'' parallel, and thus benefit from parallel execution across many devices, it remains a prohibitively
-expensive combinatorial optimization problem, as many variants of the code must
+compiler options, on very few input domains \citep{ganapathi2009case}. Although
+searching for good code configurations in this highly-discontinuous space can
+be made embarrassingly parallel, it remains a very difficult and prohibitively
+expensive combinatorial optimization problem as many variants of the code must
 generated, compiled, and benchmarked on specific input domains with meaningful
@@
 % XXX: there are 4 versions of the stencil work - consolidate to just the full journal version
-
-\iffalse
-XXX WIIFY here ??? (what's in their for you) XXX in text, say what's in this paper for:
-- compiler people
-- libs people
-- ml people
-- tools people
-- cv people
-\fi
 
@@
 The effectiveness of predictive auto-tuning depends on the mapping between the raw kernel timings $t(a, b, c)$ (i.e. in seconds) and the utility $y$ associated with that timing.
-Anticipating that regression involves minimizing the squared error of our predictor (see Eq.~\ref{XXXbelow}) it is important to choose $y$ so that differences of a given numerical {\em magnitude} correspond to {\em improvements} of a certain utility.
+Anticipating that regression involves minimizing the squared error of our predictor (see Eq.~\ref{eq:y}) it is important to choose $y$ so that differences of a given numerical {\em magnitude} correspond to {\em improvements} of a certain utility.
 In program optimization we are interested in improving the speedup over a reference implementation $b^{(\mathrm{ref})}$, so it is natural to choose
@@
 = \log\left(\frac{t(a, b^{(\mathrm{ref})}, c)}{t(a, b, c)} \right)
+\label{eq:y}
 \end{equation}
@@
 
+Figure~\ref{fig:fig_ntrain} shows how the training set size
+and the value of $\zeta$ affect the accuracy of predictive auto-tuning.
+All candidates timed during the grid and
+hill-climbing search procedures were used as training examples, so the training
+set sizes ranged from an average of 1480 (10 problem configurations) to 29 600
+(200 problem configurations).
+Training from the largest training sets took approximately 30 seconds.
+Training 50 or more problem configurations yielded quite accurate predictions
+with $\zeta=\log(0.5)$, which was the best value for $\zeta$ across the range of
+devices in our study.
+
 \begin{figure}
@@
 \caption{The effect of the invalid configuration score ($\zeta$) and training set
-size on simulated auto-tuning.
-All candidates timed during the grid and
-hill-climbing search procedures were used as training examples, so the training
-set sizes ranged from an average of 1480 (10 problem configurations) to 29 600
-(200 problem configurations).
-Training from the largest training sets took approximately 30 seconds
+size on predictive auto-tuning.
 Models trained on as few as 25 configurations outperformed the reference.
-Models trained
-but not as useful as training on 50 or more configurations.
+Models trained on 50 or more configurations rivaled empirical auto-tuning.
 Moderate values of $\zeta$ between $\log(0.25)$ and $\log(0.75)$ are best.
@@
 % Platform Space: XXX
-% 
-% XXX: GCG chapter pg 13 notes that among the grid search, the best parameter
-% setting is different for the various platforms.
-% 
-% XXX: so far we only tried 3 cards... so we're using a 1-of-N feature.
-% 
 % 

### 2011-12-14 paper/scratch.tex
m

@@
+\iffalse
+% XXX: paragraph has XXX in it and seems slightly tangential
+Even though such a ``black-box'' approach does not try to build an
+explicit model per se, it could, if care were taken, help to greatly accelerate
+our understanding of the complex interactions between the set of optimizations,
+their parameters and the hardware architectural features (XXX: that's what
+we are trying to do w/ ML-based, decision trees, etc.).
+\fi
+
+\iffalse
+XXX WIIFY here ??? (what's in their for you) XXX in text, say what's in this paper for:
+- compiler people
+- libs people
+- ml people
+- tools people
+- cv people
+\fi
+
 % DDC: XXX are these figures still in?

### 2011-12-14 paper/draft1.tex
m
Add bar plot

@@
 \definecolor{darkred}{rgb}{.5,0,0}
+
+% TODO: proof read for GFLOP vs. GFLOPS -> we want GFLOP
 
@@
 
-For the implementation features $b$, we directly used the integer and binary values (block\_w, block\_h, etc.) that parametrized the kernels.
-We did not use platform features ($c$) in our experiments. We leave the investigation
-of cross-platform predictive auto-tuning for future work.
+For the implementation features $b$, we directly used the integer and binary
+values (block\_w, block\_h, etc.) that parametrized the kernels.  We did not
+use platform features ($c$) in our experiments. We leave the investigation of
+cross-platform predictive auto-tuning for future work.
 
@@
 
-\begin{figure*}[!ht]
+\begin{figure*}[!t]
 \centering
@@
 
-Figure~\ref{fig:allstars} shows the effectiveness of empirical auto-tuning is in this setting.
-Taking the GTX 295 as an example,
-and averaging across the range of problem configurations in our study,
-we find that empirically auto-tuned implementations are on average about
-50\% faster than the reference implementation.
-The reference in turn, is about 50\% faster than implementations that were empirically auto-tuned for a [randomly chosen] different argument configuration.
-This shows that it is genereally not enough to auto-tune for particular argument configurations;
-instead it is important to choose the right kernel for the job for each unique argument configuration (input-dependent auto-tuning).
-\begin{figure}
-\centering
-\includegraphics[scale=.42]{allstars_mixup_295.pdf}
-% XXX: NPTODO: ERROR BARS
-% XXX: NPTODO: fix fonts for better proportions
-\caption{Different arguments call for different kernels:
-{\em left} is the average argument-specific empirical
-auto-tuned performance across 100 random argument configurations,
-{\em middle} is the average speed of our reference implementation,
-{\em right} is the average speed of kernels auto-tuned for different argument configurations than the one being tested.
-For 38/100 random confurations, the kernel auto-tuned for another problem could not even run on the GTX 295 hardware.
-These points contributed a speed of 0, bringing down the average much lower than the reference.
-Good performance across a variety of inputs requires {\em input-dependent auto-tuning}.
-}
-\label{fig:allstars}
-\end{figure}
+Figure~\ref{fig:allstars} shows the effectiveness of empirical auto-tuning is
+in this setting.  Taking the GTX 295 as an example, and averaging across the
+range of problem configurations in our study, we find that empirically
+auto-tuned implementations are on average about 50\% faster than the reference
+implementation.  The reference in turn, is about 50\% faster than
+implementations that were empirically auto-tuned for a \emph{randomly chosen}
+different argument configuration.  This shows that it is genereally not enough
+to auto-tune for particular argument configurations; instead it is important to
+choose the right kernel for the job for each unique argument configuration
+(input-dependent auto-tuning).
 
-Comparing the {\em grid} to {\em HC25}, {\em HC50} and {\em HC75} we found very little difference in performance.
-The HC25 was slightly poorer, but the grid, HC50, and HC75 algorithms delivered similar average results.
-None of the algorithms was strictly better than the others.
-In our predictive auto-tuning experiments we used HC75.
+Comparing the {\em grid} to {\em HC25}, {\em HC50} and {\em HC75} we found very
+little difference in performance.  The HC25 was slightly poorer, but the grid,
+       HC50, and HC75 algorithms delivered similar average results. None of the
+       algorithms was strictly better than the others.  In our predictive
+       auto-tuning experiments we used HC75.
 
@@
 
-\begin{figure}
+\begin{figure}[!t]
+\centering
+\includegraphics[width=\linewidth]{illustrator/fig_allstars_mixup_580_R1.pdf}
+\caption{Different arguments call for different kernels:
+{\em left} is the average argument-specific empirical
+auto-tuned performance across 100 random argument configurations,
+{\em middle} is the average speed of our reference implementation,
+{\em right} is the average speed of kernels auto-tuned for different argument
+configurations than the one being tested.  For 38/100 random confurations, the
+kernel auto-tuned for another problem could not even run on the GTX 295
+hardware.  These points contributed a speed of 0, bringing down the average
+much lower than the reference.  Good performance across a variety of inputs
+requires {\em input-dependent auto-tuning}.
+}
+\label{fig:allstars}
+\end{figure}
+
+\begin{figure}[!t]
 \centering

### 2011-12-14 x_allstars_mixup.sh

@@ -1,5 +1,10 @@
 #!/bin/bash
 
-python fig.py allstars_mixup timings/munctional0/timing1_580_big.pkl timings/munctional0/timing1_580_big.pkl_allstars_mixup_timings
+machine=munctional0
+what=580
+#machine=vader
+#what=295
 
-mv -vf fig_allstars_mixup_580.pdf paper
+python fig.py allstars_mixup timings/${machine}/timing1_${what}_big.pkl timings/${machine}/timing1_${what}_big.pkl_allstars_mixup_timings
+
+mv -vf fig_allstars_mixup_${what}.pdf paper

### 2011-12-14 paper/draft1.tex

@@
 a hand-chosen grid provided by the authors of the kernel used in our experiments \citep{pinto+cox:2011gcg}.
-The {\em hill-climing} (HC) search algorithm starts from the reference
+The {\em hill-climbing} (HC) search algorithm starts from the reference
 implementation and resamples each of the parameters
@@
 On each hill-climbing iteration, if the speed of the newly sampled point is greater than the previous point, then it becomes the current point.
-We show results for search variants HC25, HC50, and HC75, which correspond do hill-climbin for 25, 50, and 75 iterations respectively.
+We show results for search variants HC25, HC50, and HC75, which correspond do hill-climbing for 25, 50, and 75 iterations respectively.
 The {\em grid} algorithm is specific to the kernel used in our case study,
@@
 Each thread computes $4 \times n\_output\_4s$ channels for some column and row of $z$.
-Each block of threads computes $4 \times n\_output\_4s$ channels for a subrectangle of the output image ($z$).
+Each block of threads computes $4 \times n\_output\_4s$ channels for a sub-rectangle of the output image ($z$).
 When there are more than $4 \times n\_output\_4s$ channels in $z$, or if the
@@
 pad\_shared $True$, 
-text1dfetch $True$,
-maxrregcount $\infty$,
+use\_text1d $True$,
+maxrreg $\infty$,
 and fast\_math $False$.
@@
 Recall from the introduction (Eq.~\ref{eq:condopt}) that auto-tuning can be seen
-as a conditional optimization problem in whic we seek an implementation ($b \in \mathcal{B}$) that minimizes runtime or some other scalar-valued cost function
+as a conditional optimization problem in which we seek an implementation ($b \in \mathcal{B}$) that minimizes runtime or some other scalar-valued cost function
 for given arguments ($a \in \mathcal{A}$) on a particular platform ($c \in \mathcal{C}$).
@@
 Models trained on as few as 25 configurations outperformed the reference.
-Models trained on 50 or more configurations rivaled empirical auto-tuning.
+Models trained on 50 or more configurations rivalled empirical auto-tuning.
 Moderate values of $\zeta$ between $\log(0.25)$ and $\log(0.75)$ are best.

### 2011-12-14 paper/draft1.tex
na
spelling

@@
 
-For the implementation features $b$, we directly used the integer and binary
-values (block\_w, block\_h, etc.) that parametrized the kernels.  We did not
-use platform features ($c$) in our experiments. We leave the investigation of
-cross-platform predictive auto-tuning for future work.
+For the implementation features $b$, we directly used the integer and binary values (block\_w, block\_h, etc.) that parametrized the kernels.
+We did not use platform features ($c$) in our experiments. We leave the investigation
+of cross-platform predictive auto-tuning for future work.
 
@@
 
-\begin{figure*}[!t]
+\begin{figure*}[!ht]
 \centering
@@
 
-Figure~\ref{fig:allstars} shows the effectiveness of empirical auto-tuning is
-in this setting.  Taking the GTX 295 as an example, and averaging across the
-range of problem configurations in our study, we find that empirically
-auto-tuned implementations are on average about 50\% faster than the reference
-implementation.  The reference in turn, is about 50\% faster than
-implementations that were empirically auto-tuned for a \emph{randomly chosen}
-different argument configuration.  This shows that it is genereally not enough
-to auto-tune for particular argument configurations; instead it is important to
-choose the right kernel for the job for each unique argument configuration
-(input-dependent auto-tuning).
+Figure~\ref{fig:allstars} shows the effectiveness of empirical auto-tuning is in this setting.
+Taking the GTX 295 as an example,
+and averaging across the range of problem configurations in our study,
+we find that empirically auto-tuned implementations are on average about
+50\% faster than the reference implementation.
+The reference in turn, is about 50\% faster than implementations that were empirically auto-tuned for a [randomly chosen] different argument configuration.
+This shows that it is generally not enough to auto-tune for particular argument configurations;
+instead it is important to choose the right kernel for the job for each unique argument configuration (input-dependent auto-tuning).
+\begin{figure}
+\centering
+\includegraphics[scale=.42]{allstars_mixup_295.pdf}
+% XXX: NPTODO: ERROR BARS
+% XXX: NPTODO: fix fonts for better proportions
+\caption{Different arguments call for different kernels:
+{\em left} is the average argument-specific empirical
+auto-tuned performance across 100 random argument configurations,
+{\em middle} is the average speed of our reference implementation,
+{\em right} is the average speed of kernels auto-tuned for different argument configurations than the one being tested.
+For 38/100 random configurations, the kernel auto-tuned for another problem could not even run on the GTX 295 hardware.
+These points contributed a speed of 0, bringing down the average much lower than the reference.
+Good performance across a variety of inputs requires {\em input-dependent auto-tuning}.
+}
+\label{fig:allstars}
+\end{figure}
 
-Comparing the {\em grid} to {\em HC25}, {\em HC50} and {\em HC75} we found very
-little difference in performance.  The HC25 was slightly poorer, but the grid,
-HC50, and HC75 algorithms delivered similar average results. None of the
-algorithms was strictly better than the others.  In our predictive
-auto-tuning experiments we used HC75.
+Comparing the {\em grid} to {\em HC25}, {\em HC50} and {\em HC75} we found very little difference in performance.
+The HC25 was slightly poorer, but the grid, HC50, and HC75 algorithms delivered similar average results.
+None of the algorithms was strictly better than the others.
+In our predictive auto-tuning experiments we used HC75.
 
@@
 
-\begin{figure}[!t]
-\centering
-\includegraphics[width=\linewidth]{illustrator/fig_allstars_mixup_580_R1.pdf}
-\caption{Different arguments call for different kernels:
-{\em left} is the average argument-specific empirical
-auto-tuned performance across 100 random argument configurations,
-{\em middle} is the average speed of our reference implementation,
-{\em right} is the average speed of kernels auto-tuned for different argument
-configurations than the one being tested.  For 38/100 random confurations, the
-kernel auto-tuned for another problem could not even run on the GTX 295
-hardware.  These points contributed a speed of 0, bringing down the average
-much lower than the reference.  Good performance across a variety of inputs
-requires {\em input-dependent auto-tuning}.
-}
-\label{fig:allstars}
-\end{figure}
-
-\begin{figure}[!t]
+\begin{figure}
 \centering

### 2011-12-14 paper/draft1.tex
na

--- 2011-12-14 paper/draft1.tex
@@
 
-% TODO: proof read for GFLOP vs. GFLOPS -> we want GFLOP
+\title{Machine Learning for Predictive Auto-Tuning with Boosted Regression Trees}
 
-%XXX:  Machine Learning for Auto-tuning: A Data-Driven Approach...} ???
-%\title{Boosted Regression Trees for Data-Driven Auto-Tuning}
-\title{Boosted Regression Trees for Predictive Auto-Tuning}
-% NP: I like this title !
-%XXX:  Machine Learning for Auto-tuning: A Data-Driven Approach...} ???
+% XXX: fix fonts in figures
 
@@
 
+% UPDATE ABSTRCT.TXT IF YOU CHANGE THIS!!! (abstract.txt)
 Auto-tuning techniques are widely used and effective for optimizing a
@@
 variety of problem types, kernel types, and platforms.
+% UPDATE ABSTRCT.TXT IF YOU CHANGE THIS!!! (abstract.txt)
 
@@
 \citep{clint2001automated, bilmes1997optimizing}, sparse linear algebra
-\citep{vuduc2005oski}, signal processing \citep{frigo1998fftw}, sorting
+\citep{vuduc2005oski}, signal processing \citep{frigo2005design}, sorting
 \citep{li2004dynamically}, general stencil operations \citep{kamil2010auto}, etc.
@@
 
-Figure~\ref{fig:fig_gflop_scatter} (b) and (c) show that the implementation speeds found by predictive auto-tuning correlates very well with the speed of empirical auto-tuning across all of the devices we tested: GTX 580, GTX 480, GTX 295, Tesla C2070, and Tesla C1060 which span three generations of CUDA platforms.
+Figure~\ref{fig:fig_gflop_scatter} (b) and (c) show that the implementation
+speeds found by predictive auto-tuning correlates very well with the speed of
+empirical auto-tuning across all of the devices we tested: GTX 580, GTX 480,
+GTX 295, Tesla C2070, and Tesla C1060 which span three generations of
+CUDA platforms.
 
@@
 \centering
-\includegraphics[scale=.42]{fig_ntrain_munctional0_580.pdf} % on vader
+\includegraphics[width=\linewidth]{illustrator/fig_ntrain_munctional0_580.pdf} % on vader
 \caption{The effect of the invalid configuration score ($\zeta$) and training set

### 2011-12-14 paper/draft1.tex
m

@@
 \title{Machine Learning for Predictive Auto-Tuning with Boosted Regression Trees}
-
-% XXX: fix fonts in figures
 
@@
 \citep{clint2001automated, bilmes1997optimizing}, sparse linear algebra
-\citep{vuduc2005oski}, signal processing \citep{frigo2005design}, sorting
+\citep{vuduc2005oski}, signal processing \citep{frigo1998fftw}, sorting
 \citep{li2004dynamically}, general stencil operations \citep{kamil2010auto}, etc.
@@
 \centering
-\includegraphics[width=\linewidth]{illustrator/fig_ntrain_munctional0_580.pdf} % on vader
+\includegraphics[scale=.42]{fig_ntrain_munctional0_580.pdf} % on vader
 \caption{The effect of the invalid configuration score ($\zeta$) and training set

### 2011-12-14 paper/draft1.tex
Aspell pass

--- 2011-12-14 paper/draft1.tex
@@
 In this paper, we propose to \emph{learn the model} using non-linear regression
-modelling techniques instead of constructing a model manually. By learning the model, one can
+modeling techniques instead of constructing a model manually. By learning the model, one can
 hope to achieve elements of the best of both approaches: the search speed of model-based auto-tuning with the broad applicability
@@
 {\em right} is the average speed of kernels auto-tuned for different argument
-configurations than the one being tested.  For 38/100 random confurations, the
+configurations than the one being tested.  For 38/100 random configurations, the
 kernel auto-tuned for another problem could not even run on the GTX 295
@@
 Models trained on as few as 25 configurations outperformed the reference.
-Models trained on 50 or more configurations rivalled empirical auto-tuning.
+Models trained on 50 or more configurations rivaled empirical auto-tuning.
 Moderate values of $\zeta$ between $\log(0.25)$ and $\log(0.75)$ are best.
@@
 evaluations.  This approach allows significant flexibility to navigate
-tradeoffs between offline and run-time costs, and final auto-tuning
+trade-offs between offline and run-time costs, and final auto-tuning
 performance.  Importantly, this method treats kernels as black-boxes, allowing
@@
 machine-learning-based techniques described here provide a middle ground, where
-a library implementor can simply create an instrumented kernel, and allow
+a library developer can simply create an instrumented kernel, and allow
 generic tools to automatically generate appropriate auto-tuned implementations,

### 2011-12-14 paper/draft1.tex
labels and refs

--- 2011-12-14 paper/draft1.tex
@@
 c) predictive auto-tuning.
-%Section~\ref{sec:discussion}.
-Section~\ref{sec:previous} summarizes previous and related work involving  machine learning with performance auto-tuning.
-Section~\ref{sec:conclusion} summarizes our findings and outlines directions for future work.
+Section~\ref{sec:discuss} provides discussion of our results, and
+Section~\ref{sec:fw} outlines directions for future work.
 
@@
 
-Once an accurate regression model has been fit to the data, it remains to be optimized for novel arguments (Fig.~\ref{fig:autotune_general}, bottom, step 4).
+Once an accurate regression model has been fit to the data, it remains to be optimized for novel arguments (Fig.~\ref{fig:general_autotune}, bottom, step 4).
 An exhaustive search is the most reliable if it can be afforded, but in our experiments
@@
 \citet{pinto+cox:2011gcg} in Python using Cheetah for string
-processing~\citep{cheetah} and PyCUDA~\citep{klockner2011pycuda} for dynamic
+processing and PyCUDA~\citep{klockner2011pycuda} for dynamic
 kernel compilation and interfacing with CUDA.
@@
 \section{Experimental Setup}
+\label{sec:exp}
 
-Recall from the introduction (Eq.~\ref{eq:condopt}) that auto-tuning can be seen
+Recall from the introduction (Eq.~\ref{fig:general_autotune}) that auto-tuning can be seen
 as a conditional optimization problem in which we seek an implementation ($b \in \mathcal{B}$) that minimizes runtime or some other scalar-valued cost function
@@
 \section{Discussion}
+\label{sec:discuss}
 
@@
 \section{Future Work}
+\label{sec:fw}
 

### 2011-12-14 paper/draft1.tex

@@
 \title{Machine Learning for Predictive Auto-Tuning with Boosted Regression Trees}
+
+% XXX: fix fonts in figures
 

### 2011-12-14 paper/draft1.tex
Minor edits

--- 2011-12-14 paper/draft1.tex
@@
 approaches have emerged in the extensive literature covering the subject (see
-surveys in
+surveys in e.g.
 \citep{vuduc2001statistical, demmel2005self, vuduc2005oski, williams2008auto,
@@
 
-To ameliorate this weakness, it is intuitively appealing to combine the two
+To addess this weakness, it is intuitively appealing to combine the two
 approaches by first constraining the search space with an analytical model and

### 2011-12-14 paper/draft1.tex

@@
 
-In terms of floating point operations, a filterbank correlation requires the
+In terms of floating point operations (FLOP), a filterbank correlation requires the
 inner sums to be computed for each output pixel, yielding the quantity in
@@
 \begin{equation}
-\mathrm{FLOPS} = 2  F  H  W  D  (R - H + 1)( C- W + 1)
+\mathrm{FLOP} = 2  F  H  W  D  (R - H + 1)( C- W + 1)
 \label{eq:flops}
@@
 includes 1600 argument combinations, but we restricted our experiments to correlations that
-represented between 1 and 50 gigaflops of arithmetic.
+represented between 1 and 50 gigaflops (GFLOP) of arithmetic.
 Smaller problems do not fully utilize GPU hardware and are handled equally well by many kernel settings.
@@
 inefficiency in implementing them via multiple calls with smaller images and fewer filters.
-With the experiments searched an argument space included 602 configurations with between 1 and 50 gigaflops.
+With the experiments searched an argument space included 602 configurations with between 1 and 50 GFLOP.
 

### 2011-12-14 paper/draft1.tex

--- 2011-12-14 paper/draft1.tex
@@
 approaches have emerged in the extensive literature covering the subject (see
-surveys in e.g.
+surveys in
 \citep{vuduc2001statistical, demmel2005self, vuduc2005oski, williams2008auto,
@@
 
-To addess this weakness, it is intuitively appealing to combine the two
+To ameliorate this weakness, it is intuitively appealing to combine the two
 approaches by first constraining the search space with an analytical model and

### 2011-12-14 paper/draft1.tex
na

--- 2011-12-14 paper/draft1.tex
@@
 The hypothetical auto-tuning routine described at the top of Figure~\ref{fig:general_autotune}
-might take many minutes or hours to perform the argmin at step 3 (during which time it computes the desired result many times!)
+might take many minutes or hours to perform the $argmin$ at step 3 (during which it computes the desired result many times!)
 so it would not be suitable for a normal library subroutine implementation.
@@
 if only there were a fast (even approximate) surrogate for the costly $MeasureTime(\cdot)$ function,
-then the argmin could be done in a fraction of a second and the routine could be used normally (Figure~\ref{fig:general_autotune}, bottom).
+then the $argmin$ could be done in a fraction of a second and the routine could be used normally (Figure~\ref{fig:general_autotune}, bottom).
 
@@
 
-The heart of predictive auto-tuning is a regression model that acts as surrogate
+The heart of our predictive auto-tuning is a regression model that acts as surrogate
 for a hand-crafted hardware model or empirical timing estimates.

### 2011-12-14 paper/draft1.tex
na

@@
 \cite{pinto+cox:2011gcg} recommends as a reference implementation: block\_w 8, 
-block\_h = 8, n\_filter\_rows 1, n\_output\_4s all, spill $False$, imul\_fast $True$,
-pad\_shared $True$, 
-use\_text1d $True$,
-maxrreg $\infty$,
-and fast\_math $False$.
-This reference implementation was chosen manually based on good performance across a range of platforms from older-generation
-cards such as the
-8600GT all the way to
-current-generation flagship cards such as the GTX 580 and C2070.
+block\_h = 8, n\_filter\_rows = 1, n\_output\_4s = all, spill = $False$, imul\_fast = $True$,
+pad\_shared = $True$, 
+use\_text1d = $True$,
+maxrreg = $\infty$,
+and fast\_math = $False$.
+This reference implementation was chosen manually based on good performance
+across a range of platforms from older-generation cards such as the 8600GT all
+the way to current-generation flagship cards such as the GTX 580 and C2070.
 Given that parameters were hand-chosen for the reference kernel, no claims are
 made as to the optimality nor universality of this reference (indeed, different
-programmers would undoubtedly arrive at different results).  We use this
-kernel configuration as a reasonable indicator of typical performance
+programmers would undoubtedly arrive at different results).  We use
+this kernel configuration as a reasonable indicator of typical performance
 made possible by ad hoc experimentation with parameters.

### 2011-12-14 paper/draft1.tex
na

@@
 Figure~\ref{fig:allstars} shows the effectiveness of empirical auto-tuning is
-in this setting.  Taking the GTX 295 as an example, and averaging across the
+in this setting.  Taking the GTX 580 as an example, and averaging across the
 range of problem configurations in our study, we find that empirically

### 2011-12-14 paper/draft1.tex

@@
 % UPDATE ABSTRCT.TXT IF YOU CHANGE THIS!!! (abstract.txt)
-Auto-tuning techniques are widely used and effective for optimizing a
-parametrized GPU code template for a particular computation on particular
-hardware.
-However, a drawback of this approach is that thorough or exhaustive
-auto-tuning requires compiling many kernels and calling each one many times,
-and this process is slow.  Furthermore, it is desirable to provide users
-with unified library abstraction boundaries for operations such as image
-filtering and matrix multiplication, even those these operations actually
-correspond to a large set of potential problem configurations with a wide
-variety of memory access patterns and computational bottlenecks.  How can we
-draw on data from previous empirical auto-tuning of related problems on related
-hardware to make a just-in-time implementation decision for a novel problem?
-This paper presents a machine learning approach to auto-tuning, in which
-features of the current hardware platform, the kernel configuration and the
-problem instance are passed to a regression model (boosted regression trees)
-which predicts how much faster this kernel will be than a reference
-baseline.  Combinatorial optimization strategies for auto-tuning that would
-normally require evaluating large number of kernel configurations on the real
-hardware, are made orders of magnitude faster by evaluating the surrogate
-regression model instead.  We validate our approach using the filterbank
+The rapidly evolving landscape of multicore architectures makes
+the construction of efficient libraries a daunting task.
+A family of methods known collectively as ``auto-tuning'' has emerged to
+address this challenge.  Two major approaches to auto-tuning are
+{\em empirical} and {\em model-based}:
+empirical auto-tuning is a generic but slow approach that works by measuring runtimes of
+candidate implementations, model-based auto-tuning predicts those runtimes using simplified hardware abstractions designed by hand.
+We show that machine learning methods for non-linear regression can be used
+to estimate timing models from data, capturing the best of both approaches.
+A statistically-derived model offers the speed of a model-based approach,
+with the generality and simplicity of empirical auto-tuning.
+We validate our approach using the filterbank
 correlation kernel described in \citet{pinto+cox:2011gcg}, where we find that

### 2011-12-14 paper/draft1.tex
na

@@
 the performance of predictive auto-tuning approaches that of empirical auto-tuning
-despite taking only 0.1 seconds instead of 2 minutes.
+despite taking only 0.1 seconds instead of minutes.
 (b) and (c) For a wide range of devices, implementation speeds found by predictive auto-tuning correlate tightly with
@@
 speeds found by predictive auto-tuning correlates very well with the speed of
-empirical auto-tuning across all of the devices we tested: GTX 580, GTX 480,
-GTX 295, Tesla C2070, and Tesla C1060 which span three generations of
-CUDA platforms.
+empirical auto-tuning across all of the devices we tested: GeForce GTX 295,
+Tesla C1060, GeForce GTX 580, GeForce GTX 480 and Tesla C2070 which
+span two generations of NVIDIA CUDA platforms.
 
@@
 dealt with this scenario by backtracking through the various best-estimates
-discovered during the hill-climbing search.  Some invalid kernels can only be
+found during the hill-climbing search.  Some invalid kernels can only be
 discovered after compiling code and attempting to run the compiled code, so in
@@
 hill-climbing search procedures were used as training examples, so the training
-set sizes ranged from an average of 1480 (10 problem configurations) to 29 600
+set sizes ranged from an average of 1,480 (10 problem configurations) to 29,600
 (200 problem configurations).
@@
 {\em right} is the average speed of kernels auto-tuned for different argument
-configurations than the one being tested.  For 38/100 random configurations, the
+configurations than the one being tested.  For 38 out of 100 random configurations, the
 kernel auto-tuned for another problem could not even run on the GTX 295

### 2011-12-14 paper/draft1.tex

@@
 automatic methods known collectively as ``auto-tuning.'' Two major auto-tuning
-approaches have emerged in the extensive literature covering the subject (see
+approaches have emer-ged in the extensive literature covering the subject (see
 surveys in e.g.
@@
 
-\cite{pinto+cox:2011gcg} recommends as a reference implementation: block\_w 8, 
-block\_h = 8, n\_filter\_rows = 1, n\_output\_4s = all, spill = $False$, imul\_fast = $True$,
-pad\_shared = $True$, 
+\cite{pinto+cox:2011gcg} recommends as a reference implementation:
+block\_w = 8, 
+block\_h = 8,
+n\_filter\_rows = 1,
+n\_output\_4s = all,
+spill = $False$,
+imul\_fast = $True$,
+pad\_sha-red = $True$, %%% XXX: hypenation by hand
 use\_text1d = $True$,
@@
 For the implementation features $b$, we directly used the integer and binary
-values (block\_w, block\_h, etc.) that parametrized the kernels.  We did not
+values (block\_w, block\_h, etc.) that paramet-rized the kernels.  We did not
 use platform features ($c$) in our experiments. We leave the investigation of
@@
 Figure~\ref{fig:fig_gflop_scatter} shows how accurate predictive auto-tuning is compared with empirical auto-tuning.
-The training set $(\mathcal{X}, \mathcal{Y})$ for the regression model comprised
-all of the $<(a,b,c), y>$ pairs observed during grid search and hill-climbing search.
+The training set $(\mathcal{X}$, $\mathcal{Y})$ for the regression model comprised
+all of the ($(a,b,c)$, $y$) pairs observed during grid search and hill-climbing search.
 So for each training argument configuration $a,c$ there were 148 different values of $b$ and thus 148 training points.

### 2011-12-14 paper/draft1.tex
abstract

--- 2011-12-14 paper/draft1.tex
@@
 The hypothetical auto-tuning routine described at the top of Figure~\ref{fig:general_autotune}
-might take many minutes or hours to perform the $argmin$ at step 3 (during which it computes the desired result many times!)
+might take many minutes or hours to perform the argmin at step 3 (during which time it computes the desired result many times!)
 so it would not be suitable for a normal library subroutine implementation.
@@
 if only there were a fast (even approximate) surrogate for the costly $MeasureTime(\cdot)$ function,
-then the $argmin$ could be done in a fraction of a second and the routine could be used normally (Figure~\ref{fig:general_autotune}, bottom).
+then the argmin could be done in a fraction of a second and the routine could be used normally (Figure~\ref{fig:general_autotune}, bottom).
 
@@
 
-The heart of our predictive auto-tuning is a regression model that acts as surrogate
+The heart of predictive auto-tuning is a regression model that acts as surrogate
 for a hand-crafted hardware model or empirical timing estimates.
@@
 \cite{pinto+cox:2011gcg} recommends as a reference implementation: block\_w 8, 
-block\_h = 8, n\_filter\_rows = 1, n\_output\_4s = all, spill = $False$, imul\_fast = $True$,
-pad\_shared = $True$, 
-use\_text1d = $True$,
-maxrreg = $\infty$,
-and fast\_math = $False$.
-This reference implementation was chosen manually based on good performance
-across a range of platforms from older-generation cards such as the 8600GT all
-the way to current-generation flagship cards such as the GTX 580 and C2070.
+block\_h = 8, n\_filter\_rows 1, n\_output\_4s all, spill $False$, imul\_fast $True$,
+pad\_shared $True$, 
+use\_text1d $True$,
+maxrreg $\infty$,
+and fast\_math $False$.
+This reference implementation was chosen manually based on good performance across a range of platforms from older-generation
+cards such as the
+8600GT all the way to
+current-generation flagship cards such as the GTX 580 and C2070.
 Given that parameters were hand-chosen for the reference kernel, no claims are
 made as to the optimality nor universality of this reference (indeed, different
-programmers would undoubtedly arrive at different results).  We use
-this kernel configuration as a reasonable indicator of typical performance
+programmers would undoubtedly arrive at different results).  We use this
+kernel configuration as a reasonable indicator of typical performance
 made possible by ad hoc experimentation with parameters.
@@
 Figure~\ref{fig:allstars} shows the effectiveness of empirical auto-tuning is
-in this setting.  Taking the GTX 580 as an example, and averaging across the
+in this setting.  Taking the GTX 295 as an example, and averaging across the
 range of problem configurations in our study, we find that empirically

### 2011-12-14 paper/draft1.tex

--- 2011-12-14 paper/draft1.tex
@@
 the performance of predictive auto-tuning approaches that of empirical auto-tuning
-despite taking only 0.1 seconds instead of minutes.
+despite taking only 0.1 seconds instead of 2 minutes.
 (b) and (c) For a wide range of devices, implementation speeds found by predictive auto-tuning correlate tightly with
@@
 speeds found by predictive auto-tuning correlates very well with the speed of
-empirical auto-tuning across all of the devices we tested: GeForce GTX 295,
-Tesla C1060, GeForce GTX 580, GeForce GTX 480 and Tesla C2070 which
-span two generations of NVIDIA CUDA platforms.
+empirical auto-tuning across all of the devices we tested: GTX 580, GTX 480,
+GTX 295, Tesla C2070, and Tesla C1060 which span three generations of
+CUDA platforms.
 
@@
 dealt with this scenario by backtracking through the various best-estimates
-found during the hill-climbing search.  Some invalid kernels can only be
+discovered during the hill-climbing search.  Some invalid kernels can only be
 discovered after compiling code and attempting to run the compiled code, so in
@@
 hill-climbing search procedures were used as training examples, so the training
-set sizes ranged from an average of 1,480 (10 problem configurations) to 29,600
+set sizes ranged from an average of 1480 (10 problem configurations) to 29 600
 (200 problem configurations).
@@
 {\em right} is the average speed of kernels auto-tuned for different argument
-configurations than the one being tested.  For 38 out of 100 random configurations, the
+configurations than the one being tested.  For 38/100 random configurations, the
 kernel auto-tuned for another problem could not even run on the GTX 295

### 2011-12-14 paper/draft1.tex
Pure awesonesss

--- 2011-12-14 paper/draft1.tex
@@
 empirical auto-tuning is a generic but slow approach that works by measuring runtimes of
-candidate implementations, model-based auto-tuning predicts those runtimes using simplified hardware abstractions designed by hand.
+candidate implementations, model-based auto-tuning predicts those runtimes using simplified abstractions designed by hand.
 We show that machine learning methods for non-linear regression can be used
@@
 might take many minutes or hours to perform the $argmin$ at step 3 (during which it computes the desired result many times!)
-so it would not be suitable for a normal library subroutine implementation.
+so it would not be suitable for a normal library routine implementation.
 However, the form of the auto-tuning routine suggests the potential for enormous acceleration:
@@
 % say what we said
-In this paper, we've demonstrated a boosted regression tree-based auto-tuning
+In this paper, we have demonstrated a boosted regression tree-based auto-tuning
 method, wherein empirical performance data is used to train a machine learning

### 2011-12-14 paper/draft1.tex
marked manual hyphen

--- 2011-12-14 paper/draft1.tex
@@
 empirical auto-tuning is a generic but slow approach that works by measuring runtimes of
-candidate implementations, model-based auto-tuning predicts those runtimes using simplified abstractions designed by hand.
+candidate implementations, model-based auto-tuning predicts those runtimes using simplified hardware abstractions designed by hand.
 We show that machine learning methods for non-linear regression can be used
@@
 might take many minutes or hours to perform the $argmin$ at step 3 (during which it computes the desired result many times!)
-so it would not be suitable for a normal library routine implementation.
+so it would not be suitable for a normal library subroutine implementation.
 However, the form of the auto-tuning routine suggests the potential for enormous acceleration:
@@
 For the implementation features $b$, we directly used the integer and binary
+% XXX: MANUAL HYPHENATION
 values (block\_w, block\_h, etc.) that paramet-rized the kernels.  We did not
@@
 % say what we said
-In this paper, we have demonstrated a boosted regression tree-based auto-tuning
+In this paper, we've demonstrated a boosted regression tree-based auto-tuning
 method, wherein empirical performance data is used to train a machine learning

### 2012-03-01 paper/final.tex
Starting to work on the final version

--- 2011-12-14 paper/draft1.tex
@@
 \author{
-%\alignauthor First Last\titlenote{}\\
-\alignauthor First Last\\
-\affaddr{Affiliation line 1}\\
-\affaddr{Affiliation line 2}\\
-\email{anon@mail.com}
+\alignauthor James Bergstra\\
+\affaddr{\small{Rowland Institute at Harvard}}\\
+\affaddr{\small{Cambridge MA, US}}\\
+\email{\emph{\small{bergstra@rowland.harvard.edu}}}
 % 2nd author
-%\alignauthor First Last\titlenote{None}\\
-\alignauthor First Last\\
-\affaddr{Affiliation line 1}\\
-\affaddr{Affiliation line 2}\\
-\email{anon@mail.com}
+\alignauthor Nicolas Pinto\\
+\affaddr{\small{Rowland Institute at Harvard}}\\
+\affaddr{\small{McGovern Institute at MIT}}\\
+\affaddr{\small{Cambridge MA, USA}}\\
+\email{\emph{\small{pinto@rowland.harvard.edu}}}
 % 3rd author
-%\alignauthor First Last\titlenote{None}\\
-\alignauthor First Last\\
-\affaddr{Affiliation line 1}\\
-\affaddr{Affiliation line 2}\\
-\email{anon@mail.com}
+\alignauthor David Cox\\
+\affaddr{\small{Rowland Institute at Harvard}}\\
+\affaddr{\small{Cambridge MA, USA}}\\
+\email{\emph{\small{cox@harvard.rowland.edu}}}
 }
@@
 For the implementation features $b$, we directly used the integer and binary
-% XXX: MANUAL HYPHENATION
 values (block\_w, block\_h, etc.) that paramet-rized the kernels.  We did not
@@
 \centering
-\includegraphics[width=\linewidth]{illustrator/fig_allstars_mixup_580_R1.pdf}
+\includegraphics[width=\linewidth]{illustrator/fig_allstars_mixup_580_R2.pdf}
 \caption{Different arguments call for different kernels:

### 2012-03-02 paper/final.tex
Main figure fixes
Fix fig5
Add some more future work

@@
 
-The paper is organized as follows:
-Section~\ref{sec:brtree} describes the boosted regression tree model and the procedure for fitting it to empirical timing data.
-Section~\ref{sec:kernel} describes the sort of kernel we employ for our benchmarking,
-Section~\ref{sec:exp} presents the results of our benchmarking experiments, which
-compare a reference implementation to a) empirical auto-tuning over a domain-specific grid,
-b) empirical auto-tuning over a hill-climbing search, and
-c) predictive auto-tuning.
-Section~\ref{sec:discuss} provides discussion of our results, and
-Section~\ref{sec:fw} outlines directions for future work.
+The paper is organized as follows: Section~\ref{sec:predautotuning}
+describes the boosted regression tree model and the procedure for
+fitting it to empirical timing data.  Section~\ref{sec:fbcorr} details
+the data-parallel image processing algorithm we use to illustrate our
+auto-tuning framework.  Section~\ref{sec:kernel} describes the sort of
+kernel we employ for our benchmarking. Section~\ref{sec:exp} and
+\ref{sec:results} present the results of our benchmarking experiments,
+which compare a reference implementation to a) empirical auto-tuning
+over a domain-specific grid, b) empirical auto-tuning over
+a hill-climbing search, and c) predictive auto-tuning.
+Section~\ref{sec:discuss} provides discussion and concluding remarks,
+while Section~\ref{sec:fw} outlines directions for future work.
 
@@
 \section{Predictive Auto-Tuning}
+\label{sec:predautotuning}
 
@@
 \section{Filterbank Correlation}
-
 \label{sec:fbcorr}
@@
 
-The high potential for arithmetic intensity makes the GPU an ideal platform for computing filterbank correlations,
-and and filterbank correlation is used extensively in image and video processing,
-where it is often a computational bottleneck.
-One might expect then, that it would be easy to implement a library providing
-this operation as a simple function that takes pointers and strides for $x$, $f$, and $z$ and performs the computation.
-However, as shown in \citet{pinto+cox:2011gcg} 
-it is challenging to provide an implementation
-or even an implementation strategy that provides satisfactory performance
-across the range of inputs (shapes, physical layouts) that occur in
-typical usage.
-\citet{kamil+etal:2009} summarize a related situation related to general stencil computations in their abstract:
-``Although the auto-tuning strategy has been successfully applied to libraries,
-generalized stencil kernels are not amenable to packaging as libraries.''
+The high potential for arithmetic intensity makes the GPU an ideal
+platform for computing filterbank correlations, and filterbank
+correlation is used extensively in image and video processing, where it
+is often a computational bottleneck.  One might expect then, that it
+would be easy to implement a library providing this operation as
+a simple function that takes pointers and strides for $x$, $f$, and $z$
+and performs the computation.  However, as shown in
+\citet{pinto+cox:2011gcg} it is challenging to provide an implementation
+or even an implementation strategy that provides satisfactory
+performance across the range of inputs (shapes, physical layouts) that
+occur in typical usage.  \citet{kamil+etal:2009} summarize a related
+situation related to general stencil computations in their abstract:
+``Although the auto-tuning strategy has been successfully applied to
+libraries, generalized stencil kernels are not amenable to packaging as
+libraries.''
 
@@
 \section{Results}
+\label{sec:results}
 
@@
 \centering
-\includegraphics[width=.9\linewidth]{illustrator/fig_main_R1.pdf}
+\includegraphics[width=.9\linewidth]{illustrator/fig_main_R2.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=\linewidth]{illustrator/fig_ntrain_munctional0_580.pdf} % on vader
+\includegraphics[width=\linewidth]{illustrator/fig_ntrain_munctional0_580_R2.pdf} % on vader
 \caption{The effect of the invalid configuration score ($\zeta$) and training set
@@
 
-While the present work serves as a basic demonstration of the value of using
-machine learning models to predict optimal implementation parameters, there are
-many avenues for taking these ideas further. A natural route to extend of our
-current approach is to include more features as inputs to the predictive model.
-In addition to further instrumentation of the kernel in question, input
-features could include a much broader range of hardware-related information,
-from more detailed information about device capabilities to micro-benchmarks
-\citep{wong2010demystifying}, and results from performance limiter analyses
-\citep{micikevicius2010analysis}. Such additional features would increase the
-predictive model's ability to adapt to a wide range of different kinds of
-hardware, including new devices not available at the time of kernel creation.
+While the present work serves as a basic demonstration of the value of
+using machine learning models to predict optimal implementation
+parameters on a specific application, there are many avenues for taking
+these ideas further.
 
-Another potentially interesting avenue of research is in interpreting the model
-learned by predictive auto-tuning.  While traditional model-based auto-tuning
-approaches by design assume a given model, and empirical auto-tuning approaches
-are completely model free, the predictive approach described here
-\emph{generates} a model from performance data.  Because this generated model
-can be interrogated by a variety of means, a significant opportunity exists to
-learn about the factors that drive the performance of a given kernel.  These
-insights can be used to further guide the development and instrumentation of
-the kernel, potentially yielding even greater gains.
+A natural route to extend of our current approach is to include more
+features as inputs to the predictive model.  In addition to further
+instrumentation of the kernel in question, input features could include
+a much broader range of hardware-related information, from more detailed
+information about device capabilities to micro-benchmarks
+\citep{wong2010demystifying}, and results from performance limiter
+analyses \citep{micikevicius2010analysis}. Such additional features
+would increase the predictive model's ability to adapt to a wide range
+of different kinds of hardware, including new devices not available at
+the time of kernel creation.
+
+Another potentially interesting avenue of research is in interpreting
+the model learned by predictive auto-tuning.  While traditional
+model-based auto-tuning approaches by design assume a given model, and
+empirical auto-tuning approaches are completely model free, the
+predictive approach described here \emph{generates} a model from
+performance data.  Because this generated model can be interrogated by
+a variety of means, a significant opportunity exists to learn about the
+factors that drive the performance of a given kernel.  These insights
+can be used to further guide the development and instrumentation of the
+kernel, potentially yielding even greater gains.
+
+Other directions will include the validation of our predictive
+auto-tuning framework to other domains and environments, in particular
+general data-parallel primitives and non data-parallel applications,
+other multi-core heterogeneous architectures, and software platforms
+such as OpenCL.
 

### 2012-03-02 paper/final.tex
Aspell check

@@
 automatic methods known collectively as ``auto-tuning.'' Two major auto-tuning
-approaches have emer-ged in the extensive literature covering the subject (see
+approaches have emerged in the extensive literature covering the subject (see
 surveys in e.g.
@@
 
-To addess this weakness, it is intuitively appealing to combine the two
+To address this weakness, it is intuitively appealing to combine the two
 approaches by first constraining the search space with an analytical model and
@@
 For the implementation features $b$, we directly used the integer and binary
-values (block\_w, block\_h, etc.) that paramet-rized the kernels.  We did not
+values (block\_w, block\_h, etc.) that paraded-raced the kernels.  We did not
 use platform features ($c$) in our experiments. We leave the investigation of

### 2012-03-07 paper/final.tex

@@
 general data-parallel primitives and non data-parallel applications,
-other multi-core heterogeneous architectures, and software platforms
+other multi-core architectures, and heterogeneous software platforms
 such as OpenCL.

### 2012-03-08 paper/draft1.tex
adding authors, and inpar12.cls file
a few paper edits

@@
-\documentclass{sig-alternate}
+\documentclass{inpar12}
 \usepackage{graphicx}
@@
 \author{
+\tiny
 %\alignauthor First Last\titlenote{}\\
-\alignauthor First Last\\
-\affaddr{Affiliation line 1}\\
-\affaddr{Affiliation line 2}\\
-\email{anon@mail.com}
+\alignauthor James Bergstra\\
+\affaddr{The Rowland Institute at Harvard}\\
+\affaddr{Harvard University}\\
+\affaddr{Cambridge, MA 02142, USA}\\
+\email{bergstra@rowland.harvard.edu}
 % 2nd author
 %\alignauthor First Last\titlenote{None}\\
-\alignauthor First Last\\
-\affaddr{Affiliation line 1}\\
-\affaddr{Affiliation line 2}\\
-\email{anon@mail.com}
+\alignauthor Nicolas Pinto\\
+\affaddr{Massachusetts Institute of Technology}\\
+\affaddr{Cambridge, MA 02139, USA}\\
+\email{pinto@mit.edu}
 % 3rd author
 %\alignauthor First Last\titlenote{None}\\
-\alignauthor First Last\\
-\affaddr{Affiliation line 1}\\
-\affaddr{Affiliation line 2}\\
-\email{anon@mail.com}
+\alignauthor David D. Cox\\
+\affaddr{The Rowland Institute at Harvard}\\
+\affaddr{Harvard University}\\
+\affaddr{Cambridge, MA 02142, USA}\\
+\email{cox@rowland.harvard.edu}
 }
@@
 facing a wide diversity of low-level architectural issues
-that must be carefully balanced if one is to write code that is both
+that must be carefully balanced if they are to write code that is both
 high-performance \emph{and} portable.
@@
 Compute Unified Device Architecture (CUDA) \citep{nickolls2008scalable,
-nvidia2011cuda}, many developers default to massive amounts of manual labor
-to optimize CUDA code to specific input domains. In addition, hand-tuning
+nvidia2011cuda}, many developers apply massive amounts of manual labor
+to optimize CUDA code to specific input domains. This hand-tuning
 rarely generalizes well to new hardware generations or different input
-domains, and it can also be error-prone or far from optimal. One of the
-reason is that kernels can produce staggeringly large optimization spaces
+domains, is prone to error, results in unmaintainable code, and does not even
+guarantee optimal performance.
+One of the
+reasons is that GPU kernels can yield staggeringly large optimization spaces
 \citep{datta2008stencil}. The problem is further compounded by the fact that
 these spaces can be highly discontinuous \citep{ryoo2008program}, difficult
-to explore, and quasi-optimal solutions lie at the edge of ``performance
+to explore, and optimal performance is often realized at the edge of ``performance
 cliffs'' induced by hard device-specific constraints (e.g. register file
@@
 
-In this paper, we propose to \emph{learn the model} using non-linear regression
+In this paper, we show that it is possible to \emph{learn the model} using non-linear regression
 modeling techniques instead of constructing a model manually. By learning the model, one can
@@
 for all input configurations. One option for dealing with these invalid configurations
-would be to simply omit them from $\mathcal{X}$ and $\mathcal{Y}$, but that would lead to a biased regression model.
+would be to simply omit them from $\mathcal{X}$ and $\mathcal{Y}$, but that
+would lead to a regression model that suggests invalid configurations.
 Instead, we chose to associate invalid $(a, b, c)$ tuples with a constant $y = \zeta$.
 It makes sense to choose $\zeta < 0$ so that invalid configurations are treated as being worse than the reference,
-but the question of how much worse these points should be is an empirical one.
-Our experiments evaluate $\log(0.01) \leq \zeta \leq \log(.99)$.
+but the question of how much worse is an empirical one.
+In our experiments, we compare $\zeta$ in the range $\log(0.01) \leq \zeta \leq \log(.99)$.
 
@@
 We did not attempt a systematic study of the effect of these variables on performance.
-We chose a maximum depth of 4,
-a minimum number of examples of 10,
-and 100 iterations of boosting.
+We chose a maximum depth of 4, so that each new tree would be a {\em weak}
+learner without too much capacity;
+a minimum number of examples per leaf of 10,
+so that the ensemble would resist overfitting;
+and 100 iterations of boosting, after which the residual error on training
+data appeared to have reached a minimum.  Test examples were not used to
+choose these values.
 
@@
 probability $0.25$ (keeping the current best setting with probability $0.75$).
-On each hill-climbing iteration, if the speed of the newly sampled point is greater than the previous point, then it becomes the current point.
+On each hill-climbing iteration, if the predicted speed of the newly sampled point is greater than the previous point, then it becomes the current point.
 We show results for search variants HC25, HC50, and HC75, which correspond do hill-climbing for 25, 50, and 75 iterations respectively.

### 2012-03-11 paper/final.tex
Add acknowledgements
transferring changes from draft1 -> final

@@
 for all input configurations. One option for dealing with these invalid configurations
-would be to simply omit them from $\mathcal{X}$ and $\mathcal{Y}$, but that would lead to a biased regression model.
+would be to simply omit them from $\mathcal{X}$ and $\mathcal{Y}$, but that
+would lead to a regression model that suggests invalid configurations.
 Instead, we chose to associate invalid $(a, b, c)$ tuples with a constant $y = \zeta$.
 It makes sense to choose $\zeta < 0$ so that invalid configurations are treated as being worse than the reference,
-but the question of how much worse these points should be is an empirical one.
-Our experiments evaluate $\log(0.01) \leq \zeta \leq \log(.99)$.
+but the question of how much worse is an empirical one.
+In our experiments we compare $\zeta$ in the range $\log(0.01) \leq \zeta \leq \log(.99)$.
 

### 2012-03-11 paper/final.tex
final touchup
final spellcheck

@@
 \affaddr{\small{Rowland Institute at Harvard}}\\
-\affaddr{\small{Cambridge MA, US}}\\
+\affaddr{\small{Harvard University}}\\
+\affaddr{\small{Cambridge MA, USA}}\\
 \email{\emph{\small{bergstra@rowland.harvard.edu}}}
@@
 \affaddr{\small{Rowland Institute at Harvard}}\\
+\affaddr{\small{Harvard University}}\\
 \affaddr{\small{Cambridge MA, USA}}\\
@@
 a minimum number of examples per leaf of 10,
-so that the ensemble would resist overfitting;
+so that the ensemble would resist over-fitting;
 and 100 iterations of boosting, after which the residual error on training
@@
 implementations that were empirically auto-tuned for a \emph{randomly chosen}
-different argument configuration.  This shows that it is genereally not enough
+different argument configuration.  This shows that it is generally not enough
 to auto-tune for particular argument configurations; instead it is important to
@@
 % space included 12000 kernels.
-% XXX: how to make sense of these paramters without code listing?
+% XXX: how to make sense of these parameters without code listing?
 %
@@
 %
-% we believe that our approach is the most practial one to get high perf on
+% we believe that our approach is the most practical one to get high perf on
 % recent hardware, since there is no consensus on hw features (and it won't
@@
 %
-% auto-tuning required hardward expertise *and* application-domain knowledge, we
+% auto-tuning required hardware expertise *and* application-domain knowledge, we
 % want to remove these restrictions ultimately we want a ml-based approach to
@@
 % more features (this is the hardest, and most manual part, may require
-% domain-specific knowledge, but not necessariyly, e.g. microbenchmarking)
+% domain-specific knowledge, but not necessarily, e.g. microbenchmarking)
 % * features derived from micro-benchmarking \citep{wong2010demystifying}
@@
 % experiments \cite{HT} our method is indeed very well suited for parallel
-% explorating of multiple configurations for learning (in a cluster, e.g. in HT
+% exploration of multiple configurations for learning (in a cluster, e.g. in HT
 % context) this may get us \emph{more} than linear speed ups for better accuracy,
@@
 %
-% interpret the learned model and confront the performance against theoreticial
+% interpret the learned model and confront the performance against theoretical
 % (or approximated) models of mem/com/compute, arithmetic intensity, etc. to

### 2012-03-13 paper/final.tex
adding acknowledgement section
final touch-up

@@
 0.1 seconds of hill climbing on the regression model (``{\em
-predictive} auto-tuning'') can achieve an average of 95\% of the
-speed-up brought by minutes of empirical auto-tuning.  Our approach is not
+predictive} auto-tuning'') can achieve almost the same
+speed-up as is brought by minutes of empirical auto-tuning.  Our approach is not
 specific to filterbank correlation, nor even to GPU kernel auto-tuning, and can
@@
 
-\cite{pinto+cox:2011gcg} recommends as a reference implementation:
+\cite{pinto+cox:2011gcg} recommend as a reference implementation:
 block\_w = 8,
@@
 these cases predictive auto-tuning took up to 3 seconds.  Even in these cases
-the predictive model is much faster than empirical auto-tuning, because the
-first kernel that actually runs is a fast one.
+the predictive model is much faster than empirical auto-tuning because the
+first kernel that runs successfully is still typically a fast one.
 
@@
 %Our
+\section{Acknowledgments}
+
+This work was funded by the Rowland Institute of Harvard, the NVIDIA
+Graduate Fellowship, and the National Science Foundation (IIS 0963668).
+Hardware support was generously provided by the NVIDIA Corporation.
 

### 2012-03-13 paper/final.tex
hack a word-wrapping dash

@@
 automatic methods known collectively as ``auto-tuning.'' Two major auto-tuning
-approaches have emerged in the extensive literature covering the subject (see
+approaches have e-merged in the extensive literature covering the subject (see
 surveys in e.g.
